### 2015-01-20 papers/uai-2015/draft.tex
init commit

\documentclass[]{article}
\usepackage{proceed2e}

% Set the typeface to Times Roman
\usepackage{times}

% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
%\usepackage{subcaption} 
%usepackage[numers]{natbib}
% For citations
%\usepackage[numbers]{natbib}

% For algorithms
%\usepackage{algorithm}
%\usepackage{algorithmic}

% As of 2011, we use the hyperref package to produce hyperlinks in the
% resulting PDF.  If this breaks your system, please commend out the
% following usepackage line and replace \usepackage{icml2014} with
% \usepackage[nohyperref]{icml2014} above.
%\usepackage{hyperref}
% Packages hyperref and algorithmic misbehave sometimes.  We can fix
% this with the following command.
%\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Employ the following version of the ``usepackage'' statement for
% submitting the draft version of the paper for review.  This will set
% the note in the first column to ``Under review.  Do not distribute.''
\usepackage{bm}
\usepackage{amsmath}
%\usepackage{nips_style} 
%\usepackage{sidecap}
%\usepackage[export]{adjustbox}

% Leave date blank
\date{}

\pagestyle{myheadings}
%\usepackage{sidecap}
%\usepackage[export]{adjustbox}

%\newcommand{\theHalgorithm}{\arabic{algorithm}}

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

\input{cdefs}

\title{Likelihood-free Inference and Optimization \\ using Stochastic Gradient Approximations}

\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
          % UAI  reviewing is double-blind.

% \author{
% Edward Meeds \\
% Informatics Institute \\
% University of Amsterdam \\
% \texttt{tmeeds@gmail.com} \\
% \and
% \author{
% Robert Leenders \\
% Informatics Institute \\
% University of Amsterdam \\
% \texttt{robert.leenders@student.uva.nl} \\
% \and
% Max Welling \\
% Informatics Institute\\
% University of Amsterdam \\
%  \texttt{welling.max@gmail.com}
% }
%\nipsfinalcopy
\begin{document} 
	\vskip -0.3in

\maketitle

% 
% 
% \begin{abstract} 
% \input{abstract}
% \end{abstract} 
\begin{abstract} 
%\input{abstract}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{INTRODUCTION} \label{introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{GRADIENTS FROM FORWARD SIMULATIONS} \label{gradsforward}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Robbin's Monro: SGD}

\subsection{Spall's method: SASP}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{LANGEVIN DYNAMICS} \label{ld}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{STOCHASTIC-GRADIENT LD} \label{sgld}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{SGLD-ABC} \label{sgld}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Statistical Tests for Gradients}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{EXPERIMENTS} \label{experiments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Logistic Regression with Stochastic Gradients}

\subsection{Likelihood-free Inference of Blowfly Dynamics}

\subsection{Other Useful ABC problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{DISCUSSION} \label{discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{CONCLUSION} \label{conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection*{References}
\newpage
{
\bibliographystyle{abcsgld}
}

\end{document}

### 2015-01-20 code/demo_spsa_lr_mnist.py

@@ -1,0 +1,111 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+import gzip, cPickle
+import time
+from numpy import newaxis
+
+from ml_problems import *
+from sa_algorithms import *
+
+def onehot( t, K ):
+  N=len(t)
+  T = np.zeros( (N,K))
+  
+  for trow,tk in zip( T, t ):
+    trow[tk] = 1.0
+  return T
+def load_mnist():
+  f = gzip.open('../../data/mnist.pkl.gz', 'rb')
+  data = cPickle.load(f)
+  return data
+  
+if __name__ == "__main__":
+  #assert False, "add binary label per class, compare log-likelihoods, classification error"
+  np.random.seed(1)
+  K = 10
+  lr = 0.002
+  decay = 0.99999
+  init_w_std = 0.000001
+  mom = 0.0
+  max_steps = 300000
+  print "loading mnist..."
+  (X_train,t_train),(X_valid,t_valid),(X_test,t_test) = load_mnist()
+  print "one-hotting labels.."
+  T_train = onehot( t_train, K )
+  T_valid = onehot( t_valid, K )
+  T_test  = onehot( t_test, K )
+  
+  m = X_train.mean(0)
+  X_train -= m
+  X_valid -= m
+  X_test  -= m
+  
+  # print "constructing LR model..."
+  # #LR_model = MulticlassLogisticRegression( T_valid, X_valid, T_test, X_test )
+  LR_model = MulticlassLogisticRegression( T_train, X_train, T_test, X_test )
+  # LR_model2 = MulticlassLogisticRegression( T_train, X_train, T_test, X_test )
+  # LR_model3 = MulticlassLogisticRegression( T_train, X_train, T_test, X_test )
+  # LR_model4 = MulticlassLogisticRegression( T_train, X_train, T_test, X_test )
+  #
+  # models = [LR_model, LR_model2, LR_model3, LR_model4]
+  #ALR = AveragedLogisticRegression(models)
+  # print "gradient ascent..."
+  # ALR.gradient_ascent(lr, decay, mom, init_w_std, max_steps)
+  
+  #LR_model.gradient_ascent(lr, decay, mom, init_w_std, max_steps)
+  D = X_train.shape[1]
+
+  max_iters = 500000
+  q         = 1
+  c         = 0.2
+  N         = len(T_train)
+  batchsize = 10
+  alpha     = 0.1
+  gamma     = 0.9
+  
[…]

### 2015-01-20 code/ml_problems.py

@@ -1,0 +1,129 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+from numpy import newaxis
+
+def logsumexp(x,dim=0):
+  """Compute log(sum(exp(x))) in numerically stable way."""
+  #xmax = x.max()
+  #return xmax + log(exp(x-xmax).sum())
+  if dim==0:
+    xmax = x.max(0)
+    return xmax + np.log(np.exp(x-xmax).sum(0))
+  elif dim==1:
+    xmax = x.max(1)
+    return xmax + np.log(np.exp(x-xmax[:,newaxis]).sum(1))
+  else: 
+    raise 'dim ' + str(dim) + 'not supported'
+        
+def onehot( t, K ):
+  N=len(t)
+  T = np.zeros( (N,K))
+  
+  for trow,tk in zip( T, t ):
+    trow[tk] = 1.0
+  return T
+  
+def softmax( A, return_log=False ):
+  # input is the N*K activations
+  
+  log_probabilities = A - logsumexp(A,1).reshape( (len(A),1) )
+  
+  probabilities = np.exp( log_probabilities )
+  
+  if return_log:
+    return probabilities, log_probabilities
+  else:
+    return probabilities
+  
+def classification_error( t, y ):
+  N = len(t)
+  I = pp.find( t != y )
+  return float(len(I))/N
+
+class MulticlassLogisticRegression(object):
+  def __init__( self, T_mat, X_mat, T_test, X_test ):
+    self.X = X_mat
+    self.T = T_mat
+    self.Xtest = X_test
+    self.Ttest = T_test
+    
+    self.t_train = np.argmax(self.T,1)
+    self.t_test  = np.argmax(self.Ttest,1)
+    
+    # assume N*K for T_mat
+    self.N,self.K = T_mat.shape
+    self.Ntest,self.K = self.Ttest.shape
+    
+    # assume X_mat N*D
+    N,self.D = X_mat.shape
+  
+    assert N == self.N, "should be same shape"
+    
+  def train_cost( self, w, ids, l1=0,l2=0 ):
+    W = w.reshape( (self.D,self.K))
+    
+    utility = self.loglikelihood( W, ids = ids )
+    
+    utility -= l1*np.abs(w).sum()
+    utility -= l2*pow(w,2).sum()
+    
+    return utility 
+    
+  def class_error( self, w, X, t ):
+    W = w.reshape( (self.D,self.K))
+    Y, logY = softmax( np.dot( X, W ), return_log = True ) 
[…]

### 2015-01-20 code/sa_algorithms.py

@@ -1,0 +1,115 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+from numpy import newaxis
+import pdb
+
+def spsa_gradient( f, w, c, N, batchsize, q = 1, ids = None, batchreplace = 1.0, mask = None, l1 = 0, l2=0, g_var=None):
+  # f: the objective function, e.g. loglikelihood
+  # w: parameters at this time step, len p
+  # c: step size, constant for all dimensions
+  # N: total nbr data vectors
+  # batchsize: nbr of function evals to use in f
+  # q: nbr of repeats for the gradient
+  # batchreplace: percent of ids to replace
+  
+  c_adjust = c*np.ones(len(w))
+  #if g_var is not None:
+  #  c_adjust = len(w)*c*(np.sqrt(g_var)+1e-3)/(np.sqrt(g_var).sum()+len(w)*1e-3)
+  #  #c_adjust = len(w)*c*g_var/(g_var.sum())
+  
+  #pdb.set_trace()
+  p = len(w)
+  
+  g = np.zeros( p )
+  
+  if ids is None:
+    ids = np.random.permutation( N )[:batchsize]
+  else:
+    nbr_replace = int(batchreplace*batchsize)
+    new_ids = np.random.permutation( N )[:nbr_replace]
+    ids_in_batch = np.random.permutation( batchsize )
+    for i, idx in zip( range(nbr_replace), ids_in_batch[:nbr_replace]):
+      ids[idx] = new_ids[i]
+      
+  G = np.zeros( (q,p))
+  for j in range(q):
+    # TODO: could put different batch ids here
+    # ids=new_ids
+    
+    mask = 2*np.random.binomial(1,0.5,p)-1
+  
+    f_plus  = f( w + c_adjust*mask, ids = ids, l1=l1, l2=l2 )
+    f_minus = f( w - c_adjust*mask, ids = ids, l1=l1, l2=l2 )
+  
+    g += (f_plus-f_minus)/(2*c_adjust*mask)
+    G[j] = (f_plus-f_minus)/(2*c_adjust*mask)
+  
+  g/=q  
+  
+  #V = pow( G - g, 2).mean(0)   
+  
+  #g/=V
+  #pdb.set_trace()
+  return g, ids
+  
+  # GM = (1-mom)*GM + mom*G
+  #
+  # self.W = self.W + lr*GM
+ #
+ # historical_grad += g^2
+ # adjusted_grad = grad / (fudge_factor + sqrt(historical_grad))
[…]

### 2015-01-21 papers/uai-2015/draft.tex
draft writing + blowfly optimisation

@@
 \usepackage{amsmath}
+%\usepackage{array}
 %\usepackage{nips_style} 
@@
 %\usepackage[export]{adjustbox}
+%\include{cdefs}
 
@@
 \begin{abstract} 
-%\input{abstract}
+In this paper we apply recent techniques from Bayesian inference that use gradient information to sample efficiently from the true posterior distribution to the {\em likelihood-free} or {\em approximate Bayesian computation} (ABC) setting.
+To do this for ABC we adopt a {\em gradient-free} stochastic approximation algorithm by Spall~\cite{spall1999}.
+Together these algorithms provide both optimization and inference for likelihood-free models as the algorithm ABC-SGLD transitions from optimization to sampling.
+We demonstrate ABC-SGLD on problems where the true gradient information is known and on challenging ABC simulators.
 \end{abstract}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{itemize}
+  \item  Arguably the two most useful procedures in simulation-based science are optimization and Bayesian inference.  
+  \item  Optimization can take the form of simple grid-search to sophisticated techniques like factorial design \cite{factorialdesign}, Bayesian experiment design \cite{bayesexpdesign}, (mention others).
+  \item Recently, Bayesian optimization techniques have shown success in optimizing very expensive black-box simulators \cite{bosuccess}.
+  \item The main approach of Bayesian inference of simulator parameters is ABC.  ABC is largely based a few sampling algorithms: SMC and MCMC.  
+  \item Though gradients are not directly computable for simulators, they can be computed analytically by using finite differences (see \cite{kiefer1952}).
+This quickly becomes infeasible for large $p$ problems.
+There is however an alternative stochastic approximation algorithm by Spall \cite{spall1999} that requires only 2 simulation calls independent of $p$.
+  \item By using this approximation, gradient-based algorithms can be adopted for simulators: for optimization, using analogous stochastic gradient descent algorithms and for Bayesian inference using Langevin dynamics \cite{langevin}.
+  \item Recently, the SGLD \cite{wellingsgld} algorithm has efficiently combined optimization with inference using ideas from SGD with Langevin dynamics.
+  \item This paper describes ABC-SGLD.
+\end{itemize}
 
@@
 
-\subsection{Spall's method: SASP}
+\newcommand{\gradestimate}[2]{\hat{g}_{#1}\lp#2\rp}
+\newcommand{\gradestimaterepeat}[3]{\hat{g}^{#3}_{#1}\lp#2\rp}
+\newcommand{\feval}[2]{y_{#1}^{#2}}
+\newcommand{\cstep}[2]{c_{#1#2}}
+\newcommand{\cstepconst}[1]{c_{#1}}
+\newcommand{\dstep}[2]{\Delta_{#1#2}}
+\newcommand{\dstepconst}[1]{\Delta_{#1}}
+\newcommand{\loglik}[1]{L\lp#1\rp}
+
+\subsection{Spall's method: SPSA}
+In the gradient-free setting, Spall \cite{spall_spsa} provides a stochastic approximate to the true gradient using only 2 forward simulations (function evaluations).
+This is in contrast to multivariate finite-difference stochastic approximation FDSA \cite{kiefer1952} requiring $2p$ evaluations.
+
+The gradient estimate is
+\begin{equation}
+  \gradestimate{t}{\thetav_t} = \begin{bmatrix} 
+                                   \frac{\feval{t}{+} - \feval{t}{-}}{2 \cstep{t}{1} \dstep{t}{1}} \\
+                                   \vdots \\
+                                   \frac{\feval{t}{+} - \feval{t}{-}}{2 \cstep{t}{p} \dstep{t}{p}} \\
+                                \end{bmatrix}
+\end{equation}
+where $\cstep{t}{p}$ is a step-size that is usually constant for all dimensions $p$, but can be different (as shown in this case); $\dstep{t}{p} \in {-1, +1}$ is a {\em perturbation mask} (called symmetric Bernouilli variables by Spall), i.e.
+$\dstep{t}{p} \sim 2*\text{Bernouilli}(0.5) - 1$; and  $\feval{t}{\pm}$ are function evaluations:
+\begin{eqnarray}
+  \feval{t}{+} &= \loglik{\thetav + \cstepconst{t}\dstepconst{t}} \\
+  \feval{t}{-} &= \loglik{\thetav - \cstepconst{t}\dstepconst{t}}
+\end{eqnarray}
+A way of reducing the noise in the gradient is by averaging over $q$ draws of $\dstepconst{t}$:
+\begin{equation}
+  \gradestimate{t}{\thetav_t} = \frac{1}{q} \sum_{j=1}^q \gradestimaterepeat{t}{\thetav_t}{j}
+\end{equation}
+where for each $j$ new perturbation masks are drawn.
+
+Another variance reduction technique called the method of {\em common random numbers} (CRN) \cite{kleinman1999} sets a common seed for the random number generator (RNG) for both calls to the simulator.
+This technique can remove the effect of the simulator noise in the gradient estimate, leaving on the randomness of the perturbation masks as the source of noise.
+Consider using SPSA instead of SGD: the CRN technique is equivalent of using the same batch of data vectors to evaluate the log-likelihood which is a sensible approach.
+
+\subsubsection{Variations}
+
+\begin{itemize}
+\item Varying $c_q$: use a different per repeat.
+Seems to converge faster.
+What is the correct noise process?
+Right now trying this: flip coin, if heads perturb $c *= 1+U(0,1)$, else perturb $c /= 1 + U(0,1)$.
+\end{itemize}
 

### 2015-01-21 code/demo_blowfly_optimization.py

@@ -1,0 +1,139 @@
+from abcpy.factories import *
+from abcpy.problems.blowfly.blowfly import *
+from sa_algorithms import *
+
+import pylab as pp
+
+problem_params = default_params()
+problem_params["epsilon"] = 0.05 #10*np.array([1.0,1.0,0.5,0.5,0.5,0.1,0.01,0.5,2.0,1])
+problem_params["blowfly_filename"] = "./data/blowfly.txt"
+problem = BlowflyProblem( problem_params, force_init = True )
+
+state_params = state_params_factory.scrape_params_from_problem( problem )
+
+# set is_marginal to true so using only "current" state will force a re-run
+mcmc_params  = mcmc_params_factory.scrape_params_from_problem( problem, type="mh", is_marginal = True, nbr_samples = 1 )
+
+# ignore algo, just doing this to get state object
+algo_params = { "modeling_approach"  : "kernel",
+                "observation_groups" : problem.get_obs_groups(),
+                "state_params"       : state_params,
+                "mcmc_params"        : mcmc_params,
+                "algorithm"          : "model_mcmc"
+              }
+algo, model, state  = algo_factory.create_algo_and_state( algo_params )
+recorder     = recorder_factory.create_recorder( {} )
+
+class LikelihoodFree( object ):
+  def __init__( self, model, state, recorder ):
+    self.model = model
+    self.state = state
+    self.recorder = recorder
+    self.tau   = 15
+    
+    self.moving_average_statistics = None
+    self.count = 0
+  
+  def add_prediction( self, x_statistics ):
+    if self.moving_average_statistics is None:
+      self.moving_average_statistics = x_statistics.copy()
+      self.count = 1
+    else:
+      self.moving_average_statistics = (float( self.count)/float(self.count+1))*self.moving_average_statistics + (float( 1)/float(self.count+1))*x_statistics
+      self.count +=1
+    self.error = pow( self.state.observation_statistics - self.moving_average_statistics, 2 ).sum()
+    
+  def train_cost( self, w, seed = None ):
+    #theta = np.hstack( (w,self.tau))
+    theta = w.copy()
+    theta[-1] = int(np.exp(w[-1]))
+    #print theta
+    self.state.theta = theta#.reshape( (1,len(theta)))
+    self.state.loglikelihood_is_computed = False
+    self.model.set_current_state( self.state )
+    
[…]

### 2015-01-21 code/demo_gradients_compare.py

@@ -1,0 +1,120 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+import gzip, cPickle
+import time
+from numpy import newaxis
+
+from ml_problems import *
+from sa_algorithms import *
+
+def onehot( t, K ):
+  N=len(t)
+  T = np.zeros( (N,K))
+  
+  for trow,tk in zip( T, t ):
+    trow[tk] = 1.0
+  return T
+def load_mnist():
+  f = gzip.open('../../data/mnist.pkl.gz', 'rb')
+  data = cPickle.load(f)
+  return data
+  
+if __name__ == "__main__":
+  #assert False, "add binary label per class, compare log-likelihoods, classification error"
+  np.random.seed(1)
+  K = 10
+  lr = 0.002
+  decay = 0.99999
+  init_w_std = 0.000001
+  mom = 0.0
+  max_steps = 300000
+  print "loading mnist..."
+  (X_train,t_train),(X_valid,t_valid),(X_test,t_test) = load_mnist()
+  print "one-hotting labels.."
+  T_train = onehot( t_train, K )
+  T_valid = onehot( t_valid, K )
+  T_test  = onehot( t_test, K )
+  D = X_train.shape[1]
+  N         = len(T_train)
+  
+  m = X_train.mean(0)
+  X_train -= m
+  X_valid -= m
+  X_test  -= m
+  
+  mom = 0.5
+  max_iters = 5000
+  batchsize = 100
+  alpha     = 0.0001
+  gamma     = 0.999
+  
+  q = 100
+  c = 0.01
+  
+  # print "constructing LR model..."
+  # #LR_model = MulticlassLogisticRegression( T_valid, X_valid, T_test, X_test )
+  LR_model = MulticlassLogisticRegression( T_train, X_train, T_test, X_test )
+  
+  bcompare = True
+  np.random.seed(1)
+  w = 0.0001*np.random.randn( D*K )
+  sg_params = { "ml_problem":LR_model, "max_iters":max_iters, "N":N, "batchsize":batchsize, "alpha":alpha, "gamma":gamma,"mom":mom, "l1":100.0,"l2":0.0,"q":q,"c":c,"compare":bcompare}
+  wout, errors = sgd( w, sg_params )
+    
+  #LR_model.gradient_ascent(lr, decay, mom, init_w_std, max_steps)
+  
+
+  # max_iters = 500000
+  # q         = 1
+  # c         = 0.2
+  # N         = len(T_train)
[…]

### 2015-01-21 code/demo_spsa_lr_mnist.py

@@ -56,24 +56,30 @@ if __name__ == "__main__":
   #LR_model.gradient_ascent(lr, decay, mom, init_w_std, max_steps)
   D = X_train.shape[1]
 
-  max_iters = 500000
+  C = np.cov( X_train.T )
+  f=0.1
+  #cc = f*np.sqrt(1.0/(np.diag( C )+f*f)) 
+  cc = f*np.sqrt((np.diag( C )+f)/f)
+  ccc= np.hstack( (cc,cc,cc,cc,cc,cc,cc,cc,cc,cc))
+  #ccc=0.01
+  max_iters = 5000
   q         = 1
-  c         = 0.2
+  c         = 1
   N         = len(T_train)
   batchsize = 10
   alpha     = 0.1
   gamma     = 0.9
   
-  #cs = [0.05,0.1,0.2,0.3]
-  cs = [0.01]
-  gammas = [0.99999]
+  #cs = [0.5,0.1,0.2,0.3]
+  cs = [ccc]
+  gammas = [0.9999]
   moms = [0.5]
-  batchsizes = [20]
-  qs = [10]
+  batchsizes = [50]
+  qs = [20]
   result = []
   batchreplaces = [1]
   for c in cs:
-    alpha = 3*c/(4*N)
+    alpha = 3*0.1/(4*N)
     for gamma in gammas:
       for mom in moms:
         for batchsize in batchsizes:
@@ -81,8 +87,10 @@ if __name__ == "__main__":
             for batchreplace in batchreplaces:
               np.random.seed(1)
               w = 0.0001*np.random.randn( D*K )
-              spall_params = {"ml_problem":LR_model, "max_iters":max_iters, "q":q,"c":c,"N":N, "batchsize":batchsize, "alpha":alpha, "gamma":gamma,"mom":mom,"batchreplace":batchreplace, "l1":100.0,"l2":0.0}
+              spall_params = {"ml_problem":LR_model, "max_iters":max_iters, "q":q,"c":c,"N":N, "batchsize":batchsize, "alpha":alpha, "gamma":gamma,"mom":mom,"batchreplace":batchreplace, "l1":0.0,"l2":0.0,"diag_add":0.01}
+              
               wout, errors = spall( w, spall_params )
+              #wout, errors = spall_with_hessian( w, spall_params )
       
               result.append({"c":c,"alpha":alpha,"gamma":gamma, "mom":mom, "batchsize":batchsize,"q":q,"errors":errors,"w":wout})
   

### 2015-01-21 code/ml_problems.py

@@ -68,7 +68,7 @@ class MulticlassLogisticRegression(object):
     utility -= l1*np.abs(w).sum()
     utility -= l2*pow(w,2).sum()
     
-    return utility 
+    return -utility 
     
   def class_error( self, w, X, t ):
     W = w.reshape( (self.D,self.K))
@@ -109,13 +109,17 @@ class MulticlassLogisticRegression(object):
     else:
       return loglike_data
     
-  def gradient( self, W, Y = None ):
-    new_ids = np.random.permutation( self.N )[:self.batchsize]
-    #ids = self.ids.copy()
-    #self.ids[:self.batchsize/4] = new_ids[:self.batchsize/4]
+  def gradient( self, w, ids = None ):
+    W = w.reshape( (self.D,self.K))
+    
+    if ids is None:
+      new_ids = np.random.permutation( self.N )[:self.batchsize]
+    else:
+      new_ids = ids
+      
     self.ids = new_ids
     ids = self.ids
-    #if Y is None:
+    n = len(ids)
     Y = softmax( np.dot( self.X[ids,:], W ) )
     
     
@@ -126,4 +130,5 @@ class MulticlassLogisticRegression(object):
     for k in range(self.K):
       g_k = np.dot( DIF[:,k].T, self.X[ids,:] ).T
       G[:,k] = g_k
-    return self.batchsize*G/self.N
+    g = G.reshape( (self.D*self.K,))
+    return self.N*g/n

### 2015-01-21 code/sa_algorithms.py

@@ -4,7 +4,23 @@ import scipy as sp
 from numpy import newaxis
 import pdb
 
-def spsa_gradient( f, w, c, N, batchsize, q = 1, ids = None, batchreplace = 1.0, mask = None, l1 = 0, l2=0, g_var=None):
+def sgradient( g, w, N, batchsize, q = 1, ids = None, l1 = 0, l2=0):
+  # g: the gradient of the objective function, e.g. grad of loglikelihood
+  # w: parameters at this time step, len p
+  # N: total nbr data vectors
+  # batchsize: nbr of function evals to use in f
+  
+  p = len(w)
+  
+  
+  if ids is None:
+    ids = np.random.permutation( N )[:batchsize]
+  
+  grads = g( w, ids )    
+  
+  return grads, ids
+
+def spsa_gradient_with_hessian( f, w, c, c_tilde, N, batchsize, q = 1, ids = None, batchreplace = 1.0, mask = None, l1 = 0, l2=0, g_var=None):
   # f: the objective function, e.g. loglikelihood
   # w: parameters at this time step, len p
   # c: step size, constant for all dimensions
@@ -14,6 +30,107 @@ def spsa_gradient( f, w, c, N, batchsize, q = 1, ids = None, batchreplace = 1.0,
   # batchreplace: percent of ids to replace
   
   c_adjust = c*np.ones(len(w))
+  
+  #pdb.set_trace()
+  p = len(w)
+  
+  g = np.zeros( p )
+  h = np.zeros( p )
+  
+  if ids is None:
+    ids = np.random.permutation( N )[:batchsize]
+      
+  G = np.zeros( (q,p))
+  for j in range(q):
+    # TODO: could put different batch ids here
+    # ids=new_ids
+    
+    # TODO: uncomment for perturbed c
+    # if np.random.randn()<0:
+    #   c_adjust *= 1 + np.random.rand()
+    # else:
+    #   c_adjust /= 1 + np.random.rand()
+        
+    mask = 2*np.random.binomial(1,0.5,p)-1
+    mask_tilde = 2*np.random.binomial(1,0.5,p)-1
+  
+    f_plus  = f( w + c_adjust*mask, ids = ids, l1=l1, l2=l2 )
+    f_minus = f( w - c_adjust*mask, ids = ids, l1=l1, l2=l2 )
+    
[…]

### 2015-01-22 code/demo_blowfly_optimization.py
added hessian, good spall for blowfly

@@ -25,7 +25,8 @@ algo, model, state  = algo_factory.create_algo_and_state( algo_params )
 recorder     = recorder_factory.create_recorder( {} )
 
 class LikelihoodFree( object ):
-  def __init__( self, model, state, recorder ):
+  def __init__( self, problem, model, state, recorder ):
+    self.problem = problem
     self.model = model
     self.state = state
     self.recorder = recorder
@@ -46,7 +47,7 @@ class LikelihoodFree( object ):
   def train_cost( self, w, seed = None ):
     #theta = np.hstack( (w,self.tau))
     theta = w.copy()
-    theta[-1] = int(np.exp(w[-1]))
+    theta[-1] = np.ceil(np.exp(w[-1]))
     #print theta
     self.state.theta = theta#.reshape( (1,len(theta)))
     self.state.loglikelihood_is_computed = False
@@ -55,9 +56,14 @@ class LikelihoodFree( object ):
     if seed is not None:
       np.random.seed(seed)
     #pdb.set_trace()
-    log_posterior = self.model.log_posterior()
-    
-    return -log_posterior 
+    #log_posterior = self.model.log_posterior()
+    loglikelihood = self.model.current.loglikelihood()
+    return -loglikelihood 
+  
+  def grad_prior( self, w ):
+    theta = w.copy()
+    theta[-1] = np.ceil(np.exp(w[-1]))
+    return -self.problem.theta_prior_logpdf_grad( theta )
     
   def train_error( self, w, seed = None ):
     # ignore w (assume already computed stats)
@@ -74,7 +80,7 @@ class LikelihoodFree( object ):
     
 if __name__ == "__main__":
   
-  abc_problem = LikelihoodFree( model, state, recorder )
+  abc_problem = LikelihoodFree( problem, model, state, recorder )
   
   max_iters = 5000
   q         = 1
@@ -83,18 +89,18 @@ if __name__ == "__main__":
   gamma     = 0.9
   
   #cs = [0.5,0.1,0.2,0.3]
-  cs = [np.array([0.5,0.1,0.1,0.5,0.01,0.1])]
+  cs = [0.1] #np.array([0.5,0.1,0.1,0.5,0.01,0.1])]
   gammas = [0.9999]
-  moms = [0.9]
+  moms = [0.1]
   qs = [5]
   result = []
   for c in cs:
-    #alpha = 0.000001 #*3*c/(4)
[…]

### 2015-01-22 code/sa_algorithms.py

@@ -68,7 +68,7 @@ def spsa_gradient_with_hessian( f, w, c, c_tilde, N, batchsize, q = 1, ids = Non
     
     delta_g = (g1plus - g1minus)/(2*c_adjust*mask)
     
-    h += delta_g #.reshape((p,1)) #0.5*( delta_g.reshape((p,1))+delta_g.reshape((1,p)) )
+    h += delta_g
     
     #pdb.set_trace()
     G[j] = (f_plus-f_minus)/(2*c_adjust*mask)
@@ -82,7 +82,7 @@ def spsa_gradient_with_hessian( f, w, c, c_tilde, N, batchsize, q = 1, ids = Non
   #pdb.set_trace()
   return g, h, ids
 
-def spsa_abc_gradient( f, w, c, q = 1, seed = None, mask = None ):
+def spsa_abc_gradient( f, w, c, q = 1, seed = None, mask = None, hessian=False, c_tilde=None ):
   # f: the objective function, e.g. loglikelihood
   # w: parameters at this time step, len p
   # c: step size, constant for all dimensions
@@ -92,7 +92,8 @@ def spsa_abc_gradient( f, w, c, q = 1, seed = None, mask = None ):
   p = len(w)
   
   g = np.zeros( p )
-      
+  if hessian:
+    h = np.zeros( (p,p))
   G = np.zeros( (q,p))
   for j in range(q):
     # TODO: uncomment for perturbed c
@@ -107,19 +108,44 @@ def spsa_abc_gradient( f, w, c, q = 1, seed = None, mask = None ):
     f_plus  = f( w + c_adjust*mask, seed = seed )
     f_minus = f( w - c_adjust*mask, seed = seed )
   
+    g += (f_plus-f_minus)/(2*c_adjust*mask)
+    #print g
+    G[j] = (f_plus-f_minus)/(2*c_adjust*mask)
+    
     if np.isinf( f_plus ) or np.isinf( f_minus ):
       pdb.set_trace()
+    
+    if hessian:
+      mask_tilde = 2*np.random.binomial(1,0.5,p)-1
       
-    g += (f_plus-f_minus)/(2*c_adjust*mask)
-    G[j] = (f_plus-f_minus)/(2*c_adjust*mask)
+      # ones-sided grads
+      f_plus_1  = f( w + c_adjust*mask + c_tilde*mask_tilde, seed = seed )
+      f_minus_1 = f( w - c_adjust*mask + c_tilde*mask_tilde, seed = seed )
[…]

### 2015-01-22 code/demo_blowfly_optimization.py
good optimisation for blowfly, added prior grads

@@ -63,7 +63,7 @@ class LikelihoodFree( object ):
   def grad_prior( self, w ):
     theta = w.copy()
     theta[-1] = np.ceil(np.exp(w[-1]))
-    return -self.problem.theta_prior_logpdf_grad( theta )
+    return self.problem.theta_prior_logpdf_grad( theta )
     
   def train_error( self, w, seed = None ):
     # ignore w (assume already computed stats)

### 2015-01-22 code/sa_algorithms.py

@@ -302,7 +302,7 @@ def spall_abc( w, params ):
     # else:
     #   g_hat, h_hat, seed = spsa_abc_gradient( problem.train_cost, w, c, q=q, seed=seed, hessian= hessian, c_tilde=c_tilde )
     
-    #g_hat += problem.grad_prior( w )
+    g_hat += problem.grad_prior( w )
     
     if t==0:
       g_mom = g_hat.copy()

### 2015-01-26 code/demo_sgld.py
Added spall gradient and numerical gradient

@@ -7,11 +7,63 @@ def generate_data(theta, data_size, data_sigma):
                   np.random.normal(np.sum(theta), data_sigma**2, data_size))
 
 
-def compute_prior_gradient(theta, theta_mean, theta_sigma):
+def prior(theta, theta_mean, theta_sigma):
+    return np.log(np.prod(1.0 / (theta_sigma * np.sqrt(2*np.pi)) *
+                          np.exp(-0.5 * ((theta-theta_mean)/theta_sigma)**2)))
+
+
+def likelihood(data, data_sigma, theta):
+    likelihood = np.prod(
+        0.5 * 1.0 / (data_sigma * np.sqrt(2*np.pi)) *
+        np.exp(-0.5 * ((data-theta[0])/data_sigma)**2) +
+        0.5 * 1.0 / (data_sigma * np.sqrt(2*np.pi)) *
+        np.exp(-0.5 * ((data-np.sum(theta))/data_sigma)**2)
+    )
+    likelihood = len(data) * np.log(0.5*(2*np.pi*data_sigma**2)**-0.5) + \
+        np.sum(np.log(np.exp(-0.5 * ((data-theta[0])/data_sigma)**2) +
+                      np.exp(-0.5 * ((data-np.sum(theta))/data_sigma)**2)))
+    return likelihood
+
+
+def likelihood_spall_gradient(data, data_sigma, theta, num_repeats, epsilon):
+    gradient = np.zeros(len(theta))
+    for i in range(num_repeats):
+        mask = 2*np.random.binomial(1, 0.5, len(theta))-1
+        theta_negative = theta - epsilon*mask
+        theta_positive = theta + epsilon*mask
+        likelihood_negative = likelihood(data, data_sigma, theta_negative)
+        likelihood_positive = likelihood(data, data_sigma, theta_positive)
+        gradient += (likelihood_positive-likelihood_negative) / \
+                    (2*epsilon*mask)
+    return gradient / num_repeats
+
+
+def posterior(data, data_sigma, theta, theta_mean, theta_sigma):
+    prior = prior(theta, theta_mean, theta_sigma)
+    likelihood = likelihood(data, data_sigma, theta)
+    return prior+likelihood
+
+
+def likelihood_numerical_gradient(data, data_sigma, theta):
+    scale = 0.01
+    gradient = np.zeros(len(theta))
+    for i in range(len(theta)):
+        new_theta = np.copy(theta)
+        new_theta[i] -= scale
[…]

### 2015-01-26 code/demo_blowfly_optimization.py
ADAgrad RSMProp ADAm

@@ -5,8 +5,9 @@ from sa_algorithms import *
 import pylab as pp
 
 problem_params = default_params()
-problem_params["epsilon"] = 0.05 #10*np.array([1.0,1.0,0.5,0.5,0.5,0.1,0.01,0.5,2.0,1])
+problem_params["epsilon"] = 0.01 #10*np.array([1.0,1.0,0.5,0.5,0.5,0.1,0.01,0.5,2.0,1])
 problem_params["blowfly_filename"] = "./data/blowfly.txt"
+#problem_params["blowfly_filename"] = "./data/bf2.txt"
 problem = BlowflyProblem( problem_params, force_init = True )
 
 state_params = state_params_factory.scrape_params_from_problem( problem )
@@ -82,25 +83,30 @@ if __name__ == "__main__":
   
   abc_problem = LikelihoodFree( problem, model, state, recorder )
   
-  max_iters = 5000
+  max_iters = 2000
   q         = 1
   c         = 0.2
   alpha     = 0.1
-  gamma     = 0.9
-  
+  gamma     = 0.75
+  mom_beta1 = 0.75 # on gradient
+  mom_beta2 = 0.9 # on gradient_squared
   #cs = [0.5,0.1,0.2,0.3]
   cs = [0.1] #np.array([0.5,0.1,0.1,0.5,0.01,0.1])]
   gammas = [0.9999]
-  moms = [0.1]
+  moms = [0.0]
   qs = [5]
   result = []
   for c in cs:
-    alpha = 0.0001 #*3*c/(4)
+    # for "grad"
+    #alpha = 1e-3 #*3*c/(4)
+    
+    # for others
+    alpha = 1e-1 #*3*c/(4)
     #alpha  = 1.001 #0.00002 #*3*c/(4)
     for gamma in gammas:
       for mom in moms:
         for q in qs:
-          np.random.seed(1)
+          np.random.seed(5)
           theta_rand = problem.theta_prior_rand()
           w = theta_rand #[:5]
           w[-1] = np.log(w[-1])
@@ -109,13 +115,18 @@ if __name__ == "__main__":
                               "max_iters":max_iters, 
                               "q":q,
                               "c":c,
-                              "alpha":problem_params["epsilon"]*alpha*1, 
-                              "gamma":gamma,
-                              "mom":mom,
+                              "alpha":alpha, 
+                              "gamma_alpha":1,
+                              "gamma_c":0.9999,
+                              "gamma_eps":0.9999,
[…]

### 2015-01-26 code/demo_spsa_lr_mnist.py

@@ -27,7 +27,7 @@ if __name__ == "__main__":
   lr = 0.002
   decay = 0.99999
   init_w_std = 0.000001
-  mom = 0.0
+  mom = 0.5
   max_steps = 300000
   print "loading mnist..."
   (X_train,t_train),(X_valid,t_valid),(X_test,t_test) = load_mnist()
@@ -41,9 +41,12 @@ if __name__ == "__main__":
   X_valid -= m
   X_test  -= m
   
+  l1 = 0.0
+  l2 = 0.0
+  
   # print "constructing LR model..."
   # #LR_model = MulticlassLogisticRegression( T_valid, X_valid, T_test, X_test )
-  LR_model = MulticlassLogisticRegression( T_train, X_train, T_test, X_test )
+  LR_model = MulticlassLogisticRegression( T_train, X_train, T_test, X_test, l1,  l2 )
   # LR_model2 = MulticlassLogisticRegression( T_train, X_train, T_test, X_test )
   # LR_model3 = MulticlassLogisticRegression( T_train, X_train, T_test, X_test )
   # LR_model4 = MulticlassLogisticRegression( T_train, X_train, T_test, X_test )
@@ -62,32 +65,54 @@ if __name__ == "__main__":
   cc = f*np.sqrt((np.diag( C )+f)/f)
   ccc= np.hstack( (cc,cc,cc,cc,cc,cc,cc,cc,cc,cc))
   #ccc=0.01
-  max_iters = 5000
+  max_iters = 10000
   q         = 1
   c         = 1
   N         = len(T_train)
   batchsize = 10
   alpha     = 0.1
   gamma     = 0.9
+  mom_beta1 = 0.9 # on gradient
+  mom_beta2 = 0.9 # on gradient_squared
   
   #cs = [0.5,0.1,0.2,0.3]
   cs = [ccc]
   gammas = [0.9999]
-  moms = [0.5]
-  batchsizes = [50]
-  qs = [20]
+  moms = [0.0]
+  batchsizes = [100]
+  qs = [10]
   result = []
   batchreplaces = [1]
   for c in cs:
-    alpha = 3*0.1/(4*N)
+    # for "grad"
+    #alpha = 0.01/N
+    
+    # for others
+    alpha = 0.01 #e-1 #*3*c/(4)
+    
     for gamma in gammas:
       for mom in moms:
         for batchsize in batchsizes:
           for q in qs:
             for batchreplace in batchreplaces:
[…]

### 2015-01-26 code/ml_problems.py

@@ -42,7 +42,7 @@ def classification_error( t, y ):
   return float(len(I))/N
 
 class MulticlassLogisticRegression(object):
-  def __init__( self, T_mat, X_mat, T_test, X_test ):
+  def __init__( self, T_mat, X_mat, T_test, X_test, l1=0, l2=0 ):
     self.X = X_mat
     self.T = T_mat
     self.Xtest = X_test
@@ -57,18 +57,26 @@ class MulticlassLogisticRegression(object):
     
     # assume X_mat N*D
     N,self.D = X_mat.shape
+    
+    self.l1 = l1
+    self.l2 = l2
   
     assert N == self.N, "should be same shape"
     
-  def train_cost( self, w, ids, l1=0,l2=0 ):
+  def train_cost( self, w, ids ):
     W = w.reshape( (self.D,self.K))
     
     utility = self.loglikelihood( W, ids = ids )
     
-    utility -= l1*np.abs(w).sum()
-    utility -= l2*pow(w,2).sum()
+    # utility -= l1*np.abs(w).sum()
+    # utility -= l2*pow(w,2).sum()
     
     return -utility 
+  
+  def grad_prior( self, w ):
+    g = self.l1*np.sign(w)
+    g = self.l2*w
+    return g
     
   def class_error( self, w, X, t ):
     W = w.reshape( (self.D,self.K))

### 2015-01-26 code/sa_algorithms.py

@@ -20,67 +20,67 @@ def sgradient( g, w, N, batchsize, q = 1, ids = None, l1 = 0, l2=0):
   
   return grads, ids
 
-def spsa_gradient_with_hessian( f, w, c, c_tilde, N, batchsize, q = 1, ids = None, batchreplace = 1.0, mask = None, l1 = 0, l2=0, g_var=None):
-  # f: the objective function, e.g. loglikelihood
-  # w: parameters at this time step, len p
-  # c: step size, constant for all dimensions
-  # N: total nbr data vectors
-  # batchsize: nbr of function evals to use in f
-  # q: nbr of repeats for the gradient
-  # batchreplace: percent of ids to replace
-  
-  c_adjust = c*np.ones(len(w))
-  
-  #pdb.set_trace()
-  p = len(w)
-  
-  g = np.zeros( p )
-  h = np.zeros( p )
-  
-  if ids is None:
-    ids = np.random.permutation( N )[:batchsize]
-      
-  G = np.zeros( (q,p))
-  for j in range(q):
-    # TODO: could put different batch ids here
-    # ids=new_ids
-    
-    # TODO: uncomment for perturbed c
-    # if np.random.randn()<0:
-    #   c_adjust *= 1 + np.random.rand()
-    # else:
-    #   c_adjust /= 1 + np.random.rand()
-        
-    mask = 2*np.random.binomial(1,0.5,p)-1
-    mask_tilde = 2*np.random.binomial(1,0.5,p)-1
-  
-    f_plus  = f( w + c_adjust*mask, ids = ids, l1=l1, l2=l2 )
-    f_minus = f( w - c_adjust*mask, ids = ids, l1=l1, l2=l2 )
-    
-    f_plus_1  = f( w + c_adjust*mask + c_tilde*mask_tilde, ids = ids, l1=l1, l2=l2 )
-    f_minus_1 = f( w - c_adjust*mask + c_tilde*mask_tilde, ids = ids, l1=l1, l2=l2 )
-  
-    g += (f_plus-f_minus)/(2*c_adjust*mask)
-    
-    # ones-sided grads
-    g1plus  = (f_plus-f_plus_1)/(c_tilde*mask_tilde)
-    g1minus = (f_minus-f_minus_1)/(c_tilde*mask_tilde)
-    
-    delta_g = (g1plus - g1minus)/(2*c_adjust*mask)
-    
-    h += delta_g
-    
-    #pdb.set_trace()
[…]

### 2015-01-26 code/sa_algorithms.py
conflict fixes

@@ -20,7 +20,6 @@ def sgradient( g, w, N, batchsize, q = 1, ids = None, l1 = 0, l2=0):
 
   return grads, ids
 
-<<<<<<< HEAD
 # def spsa_gradient_with_hessian( f, w, c, c_tilde, N, batchsize, q = 1, ids = None, batchreplace = 1.0, mask = None, l1 = 0, l2=0, g_var=None):
 #   # f: the objective function, e.g. loglikelihood
 #   # w: parameters at this time step, len p
@@ -82,69 +81,6 @@ def sgradient( g, w, N, batchsize, q = 1, ids = None, l1 = 0, l2=0):
 #   #g/=V
 #   #pdb.set_trace()
 #   return g, h, ids
-=======
-def spsa_gradient_with_hessian( f, w, c, c_tilde, N, batchsize, q = 1, ids = None, batchreplace = 1.0, mask = None, l1 = 0, l2=0, g_var=None):
-  # f: the objective function, e.g. loglikelihood
-  # w: parameters at this time step, len p
-  # c: step size, constant for all dimensions
-  # N: total nbr data vectors
-  # batchsize: nbr of function evals to use in f
-  # q: nbr of repeats for the gradient
-  # batchreplace: percent of ids to replace
-
-  c_adjust = c*np.ones(len(w))
-
-  #pdb.set_trace()
-  p = len(w)
-
-  g = np.zeros( p )
-  h = np.zeros( p )
-
-  if ids is None:
-    ids = np.random.permutation( N )[:batchsize]
-
-  G = np.zeros( (q,p))
-  for j in range(q):
-    # TODO: could put different batch ids here
-    # ids=new_ids
-
-    # TODO: uncomment for perturbed c
-    # if np.random.randn()<0:
-    #   c_adjust *= 1 + np.random.rand()
-    # else:
-    #   c_adjust /= 1 + np.random.rand()
-
-    mask = 2*np.random.binomial(1,0.5,p)-1
-    mask_tilde = 2*np.random.binomial(1,0.5,p)-1
-
-    f_plus  = f( w + c_adjust*mask, ids = ids, l1=l1, l2=l2 )
-    f_minus = f( w - c_adjust*mask, ids = ids, l1=l1, l2=l2 )
-
[…]

### 2015-01-26 code/demo_spsa_lr_mnist.py
minor updates

@@ -17,6 +17,11 @@ if __name__ == "__main__":
   num_training, dimension = X_train.shape
   num_classes = T_train.shape[1]
 
+  N=num_training; D = dimension; K = num_classes
+  
+  l1 = 0
+  l2 = 0
+  
   LR_model = MulticlassLogisticRegression( T_train, X_train, T_test, X_test, l1,  l2 )
   covariance = np.cov( X_train.T )
   f=0.1
@@ -25,7 +30,7 @@ if __name__ == "__main__":
   ccc= np.hstack(tuple([cc]*num_classes))
   #ccc=0.01
 
-  max_iters = 10000
+  max_iters = 50000
   q         = 1
   c         = 1
   N         = len(T_train)
@@ -40,7 +45,7 @@ if __name__ == "__main__":
   gammas = [0.9999]
   moms = [0.0]
   batchsizes = [100]
-  qs = [10]
+  qs = [100]
   result = []
   batchreplaces = [1]
   for c in cs:
@@ -48,7 +53,7 @@ if __name__ == "__main__":
     #alpha = 0.01/N
     
     # for others
-    alpha = 0.01 #e-1 #*3*c/(4)
+    alpha = 0.001 #e-1 #*3*c/(4)
     
     for gamma in gammas:
       for mom in moms:

### 2015-01-26 code/helpers.py

@@ -43,7 +43,7 @@ def onehot( t, K ):
 
 
 def load_mnist():
-  file = gzip.open('../data/mnist.pkl.gz', 'rb')
+  file = gzip.open('./data/mnist.pkl.gz', 'rb')
   (X_train, t_train), (X_valid, t_valid), (X_test, t_test) = cPickle.load(file)
   num_classes = 10
 

### 2015-02-11 papers/uai-2015/draft.tex
ADAgrad RSMProp ADAm
sghmc and sg thermostats mostly

@@
 
-\newcommand{\gradestimate}[2]{\hat{g}_{#1}\lp#2\rp}
-\newcommand{\gradestimaterepeat}[3]{\hat{g}^{#3}_{#1}\lp#2\rp}
-\newcommand{\feval}[2]{y_{#1}^{#2}}
-\newcommand{\cstep}[2]{c_{#1#2}}
-\newcommand{\cstepconst}[1]{c_{#1}}
-\newcommand{\dstep}[2]{\Delta_{#1#2}}
-\newcommand{\dstepconst}[1]{\Delta_{#1}}
-\newcommand{\loglik}[1]{L\lp#1\rp}
-
 \subsection{Spall's method: SPSA}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{eqnarray}
+  h & = & \frac{N}{n} \sum_{i=1}^n \nabla \log p(\bx_i | \thetav)- \sum_{i=1}^N \nabla \log p(\bx_i | \thetav ) \\
+    & = & \frac{N}{n} \sum_{i=1}^n s_i \\
+  s_i & = & \nabla \log p(\bx_i | \thetav) + \frac{1}{N} \nabla p(\thetav) \\
+  \text{V}(h) & = & \frac{N^2}{n^2} \sum_{i=1}^n \text{Var}(s_i) \\
+                & = & \frac{N^2}{n^2} \sum_{i=1}^n \frac{1}{n} \lp \sum_{j=1} (s_i - \bar{s})^2 \rp \\
+                & = & \frac{N^2}{n^2} \sum_{i=1}^n V_s \\
+                & = & \frac{N^2}{n^2} n V_s \\
+                & = & \frac{N^2}{n} V_s 
+  %\text{Var}(\thetav) & = & \frac{\epsilon^2}{4}\text{Var}(h) + \epsilon \\
+  %       \frac{\epsilon^2}{4}\text{Var}(h) & << & \epsilon \\
+  %       \frac{\epsilon N^2}{4 n} V_s & << & 1       
+\end{eqnarray}
+\begin{eqnarray}
+  \text{Var}(\thetav) & = & \frac{\epsilon^2}{4}\text{Var}(h) + \epsilon \\
+         \frac{\epsilon^2}{4}\text{Var}(h) & << & \epsilon \\
+         \frac{\epsilon N^2}{4 n} V_s & << & 1       
+\end{eqnarray}
 
@@
 
+\subsection{ABC Cost Function}
+
+The ABC likelihood is proportional to the convolution between a kernel around the observation statistics $\y$ and the generator of pseudo-statistics $\x$, i.e.
+the simulator.
+Since pseudo-statistics can only be generated, we resort to a Monte Carlo estimate for the ABC likelihood based on $S$ draws from the simulator using the same parameter vector $\thetav$ but different random seeds:
+\begin{eqnarray}
+  \pi\lp \y | \thetav \rp & = &  \int \pi_{\epsvec}\lp \y | \x \rp \pi\lp \x | \thetav \rp d \x \\
+   & \approx & \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}\lp \y | \x^{(s)} \rp
+\end{eqnarray}
+where $\x^{(s)} = \pi\lp \x | \thetav, \omega_s \rp$.
+Note we have passed a unique random seed $\omega_s$ into the simulator, making $\x^{(s)}$ a deterministic function of the simulator with parameters $\{\thetav, \omega_s\}$.
+It will be useful to rewrite this deterministic simulator function as $f$: $\x^{(s)} = f\lp \thetav, \omega_s \rp$, allowing us to rewrite the likelihood as
+\begin{eqnarray}
+  \pi\lp \y | \thetav \rp & \approx & \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}\lp \y | f\lp \thetav, \omega_s \rp \rp
+\end{eqnarray}
+
+For optimization purposes we can search for the maximum a posteriori (MAP) estimator $\hat{\thetav}$ using the log-joint distribution $\hat{\thetav} = \argmax_{\thetav} \mathcal{MAP}$, where $\mathcal{MAP}$ is
+\begin{eqnarray}
+  \mathcal{LJ} &    =    & \log \pi(\thetav) + \log \pi\lp \y | \thetav \rp \\
+                & \approx &\log \pi(\thetav) + \log \sum_{s=1}^S \pi_{\epsvec}\lp \y | f\lp \thetav, \omega_s \rp \rp - \log S
+\end{eqnarray} 
+we can also maximize the lower-bound of $\mathcal{LJ} \geq \mathcal{LB}$, where 
+\begin{eqnarray}
+  \mathcal{LB} & \approx &\log \pi(\thetav) + \sum_{s=1}^S \log  \pi_{\epsvec}\lp \y | f\lp \thetav, \omega_s \rp \rp
+\end{eqnarray} 
+which will be more convenient for optimization for typical kernels (i.e. $\pi_{\epsilon}$ is Gaussian).
+
+The simplest optimization is to set the objective function  $C$ to be $-\mathcal{LJ}$ or $-\mathcal{LB}$ and compute stochastic gradients using 2SPSA with $q$ repeats (negative log-likelihood because we will assume minimization):
+\begin{eqnarray}
+  \frac{\partial C}{\partial \thetav} & = & \frac{1}{q} \sum_{j=1}^{q} \frac{C\lp \thetav + c \Delta_j\rp - C{\lp\thetav - c \Delta_j\rp}}{2 c \Delta_j}
+\end{eqnarray}
+or 1SPSA
+\begin{eqnarray}
+  \frac{\partial C}{\partial \thetav} & = & \frac{1}{q} \sum_{j=1}^{q}  \frac{C\lp \thetav + c \Delta_j\rp - C(\thetav)}{c \Delta_j}
+\end{eqnarray}
+which is less accurate but is less computation if $C(\thetav)$ is computed at each step (TODO: possible algorithm).  TODO: rewrite $\frac{\partial C}{\partial \thetav}$ as $\hat{g}$ or similar.
+
+\subsection{Special Case: Gaussian Kernels}
+When the kernel is known we can use the chain rule to compute the gradients with the first part from the kernel and the second from the variation of the simulator pseudo-statistics, whose gradient can be estimated using SPSA.
+Expanding the likelihood over $J$ statistics, the lower bound on the loglikelihood becomes:
+\begin{eqnarray}
+  \hat{L} & = & \sum_s \log \pi_{\epsilon}\lp \y | f(\thetav, \omega_s)\rp \\
+          & = & \sum_s \log \prod_{j=1}^J \pi_{\epsilon_j}\lp y_j | f_j(\thetav, \omega_s)\rp \\
+          & = & \sum_s \sum_j \log  \pi_{\epsilon_j}\lp y_j | f_j(\thetav, \omega_s)\rp \\
+          & = & \sum_s \sum_j -\frac{1}{2} \frac{\lp y_j - f_j(\thetav, \omega_s)\rp^2}{\epsilon_j^2} - Z_j \\
+\end{eqnarray}
+
+The gradient:
+\begin{eqnarray} 
+  \nabla \hat{L} & = & \sum_s \sum_j \frac{\lp y_j - f_j(\thetav, \omega_s)\rp}{\epsilon_j^2} \cdot \frac{\partial f_j(\thetav, \omega_s)}{\partial \thetav} \\
+                 & = & \sum_s \sum_j e_{js} \nabla_{js}
+\end{eqnarray}
+The first part $e_{js}$ is analytic, the second part $\nabla_{js}$ requires estimation by 2SPSA:
+\begin{eqnarray} 
+  \hat{ \nabla }_{js}  & = & \frac{1}{q} \sum_{r=1}^q \frac{f_j(\thetav + c \Delta_r , \omega_s) - f_j(\thetav - c \Delta_r , \omega_s)}{2 c \Delta_r}
+\end{eqnarray}
+or 1SPSA
+\begin{eqnarray} 
+  \hat{ \nabla }_{js} & = & \frac{1}{q} \sum_{r=1}^q \frac{f_j(\thetav + c \Delta_r , \omega_s) - f_j(\thetav, \omega_s)}{c \Delta_r}
+\end{eqnarray}
+the benefit of 1SPSA in this case is that the first gradients  $e_{js}$ can be computed from $f_j(\thetav, \omega_s)$.
+
+\subsection{SGD, AdaGrad, RMSProp, AdaM}
+
+With our stochastic gradient in hand, we can make use of robust stochastic gradient rules, including recent advances from machine learning.
+
+\subsubsection{SGD}
+The basic update rule is at time $t$ for minimization is
+\begin{eqnarray}
+  \Delta\thetav & = & - \alpha_t \hat{g_t}
+\end{eqnarray} 
+with conditions $\sum_{t=1}^{\infty} \alpha_t = \infty$ and $\sum_{t=1}^{\infty} \alpha_t^2 < \infty$ required for convergence.
+An additional momentum parameter $\beta_1$ can be added to improve convergence
+\begin{eqnarray}
+  \gmom & = & \beta_1 \gmom + (1-\beta_1)*\hat{g_t} \\
+  \Delta\thetav & = & - \alpha_t \gmom
+\end{eqnarray}
+
+\subsubsection{AdaGrad}
+AdaGrad uses the total cumulative element-wise square of the gradients to automatically both decrease $\alpha_t$ and scale the learning rates for each dimension: 
+\begin{eqnarray}
+  \gmom    & = & \beta_1 \gmom + (1-\beta_1)*\hat{g_t} \\
+  \gsquare & = & \sum_{t'=1}^t \hat{g_t}^2 \\
+  \Delta\thetav & = & - \alpha_t \gmom / \sqrt{\gsquare}
+\end{eqnarray}
+
+\subsubsection{RMSProp}
+\begin{eqnarray}
+  \gmom    & = & \beta_1 \gmom + (1-\beta_1)*\hat{g_t} \\
+  \gsquare & = & \beta_2 \gsquare + (1-\beta_2)*\hat{g_t}^2 \\
+  \Delta\thetav & = & - \alpha_t \gmom / \sqrt{\gsquare}
+\end{eqnarray}
+
+\subsubsection{AdaM}
+\begin{eqnarray}
+  \gmom    & = & \beta_1 \gmom + (1-\beta_1)*\hat{g_t} \\
+  \gsquare & = & \beta_2 \gsquare + (1-\beta_2)*\hat{g_t}^2 \\
+  \gamma & = & \frac{ \sqrt{1-(1-\beta_2)^t}}{(1-(1-\beta_1)^t)} \\
+  \Delta\thetav & = & - \alpha_t \gamma \gmom / \sqrt{\gsquare}
+\end{eqnarray}
+
+\section{Applying SGLD to ABC}
+
+There are two main sources of randomness in the ABC gradient computation: the random perturbation mask and the random seed of the simulator model, which controls the simulator noise.
+This noise can be eliminated by using common random numbers \cite{spall-crn}.
+On the other hand, it is not clear that this is the best strategy, as the noise from the masks at a fixed seed may be higher than
+
+\begin{eqnarray}
+  V_{\gradestimaterepeat{t}{\thetav_t}{j}} & = & \frac{1}{q-1} \sum_{j=1}^q (\gradestimaterepeat{t}{\thetav_t}{j} - \gradestimate{t}{\thetav_t})^2 \\
+  V_{\gradestimate{t}{\thetav_t}} & = & \frac{1}{q-1} V_{\gradestimaterepeat{t}{\thetav_t}{j}} \\
+  \frac{\epsilon^2}{4 }V_{\gradestimate{t}{\thetav_t}} & = & \frac{\epsilon^2}{4 q }V_{\gradestimaterepeat{t}{\thetav_t}{j}} \\
+                                                       & << & \epsilon \\
+                    \frac{\epsilon}{4 (q-1) }V_{\gradestimaterepeat{t}{\thetav_t}{j}}  & << & 1                                  
+\end{eqnarray}
+
+Check for $g(\thetav) + h(\thetav) + \eta$ that $h(\thetav) + \eta \approx \eta$
+
+Also $\theta_{t+1} = \theta_t + \frac{\epsilon}{2}\gradestimate{t}{\thetav_t} + \sqrt{\epsilon}\mathcal{N}(0,1)$
+
 \subsection{Statistical Tests for Gradients}
+
+From Byrd:
+\begin{eqnarray}
+\frac{||V_{\gradestimaterepeat{t}{\thetav_t}{j}} ||_1}{q} & \leq r^2 || \hat{g}_t ||_2^2 \\
+\hat{q} & = & \frac{||V_{\gradestimaterepeat{t}{\thetav_t}{j}} ||_1}{r^2 || \hat{g}_t ||_2^2}
+\end{eqnarray}
+
+TODO:
+\begin{itemize}
+  \item Check that $q$ gradients are normal.  Especially as $q$ increases.
+  \item Merge Byrd's test for $\hat{q}$ and condition on SGLD-ABC.
+  \item Variance check for increasing $q$ can also be used to stop optimization since the variance is less than the noise in the gradients.
+\end{itemize}
 
@@
 \subsection{Logistic Regression with Stochastic Gradients}
+
+\subsection{Optimization of Blowfly Dynamics}
+
+\subsubsection{Using AdaM}
+
+\begin{figure}[h]
+\vspace{1in}
+\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_post_series.png}
+\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_post_series_sgld5k.png}
+\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_post.png}
+\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_post_sgld5k.png}
+\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_postpred_series.png}
+\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_postpred_series_sgld5k.png}
+\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_postpred.png}
+\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_postpred_sgld5k.png}
+\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_noise.png}
+\includegraphics[width=0.45\columnwidth]{./images/blowfly_adam_noise_sgld5k.png}
+\caption{Optimizing Blowfly using $q=5$, $\epsilon = 0.01$, $\beta_1=0.75$,$\beta_2=0.9$,$\alpha=0.1$ and $c=1$.  Left: Optimization using AdaM, Right: ``SGLD'' with estimated noise. }
+\end{figure}
+
+\subsubsection{Using AdaM with ``SGLD" injected noise}
 

### 2015-02-11 code/demo_blowfly_graddescent.py

@@ -1,0 +1,231 @@
+from abcpy.factories import *
+from abcpy.problems.blowfly.blowfly import *
+from sa_algorithms import *
+from scipy import stats as spstats
+import pylab as pp
+
+problem_params = default_params()
+problem_params["std_log_tau"] = 0.2
+problem_params["tau_is_log_normal"] = True
+problem_params["epsilon"] = 0.1 #0.01 #10*np.array([1.0,1.0,0.5,0.5,0.5,0.1,0.01,0.5,2.0,1])
+#problem_params["epsilon"] = 0.1*np.array([0.1,0.1,0.1,0.1,0.1,0.01,0.01,0.1,1.0,1.0])
+problem_params["epsilon"] = 0.01*np.array([  0.91064006,   0.12437852,   1.06735857,   1.70135243,1.10402222,   0.22966667, 0.08973333,   1.28127273, 10.        ,   9.        ])
+
+problem_params["epsilon"] = 2*np.array([  0.05,   0.05,   0.05,   0.05, 0.05,  0.02, 0.02,  0.05, 0.01        ,  3.0        ])
+
+problem_params["blowfly_filename"] = "./data/blowfly.txt"
+#problem_params["blowfly_filename"] = "./data/bf2.txt"
+problem = BlowflyProblem( problem_params, force_init = True )
+
+state_params = state_params_factory.scrape_params_from_problem( problem, S=5 )
+
+# set is_marginal to true so using only "current" state will force a re-run
+mcmc_params  = mcmc_params_factory.scrape_params_from_problem( problem, type="mh", is_marginal = True, nbr_samples = 1 )
+
+# ignore algo, just doing this to get state object
+algo_params = { "modeling_approach"  : "kernel",
+                "observation_groups" : problem.get_obs_groups(),
+                "state_params"       : state_params,
+                "mcmc_params"        : mcmc_params,
+                "algorithm"          : "model_mcmc"
+              }
+algo, model, state  = algo_factory.create_algo_and_state( algo_params )
+recorder     = recorder_factory.create_recorder( {} )
+
+class LikelihoodFree( object ):
+  def __init__( self, problem, model, state, recorder ):
+    self.problem = problem
+    self.model = model
+    self.state = state
+    self.recorder = recorder
+    self.tau   = 15
+    
[…]

### 2015-02-11 code/demo_blowfly_ld.py

@@ -1,16 +1,23 @@
 from abcpy.factories import *
 from abcpy.problems.blowfly.blowfly import *
 from sa_algorithms import *
-
+from scipy import stats as spstats
 import pylab as pp
 
 problem_params = default_params()
-problem_params["epsilon"] = 0.01 #10*np.array([1.0,1.0,0.5,0.5,0.5,0.1,0.01,0.5,2.0,1])
+problem_params["std_log_tau"] = 0.2
+problem_params["tau_is_log_normal"] = True
+problem_params["epsilon"] = 0.1 #0.01 #10*np.array([1.0,1.0,0.5,0.5,0.5,0.1,0.01,0.5,2.0,1])
+#problem_params["epsilon"] = 0.1*np.array([0.1,0.1,0.1,0.1,0.1,0.01,0.01,0.1,1.0,1.0])
+problem_params["epsilon"] = 0.01*np.array([  0.91064006,   0.12437852,   1.06735857,   1.70135243,1.10402222,   0.22966667, 0.08973333,   1.28127273, 10.        ,   9.        ])
+
+problem_params["epsilon"] = 2*np.array([  0.05,   0.05,   0.05,   0.05, 0.05,  0.02, 0.02,  0.05, 0.01        ,  3.0        ])
+
 problem_params["blowfly_filename"] = "./data/blowfly.txt"
 #problem_params["blowfly_filename"] = "./data/bf2.txt"
 problem = BlowflyProblem( problem_params, force_init = True )
 
-state_params = state_params_factory.scrape_params_from_problem( problem )
+state_params = state_params_factory.scrape_params_from_problem( problem, S=1 )
 
 # set is_marginal to true so using only "current" state will force a re-run
 mcmc_params  = mcmc_params_factory.scrape_params_from_problem( problem, type="mh", is_marginal = True, nbr_samples = 1 )
@@ -38,17 +45,27 @@ class LikelihoodFree( object ):
   
   def add_prediction( self, x_statistics ):
     if self.moving_average_statistics is None:
-      self.moving_average_statistics = x_statistics.copy()
+      self.moving_average_statistics = x_statistics.mean(0).copy()
       self.count = 1
     else:
-      self.moving_average_statistics = (float( self.count)/float(self.count+1))*self.moving_average_statistics + (float( 1)/float(self.count+1))*x_statistics
[…]

### 2015-02-11 code/demo_sghmc_fig1.py

@@ -1,0 +1,45 @@
+import numpy as np
+import pylab as pp
+
+from sghmc_problems import *
+from sg import *
+
+
+    
+    
+if __name__ == "__main__":
+  pp.close('all')
+  T             = 5
+  L             = 20
+  epsilon       = 0.1
+  current_q     = np.random.randn()
+  mh_correction = True
+  step_method   = leapfrog_step
+  
+  injected_gradient_noise_std = 2.0
+  
+  problem = generate_sghmc_fig1_problem(noise=injected_gradient_noise_std)
+  problem["other_params"]={"C":6,"V":injected_gradient_noise_std**2}
+
+  Qleap,Pleap = run_hmc_with_problem( T, problem, leapfrog_step, mh_correction, current_q, epsilon, L  )
+  Qleap_no_mh,Pleap_no_mh = run_hmc_with_problem( T, problem, leapfrog_step, False, current_q, epsilon, L  )
+  Qsghmc,Psghmc = run_hmc_with_problem( T, problem, sghmc_step, False, current_q, epsilon, L  )
+  Qsgnht,Psgnht = run_hmc_with_problem( T, problem, sgnht_step, False, current_q, epsilon, L  )
+  #Qmeul,Pmeul = run_hmc_with_problem( T, problem, mod_euler_step, mh_correction, current_q, epsilon, L  )
+  #Qeul,Peul = run_hmc_with_problem( T, problem, euler_step, mh_correction, current_q, epsilon, L  )
+  
+  pp.figure(1)
+  pp.clf()
+  pp.plot( problem["theta_range"](), problem["true_posterior"](problem["theta_range"]()),  'k--', lw=4)
+  pp.hist( Qleap, 20,histtype='step', normed=True, alpha=0.75, linewidth=3 )
+  pp.hist( Qleap_no_mh, 20,histtype='step', normed=True, alpha=0.75, linewidth=3 )
+  pp.hist( Qsghmc, 20,histtype='step', normed=True, alpha=0.75, linewidth=3 )
+  pp.hist( Qsgnht, 20,histtype='step', normed=True, alpha=0.75, linewidth=3 )
+  #pp.hist( Qmeul, 50,histtype='step',normed=True, alpha=0.5, linewidth=4 )
+  #pp.hist( Qeul, 50,histtype='step',normed=True, alpha=0.5, linewidth=4 )
[…]

### 2015-02-11 code/demo_sgld_fig1.py

@@ -1,0 +1,44 @@
+import numpy as np
+import pylab as pp
+
+from sghmc_problems import *
+from sg import *
+
+
+    
+    
+if __name__ == "__main__":
+  pp.close('all')
+  T             = 5000
+  L             = 2
+  epsilon       = 0.05
+  current_q     = 0*np.random.randn()*np.ones(2)
+  mh_correction = True
+  step_method   = leapfrog_step
+  
+  injected_gradient_noise_std = 4.0
+  
+  problem = generate_sgld_fig1_problem()
+  
+
+  Qleap,Pleap = run_hmc_with_problem( T, problem, leapfrog_step, mh_correction, current_q, epsilon, L  )
+  #Qleap_no_mh,Pleap_no_mh = run_hmc_with_problem( T, problem, leapfrog_step, False, current_q, epsilon, L  )
+  #Qmeul,Pmeul = run_hmc_with_problem( T, problem, mod_euler_step, mh_correction, current_q, epsilon, L  )
+  #Qeul,Peul = run_hmc_with_problem( T, problem, euler_step, mh_correction, current_q, epsilon, L  )
+  
+  pp.figure(1)
+  pp.clf()
+  
+  pp.plot( Qleap[:,0], Qleap[:,1], '.')
+  pp.axis( [-1.5,2.5,-3,3])
+  #pp.plot( problem["theta_range"](), problem["true_posterior"](problem["theta_range"]()),  'k--', lw=4)
+  #pp.hist( Qleap, 20,histtype='step', normed=True, alpha=0.5, linewidth=4 )
+  #pp.hist( Qleap_no_mh, 20,histtype='step', normed=True, alpha=0.5, linewidth=4 )
+  #pp.hist( Qmeul, 50,histtype='step',normed=True, alpha=0.5, linewidth=4 )
+  #pp.hist( Qeul, 50,histtype='step',normed=True, alpha=0.5, linewidth=4 )
+  #pp.legend( ["True","Leap + mh","Leap no mh"])
+  #pp.legend( ["True","Leap","mEuler"])
+  #pp.legend( ["True","Leap","mEuler","euler"])
+  pp.show()
+  
+  

### 2015-02-11 code/demo_sgnht_fig1_double_well.py

@@ -1,0 +1,199 @@
+import numpy as np
+import pylab as pp
+
+from sghmc_problems import *
+from sgnht_problems import *
+from sg import *
+
+def run_sghmc_with_problem( problem, T, h, A, Bhat, current_q, current_p = None, current_xi = None ):
+  # extract problem functions
+  U  = problem["U"]
+  K  = problem["K"]
+  dU = problem["dU"]
+  
+  # initialize
+  if current_p is None:
+    p = np.random.randn(1)
+  else:
+    p = current_p
+  current_p = None
+  q = current_q
+  
+  if current_xi is None:
+    xi = A
+  else:
+    xi = current_xi
+    
+  P  = [p]
+  Q  = [q]
+  XI = [xi]
+  for t in range(T):
+    
+    
+    q = q + p*h
+    
+    grad_U = dU( q )
+    p = p - A*p*h - grad_U*h + np.sqrt(2.0*(A-Bhat)*h)*np.random.randn()
+    
+    
+    # update xi if it wasn't initialize
+    if current_xi is None:
+      xi = xi + (p**2 - 1.0)*h
+  
+    P.append( p )
+    Q.append( q )
+    XI.append( xi )
+    
+  return np.squeeze(np.array(Q)), np.squeeze(np.array(P)), np.squeeze(np.array(XI))
+  
+def run_sgnht_with_problem( problem, T, h, A, current_q, current_p = None, current_xi = None ):
+  # extract problem functions
+  U  = problem["U"]
+  K  = problem["K"]
+  dU = problem["dU"]
+  
+  # initialize
+  if current_p is None:
+    p = np.random.randn(1)
+  else:
+    p = current_p
+  current_p = None
+  q = current_q
+  
+  if current_xi is None:
+    xi = A
+  else:
+    xi = current_xi
+    
+  P  = [p]
+  Q  = [q]
+  XI = [xi]
+  for t in range(T):
+    grad_U = dU( q )
+    p = p - xi*p*h - grad_U*h + np.sqrt(2.0*A*h)*np.random.randn()
+    q = q + p*h
+    
+    # update xi if it wasn't initialize
[…]

### 2015-02-11 code/demo_sgnht_fig2.py

@@ -1,0 +1,267 @@
+import numpy as np
+import pylab as pp
+
+from sghmc_problems import *
+from sgnht_problems import *
+from sg import *
+
+def run_sghmc_with_problem( problem, T, h, A, Bhat, current_q, current_p = None, current_xi = None ):
+  # extract problem functions
+  U  = problem["U"]
+  K  = problem["K"]
+  dU = problem["dU"]
+  
+  # initialize
+  if current_p is None:
+    p = np.random.randn(1)
+  else:
+    p = current_p
+  current_p = None
+  q = current_q
+  
+  if current_xi is None:
+    xi = A
+  else:
+    xi = current_xi
+    
+  P  = [p]
+  Q  = [q]
+  XI = [xi]
+  for t in range(T):
+    
+    
+    q = q + p*h
+    
+    grad_U = dU( q )
+    p = p - A*p*h - grad_U*h + np.sqrt(2.0*(A-Bhat)*h)*np.random.randn()
+    
+    
+    # update xi if it wasn't initialize
+    if current_xi is None:
+      xi = xi + (p**2 - 1.0)*h
+  
+    P.append( p )
+    Q.append( q )
+    XI.append( xi )
+    
+  return np.squeeze(np.array(Q)), np.squeeze(np.array(P)), np.squeeze(np.array(XI))
+  
+def run_sgnht_with_problem( problem, T, h, A, current_q, current_p = None, current_xi = None ):
+  # extract problem functions
+  U  = problem["U"]
+  K  = problem["K"]
+  dU = problem["dU"]
+  
+  # initialize
+  if current_p is None:
+    p = np.random.randn(1)
+  else:
+    p = current_p
+  current_p = None
+  q = current_q
+  
+  if current_xi is None:
+    xi = A
+  else:
+    xi = current_xi
+    
+  P  = [p]
+  Q  = [q]
+  XI = [xi]
+  for t in range(T):
+    grad_U = dU( q )
+    p = p - xi*p*h - grad_U*h + np.sqrt(2.0*A*h)*np.random.randn()
[…]

### 2015-02-11 code/demo_sgnht_fig2right.py

@@ -1,0 +1,258 @@
+import numpy as np
+import pylab as pp
+
+from sghmc_problems import *
+from sgnht_problems import *
+from sg import *
+
+def sample_density( samples, bins ):
+  bw = bins[1]-bins[0]
+  cnts, bins, patches = pp.hist( samples, bins )
+  pdf = cnts / (bw*float(len(samples)))
+  return pdf
+    
+def run_sghmc_with_problem( problem, T, h, A, Bhat, current_q, current_p = None, current_xi = None ):
+  # extract problem functions
+  U  = problem.U
+  K  = problem.K
+  dU = problem.dU
+  
+  # initialize
+  if current_p is None:
+    p = np.random.randn(2)
+  else:
+    p = current_p
+  current_p = None
+  q = current_q
+  
+  if current_xi is None:
+    xi = A
+  else:
+    xi = current_xi
+    
+  P  = [p]
+  Q  = [q]
+  XI = [xi]
+  for t in range(T):
+    
+    
+    q = q + p*h
+    
+    grad_U = dU( q )
+    p = p - A*p*h - grad_U*h + np.sqrt(2.0*(A-Bhat)*h)*np.random.randn()
+    
+    
+    # update xi if it wasn't initialize
+    if current_xi is None:
+      xi = xi + (0.5*np.sum( p**2 ) - 1.0)*h
+  
+    P.append( p )
+    Q.append( q )
+    XI.append( xi )
+    
+  return np.squeeze(np.array(Q)), np.squeeze(np.array(P)), np.squeeze(np.array(XI))
+  
+def run_sgnht_with_problem( problem, T, h, A, current_q, current_p = None, current_xi = None ):
+  # extract problem functions
+  U  = problem.U
+  K  = problem.K
+  dU = problem.dU
+  
+  # initialize
+  if current_p is None:
+    p = np.random.randn(2)
+  else:
+    p = current_p
+  current_p = None
+  q = current_q
+  
+  if current_xi is None:
+    xi = A
+  else:
+    xi = current_xi
+    
+  P  = [p]
+  Q  = [q]
+  XI = [xi]
+  for t in range(T):
+    grad_U = dU( q )
[…]

### 2015-02-11 code/demo_spsa_lr_mnist.py

@@ -30,30 +30,30 @@ if __name__ == "__main__":
   ccc= np.hstack(tuple([cc]*num_classes))
   #ccc=0.01
 
-  max_iters = 50000
+  max_iters = 5000
   q         = 1
   c         = 1
   N         = len(T_train)
   batchsize = 10
   alpha     = 0.1
   gamma     = 0.9
-  mom_beta1 = 0.9 # on gradient
-  mom_beta2 = 0.9 # on gradient_squared
+  mom_beta1 = 0.0 # on gradient
+  mom_beta2 = 0.0 # on gradient_squared
   
   #cs = [0.5,0.1,0.2,0.3]
   cs = [ccc]
   gammas = [0.9999]
   moms = [0.0]
   batchsizes = [100]
-  qs = [100]
+  qs = [10]
   result = []
   batchreplaces = [1]
   for c in cs:
     # for "grad"
-    #alpha = 0.01/N
+    alpha = 0.01/N
     
     # for others
-    alpha = 0.001 #e-1 #*3*c/(4)
+    #alpha = 0.01 #e-1 #*3*c/(4)
     
     for gamma in gammas:
       for mom in moms:
@@ -75,11 +75,12 @@ if __name__ == "__main__":
                                 "gamma_eps":0.9999,
                                 "mom_beta1":mom_beta1,
                                 "mom_beta2":mom_beta2,
-                                "update_method":"adam",
+                                "update_method":"grad",
                                 "batchreplace":batchreplace, 
                                 "diag_add":0.01}
               
-              wout, errors = spall( w, spall_params )
+              #wout, errors = spall( w, spall_params )
+              wout, errors = spall_sgld( w, spall_params )
               #wout, errors = spall_with_hessian( w, spall_params )
       
               result.append({"c":c,"alpha":alpha,"gamma":gamma, "mom":mom, "batchsize":batchsize,"q":q,"errors":errors,"w":wout})

### 2015-02-11 code/demo_trajectories.py

@@ -1,0 +1,622 @@
+import numpy as np
+import pylab as pp
+
+def U(q):
+  return 0.5*q*q
+  
+def K(p):
+  return 0.5*p*p
+
+def hamiltonian( q, p ):
+  return U(q) + K(p)
+  
+def grad_U( q ):
+  return q
+  
+def grad_K( p ):
+  return p
+
+def Hamiltonian_dynamics_with_noised_gradient(U, grad_U, epsilon, L, current_q, p = None, noise_in_u=0):
+  # %% Hamiltonian dynamics with noised gradient
+  # x = xstart;
+  # p = pstart;
+  q = current_q
+  if p is None:
+    p = np.random.randn()
+  nstep=L[0]
+  niter=L[1]
+  
+  P = [p]
+  Q = [q]
+  
+  # xs = zeros( nstep, 1 );
+  # ys = zeros( nstep, 1 );
+  # % do leapfrog
+  # for i = 1 : nstep
+  #     for j = 1: niter
+  print "p half-step"
+  
+  #p = p - epsilon*( grad_U( q ) + noise_in_u*np.random.randn() ) / 2.0
+  for i in range(nstep):
+    p = p - epsilon*( grad_U(q)+noise_in_u*np.random.randn() )/2.0
+    for j in range(niter):
+      print "q full-step"
+      q = q + epsilon*p
+      
+      
+      if j+1 < L:
+        print "p full-step"
+        #P.append( p - epsilon*( grad_U(q)+noise_in_u*np.random.randn() )/2.0)
+        p = p - epsilon*( grad_U(q)+noise_in_u*np.random.randn() )
+    
+    p = p - epsilon*( grad_U( q ) + noise_in_u*np.random.randn() ) / 2.0   
+    P.append(p) 
+    Q.append(q)   
+      
+  return q, p,np.array(Q), np.array(P)
+  # end
+  # plot( xs, ys, 'rv', 'MarkerSize', 3 );
+  # hold on;
+  
+def Hamiltonian_dynamics_with_noised_gradient_bad(U, grad_U, epsilon, L, current_q, p = None, noise_in_u=0):
+  # %% Hamiltonian dynamics with noised gradient
+  # x = xstart;
+  # p = pstart;
+  q = current_q
+  if p is None:
+    p = np.random.randn()
+  nstep=L[0]
+  niter=L[1]
+  
+  P = [p]
+  Q = [q]
+  
[…]

### 2015-02-11 code/hmc_demo_sample.py

@@ -1,0 +1,469 @@
+import numpy as np
+import pylab as pp
+
+def U(q):
+  return 0.5*q*q
+  
+def K(p):
+  return 0.5*p*p
+
+def hamiltonian( q, p ):
+  return U(q) + K(p)
+  
+def grad_U( q ):
+  return q
+  
+def grad_K( p ):
+  return p
+
+def Euler_step( U, grad_U, epsilon, L, current_q, p = None, noise_in_u=0 ):
+  q = current_q
+  if p is None:
+    p = np.random.randn()
+  # else:
+  #   p = 1
+    
+  current_p = p
+  
+  P = [p]
+  Q = [q]
+  
+  old_p = p
+  old_q = q
+  # half step for p
+  #p = p - epsilon*grad_U(q)/2.0
+  
+  for i in range(L):
+    # full step for position
+    q = q + epsilon*old_p
+    
+    p = old_p - epsilon*( grad_U(old_q)+noise_in_u*np.random.randn() )
+    
+    
+    old_p = p
+    old_q = q
+    
+    Q.append(q)
+    P.append(p)
+  
+  # not needed, but negate momentum to preserve reversibility
+  p = -p
+  
+  current_U = U( current_q )
+  current_K = K( current_p )
+  proposed_U = U( q )
+  proposed_K = K( p )
+  
+  if np.random.rand() < np.exp( current_U - proposed_U + current_K - proposed_K ):
+    return q, p,np.array(Q), np.array(P)
+  else:
+    return current_q, current_p,np.array(Q), np.array(P)
+
+def mod_Euler_step( U, grad_U, epsilon, L, current_q, p = None, noise_in_u=0 ):
+  q = current_q
+  if p is None:
+    p = np.random.randn()
+  # else:
+  #   p = 1 
+    
+  current_p = p
+  
+  P = [p]
+  Q = [q]
+  
+  # half step for p
+  #p = p - epsilon*grad_U(q)/2.0
+  
+  for i in range(L):
+    # full step for position
+    q = q + epsilon*p
+    
+    p = p - epsilon*( grad_U(q)+noise_in_u*np.random.randn() )
+    
+    
+    Q.append(q)
+    P.append(p)
+  
+  # not needed, but negate momentum to preserve reversibility
+  p = -p
+  
+  current_U = U( current_q )
[…]

### 2015-02-11 code/hmc_demo_with_noise.py

@@ -1,0 +1,685 @@
+import numpy as np
+import pylab as pp
+import pdb
+
+def U(q):
+  return 0.5*q*q
+  
+def K(p):
+  return 0.5*p*p
+
+def hamiltonian( q, p ):
+  return U(q) + K(p)
+  
+def grad_U( q ):
+  return q
+  
+def grad_K( p ):
+  return p
+
+def Hamiltonian_dynamics_with_noised_gradient(U, grad_U, epsilon, L, current_q, p = None, noise_in_u=0):
+  # %% Hamiltonian dynamics with noised gradient
+  # x = xstart;
+  # p = pstart;
+  q = current_q
+  if p is None:
+    p = np.random.randn()
+  nstep=L
+  niter=1
+  
+  P = [p]
+  Q = [q]
+  
+  # xs = zeros( nstep, 1 );
+  # ys = zeros( nstep, 1 );
+  # % do leapfrog
+  # for i = 1 : nstep
+  #     for j = 1: niter
+  print "p half-step"
+  p = p - epsilon*( grad_U(q)+noise_in_u*np.random.randn() )/2.0
+  #p = p - epsilon*( grad_U( q ) + noise_in_u*np.random.randn() ) / 2.0
+  for i in range(L):
+    #for j in range(nstep):
+      print "q full-step"
+      q = q + epsilon*p
+      Q.append(q)
+      
+      if i+1 < L:
+        print "p full-step"
+        P.append( p - epsilon*( grad_U(q)+noise_in_u*np.random.randn() )/2.0)
+        p = p - epsilon*( grad_U(q)+noise_in_u*np.random.randn() )
+    #if i+1 < L:
+    #  p = p - epsilon*( grad_U( q ) + noise_in_u*np.random.randn() ) / 2.0
+  p = p - epsilon*( grad_U( q ) + noise_in_u*np.random.randn() ) / 2.0   
+  P.append(p)    
+      
+  return q, p,None,np.array(Q), np.array(P),None
+  # end
+  # plot( xs, ys, 'rv', 'MarkerSize', 3 );
+  # hold on;
+  
+def Hamiltonian_dynamics_with_noised_gradient_bad(U, grad_U, epsilon, L, current_q, p = None, noise_in_u=0):
+  # %% Hamiltonian dynamics with noised gradient
+  # x = xstart;
+  # p = pstart;
+  q = current_q
+  if p is None:
[…]

### 2015-02-11 code/neal_hmc_demo.py

@@ -1,0 +1,229 @@
+import numpy as np
+import pylab as pp
+
+def U(q):
+  return 0.5*q*q
+  
+def K(p):
+  return 0.5*p*p
+
+def hamiltonian( q, p ):
+  return U(q) + K(p)
+  
+def grad_U( q ):
+  return q
+  
+def grad_K( p ):
+  return p
+
+def Euler_step( U, grad_U, epsilon, L, current_q, p = None ):
+  q = current_q
+  if p is None:
+    p = np.random.randn()
+  else:
+    p = 1 
+    
+  current_p = p
+  
+  P = [p]
+  Q = [q]
+  
+  old_p = p
+  old_q = q
+  # half step for p
+  #p = p - epsilon*grad_U(q)/2.0
+  
+  for i in range(L):
+    p = old_p - epsilon*grad_U(q)
+    
+    # full step for position
+    q = q + epsilon*old_p
+    
+    old_p = p
+    #old_q = q
+    
+    Q.append(q)
+    P.append(p)
+  
+  # not needed, but negate momentum to preserve reversibility
+  p = -p
+  
+  current_U = U( current_q )
+  current_K = K( current_p )
+  proposed_U = U( q )
+  proposed_K = K( p )
+  
+  if np.random.rand() < np.exp( current_U - proposed_U + current_K - proposed_K ):
+    return q, p,np.array(Q), np.array(P)
+  else:
+    return current_q, current_p,np.array(Q), np.array(P)
+
+def mod_Euler_step( U, grad_U, epsilon, L, current_q, p = None ):
+  q = current_q
+  if p is None:
+    p = np.random.randn()
+  else:
+    p = 1 
+    
+  current_p = p
+  
+  P = [p]
+  Q = [q]
+  
+  # half step for p
+  #p = p - epsilon*grad_U(q)/2.0
+  
+  for i in range(L):
+    p = p - epsilon*grad_U(q)
+    
+    # full step for position
+    q = q + epsilon*p
+    
+    Q.append(q)
+    P.append(p)
+  
+  # not needed, but negate momentum to preserve reversibility
+  p = -p
+  
+  current_U = U( current_q )
+  current_K = K( current_p )
+  proposed_U = U( q )
+  proposed_K = K( p )
+  
[…]

### 2015-02-11 code/sa_algorithms.py

@@ -82,6 +82,37 @@ def sgradient( g, w, N, batchsize, q = 1, ids = None, l1 = 0, l2=0):
 #   #pdb.set_trace()
 #   return g, h, ids
 
+def finite_difference_abc_gradient( f, w, c, seed = None ):
+  # f: the objective function, e.g. loglikelihood
+  # w: parameters at this time step, len p
+  # c: step size, constant for all dimensions
+  # q: nbr of repeats for the gradient
+
+  p = len(w)
+  g = np.zeros( p )
+  
+  for i in range(p):
+    
+    mask = np.zeros(p)
+    mask[i] = 1.0
+    
+    #pdb.set_trace()
+    f_plus  = f( w + c*mask, seed = seed )
+    f_minus = f( w - c*mask, seed = seed )
+
+    g[i] = (f_plus-f_minus)/(2*c)
+
+    if np.isinf( f_plus ) or np.isinf( f_minus ):
+      pdb.set_trace()
+
+  h = None
+  if seed is not None:
+    new_seed = seed + 1
+  else:
+    new_seed = None
+  return g, new_seed
+
+
 def spsa_abc_gradient( f, w, c, q = 1, seed = None, mask = None, hessian=False, c_tilde=None ):
   # f: the objective function, e.g. loglikelihood
   # w: parameters at this time step, len p
@@ -95,19 +126,24 @@ def spsa_abc_gradient( f, w, c, q = 1, seed = None, mask = None, hessian=False, 
   if hessian:
     h = np.zeros( (p,p))
   G = np.zeros( (q,p))
+  
+  masks = 2*np.random.binomial(1,0.5,(q,p))-1
+  
   for j in range(q):
     # TODO: uncomment for perturbed c
-    # if np.random.randn()<0:
-    #   c_adjust *= 1 + np.random.rand()
-    # else:
-    #   c_adjust /= 1 + np.random.rand()
+    #if np.random.randn()<0:
+    #  c_adjust *= 1 + 0.25*np.random.rand()
+    #else:
+    #  c_adjust /= 1 + 0.25*np.random.rand()
 
-    mask = 2*np.random.binomial(1,0.5,p)-1
+    mask = masks[j]
 
     #pdb.set_trace()
     f_plus  = f( w + c_adjust*mask, seed = seed )
     f_minus = f( w - c_adjust*mask, seed = seed )
 
+    #seed += 1
[…]

### 2015-02-11 code/sg.py

@@ -1,0 +1,313 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+import pdb
+
+def euler_step( U, grad_U, K, epsilon, L, current_q, p, mh_correction, other_params = None ):
+  q = current_q
+  if p is None:
+    p = np.random.randn()
+    
+  current_p = p
+  
+  P = [p]
+  Q = [q]
+  
+  old_p = p
+  old_q = q
+  # half step for p
+  #p = p - epsilon*grad_U(q)/2.0
+  
+  for i in range(L):
+    p = old_p - epsilon*grad_U(q)
+    
+    # full step for position
+    q = q + epsilon*old_p
+    
+    old_p = p
+    #old_q = q
+    
+    Q.append(q)
+    P.append(p)
+  
+  # not needed, but negate momentum to preserve reversibility
+  p = -p
+  
+  current_U = U( current_q )
+  current_K = K( current_p )
+  proposed_U = U( q )
+  proposed_K = K( p )
+  
+  if mh_correction:
+    if np.random.rand() < np.exp( current_U - proposed_U + current_K - proposed_K ):
+      return q, p,np.array(Q), np.array(P)
+    else:
+      return current_q, current_p, np.array(Q), np.array(P)
+  else:
+    return q, p, np.array(Q), np.array(P)
+    
+def mod_euler_step( U, grad_U, K, epsilon, L, current_q, p, mh_correction, other_params = None ):
+  q = current_q
+  if p is None:
+    p = np.random.randn()
+    
+  current_p = p
+  
+  P = [p]
+  Q = [q]
+  
+  # half step for p
+  #p = p - epsilon*grad_U(q)/2.0
+  
+  for i in range(L):
+    p = p - epsilon*grad_U(q)
+    
+    # full step for position
+    q = q + epsilon*p
+    
+    Q.append(q)
+    P.append(p)
+  
+  # not needed, but negate momentum to preserve reversibility
+  p = -p
+  
+  current_U = U( current_q )
+  current_K = K( current_p )
+  proposed_U = U( q )
+  proposed_K = K( p )
+  
+  if mh_correction:
+    if np.random.rand() < np.exp( current_U - proposed_U + current_K - proposed_K ):
+      return q, p,np.array(Q), np.array(P)
+    else:
+      return current_q, current_p, np.array(Q), np.array(P)
+  else:
[…]

### 2015-02-11 code/sghmc_problems.py

@@ -1,0 +1,169 @@
+import numpy as np
+import pylab as pp
+from scipy import stats
+import pdb
+
+def logsumexp(x,dim=0):
+    """Compute log(sum(exp(x))) in numerically stable way."""
+    #xmax = x.max()
+    #return xmax + log(exp(x-xmax).sum())
+    if dim==0:
+        xmax = x.max(0)
+        return xmax + np.log(np.exp(x-xmax).sum(0))
+    elif dim==1:
+        xmax = x.max(1)
+        return xmax + np.log(np.exp(x-xmax[:,np.newaxis]).sum(1))
+    else: 
+        raise 'dim ' + str(dim) + 'not supported'
+        
+def generate_sgld_fig1_problem( theta1 = 0, theta2 = 1, var1 = 10, var2 = 1, varx = 2, N = 100, batchsize = 1, seed = 1 ):
+  
+  std1 = np.sqrt(var1)
+  std2 = np.sqrt(var2)
+  stdx = np.sqrt(varx)
+  
+  # 0/1 assignments to first or second gaussian
+  Z = (np.random.randn( N ) > 0).astype(int)
+  
+  # use Z as selectors
+  X = Z*( theta1 + stdx*np.random.randn(N) ) + (1-Z)*(theta1+theta2 + stdx*np.random.randn(N))
+  
+  #prior = stats.mvn( np.zeros(2), )
+  def U(q):
+    # initialize model at q's locations
+    p1 = stats.norm( q[0], stdx )
+    p2 = stats.norm( q[0]+ q[1] , stdx )
+    
+    
+    
+    # full dataset:
+    if True:
+      L  = np.zeros( (N,2) )
+      L[:,0] = np.log(0.5) + p1.logpdf( X )
+      L[:,1] = np.log(0.5) + p2.logpdf( X )
+    
+      prior1 = stats.norm( 0, std1 )
+      prior2 = stats.norm( 0, std2 )
+      return -(np.sum( logsumexp( L, 1 ) )  + prior1.logpdf( q[0] ) + prior2.logpdf( q[1] ))
+    else:
+      I = np.random.permutation(N)[:batchsize]
+      data = X[I]
+      scale = float(N) / float(batchsize)
+      L  = np.zeros( (batchsize,2) )
+      L[:,0] = np.log(0.5) + p1.logpdf( data )
[…]

### 2015-02-11 code/sgnht_problems.py

@@ -1,0 +1,245 @@
+import numpy as np
+import pylab as pp
+from scipy import stats
+import pdb
+
+class generate_1d_gaussian_var_unknown( object ):
+  def __init__( self, N = 100, Ntilde = 10, mu_star = 0.0, gamma_star = 1.0,  Nx = 100  ):
+  
+    self.mu0 = 0
+    self.N = N
+    self.Ntilde = Ntilde
+    self.mu_star = mu_star
+    self.gamma_star = gamma_star
+    self.varx = 1.0 / gamma_star
+    self.stdx = np.sqrt(self.varx)
+    self.X = self.mu_star + self.stdx*np.random.randn(self.N)
+    self.muX = self.X.mean()
+  
+    self.gamma_a = 1.0
+    self.gamma_b = 1.0
+  
+    self.gamma_prior = stats.gamma( self.gamma_a, 0.0, 1.0/self.gamma_b )
+    self.mu_prior    = stats.norm( self.mu0, self.stdx )
+  
+    self.SS     = np.sum( (self.X - self.muX)**2 )
+    self.muN    = self.muX*float(self.N)/float(self.N+1.0)
+    self.kappaN = float(self.N+1.0)
+    
+    self.aN = self.gamma_a + float(self.N)/2.0
+    self.bN = self.gamma_b + 0.5*self.SS + (float(self.N)/float(2*(1+self.N)))*( self.muX - self.mu0 )**2
+    
+    self.post_var = 1.0/self.N
+    self.post_std = np.sqrt(self.post_var)
+  
+  
+    self.SS0 = 2*self.gamma_b
+    self.nu0 = 2*self.gamma_a
+    self.nuN = self.nu0 + self.N
+    
+    self.mu_post    = stats.nct( self.nuN, self.muN, np.sqrt(((1.0/(1.0+self.N))*(1.0/self.nuN)*(self.SS0 + self.SS + self.kappaN*( self.muX - self.mu0 )**2)))  )
+    
+    self.gamma_post = stats.gamma( self.aN, 0.0, 1.0/self.bN )
+    
+    self.gamma_post_samples = self.gamma_post.rvs(100000)
+    self.mu_post_samples    = self.muN + np.sqrt(1.0 / (self.kappaN*self.gamma_post_samples) )*np.random.randn(100000)
+    self.Nx = Nx
+    
+    self.bins_mu = np.linspace( -0.6,1.0, Nx )
+    self.bins_gamma = np.linspace( 0,2,Nx)
[…]

### 2015-02-11 code/demo_blowfly_optimization.py
blowfly opt

@@ -1,0 +1,229 @@
+from abcpy.factories import *
+from abcpy.problems.blowfly.blowfly import *
+from sa_algorithms import *
+from scipy import stats as spstats
+import pylab as pp
+
+problem_params = default_params()
+problem_params["std_log_tau"] = 0.2
+problem_params["tau_is_log_normal"] = True
+problem_params["epsilon"] = 0.01 #0.01 #10*np.array([1.0,1.0,0.5,0.5,0.5,0.1,0.01,0.5,2.0,1])
+#problem_params["epsilon"] = 0.1*np.array([0.1,0.1,0.1,0.1,0.1,0.01,0.01,0.1,1.0,1.0])
+#problem_params["epsilon"] = 0.01*np.array([  0.91064006,   0.12437852,   1.06735857,   1.70135243,1.10402222,   0.22966667, 0.08973333,   1.28127273, 10.        ,   9.        ])
+
+problem_params["blowfly_filename"] = "./data/blowfly.txt"
+#problem_params["blowfly_filename"] = "./data/bf2.txt"
+problem = BlowflyProblem( problem_params, force_init = True )
+
+state_params = state_params_factory.scrape_params_from_problem( problem, S=1 )
+
+# set is_marginal to true so using only "current" state will force a re-run
+mcmc_params  = mcmc_params_factory.scrape_params_from_problem( problem, type="mh", is_marginal = True, nbr_samples = 1 )
+
+# ignore algo, just doing this to get state object
+algo_params = { "modeling_approach"  : "kernel",
+                "observation_groups" : problem.get_obs_groups(),
+                "state_params"       : state_params,
+                "mcmc_params"        : mcmc_params,
+                "algorithm"          : "model_mcmc"
+              }
+algo, model, state  = algo_factory.create_algo_and_state( algo_params )
+recorder     = recorder_factory.create_recorder( {} )
+
+class LikelihoodFree( object ):
+  def __init__( self, problem, model, state, recorder ):
+    self.problem = problem
+    self.model = model
+    self.state = state
+    self.recorder = recorder
+    self.tau   = 15
+    
+    self.moving_average_statistics = None
+    self.count = 0
+  
+  def add_prediction( self, x_statistics ):
+    if self.moving_average_statistics is None:
[…]

### 2015-02-24 papers/uai-2015/draft.tex
new draft, lots code

--- 2015-01-20 papers/uai-2015/draft.tex
@@
 \usepackage{amsmath}
+\usepackage[numbers]{natbib}
+%\usepackage{array}
 %\usepackage{nips_style} 
@@
 
-\title{Likelihood-free Inference and Optimization \\ using Stochastic Gradient Approximations}
+\title{Hamiltonian ABC}
 
@@
 \begin{abstract} 
-%\input{abstract}
+  Approximate Bayesian computation (ABC) is a powerful and elegant framework for performing inference in simulation-based models.
+However, due to the difficulty in scaling likelihood estimates, ABC remains useful for relatively low-dimensional problems.
+[REASONS] We introduce Hamiltonian ABC (HABC), a set of likelihood-free algorithms that apply recent advances in scaling Bayesian learning using stochastic gradients with Hamiltonian dynamics to simulation models.
+[WHY GOOD: low nbr sims and non-random walk]  We find that a small number forward simulations can effectively approximate the true ABC gradient, allowing Hamiltonian dynamics to efficiently traverse parameter spaces.
+
+  Models in simulation-based science domains are considered likelihood-free (LF) since they are not defined by a probabilistic model, but by a simulator that maps parameters to pseudo-observations.
+To perform parameter inference in the models, a powerful set of statistical procedures called Approximate Bayesian Computation (ABC) can be applied.
+Although there is a wide variety of ABC algorithms, they are dominated by SMC/PMC and MCMC.
+These have been successful, but their inefficiency of producing samples from the approximate posterior has limited ABC to simulations with relatively few parameters.
+Hamiltonian Monte Carlo (HMC)  provides a framework for extending ABC to high-dimensions, but requires computing the gradient of the log-likelihood.
+With Hamiltonian ABC, we show how recent advances of stochastic gradient HMC algorithms applied to the big data domain can be also used for efficient sampling for LF models.
+We find that a small number forward simulations can effectively approximate the true ABC gradient, allowing Hamiltonian dynamics to efficiently traverse parameter spaces.
+We demonstrate HABC on several typical ABC problems, and show that HABC performs comparably to regular Bayesian inference on a high-dimensional problem from machine learning.
+%
+%
+%   Bayesian inference in simulation-based science relies
+%   Approximate Bayesian Computation (ABC) is a set of inference procedures
+%   for so-called {\em likelihood-free} models
+%
+% In this paper we apply recent techniques from Bayesian inference that use gradient information to sample efficiently from the true posterior distribution to the {\em likelihood-free} or {\em approximate Bayesian computation} (ABC) setting.
+To do this for ABC we adopt a {\em gradient-free} stochastic approximation algorithm by Spall~\cite{spall1999}.
+Together these algorithms provide both optimization and inference for likelihood-free models as the algorithm ABC-SGLD transitions from optimization to sampling.
+We demonstrate ABC-SGLD on problems where the true gradient information is known and on challenging ABC simulators.
 \end{abstract}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+In simulation-based science, models are defined by simulator and parameters.
+These are likelihood-free models because, in contrast to probabilistic models, their likelihoods are either intractable or must be approximated by simulations.
+
+Recent advances in Bayesian inference within the {\em big data} domain have combined the efficiency of stochastic gradient algorithms used for optimization with Hamiltonian dynamics;  these include Stochastic Gradient Langevin Dynamics (SGLD) \cite{welling2011bayesian}, Stochastic Gradient HMC \cite{chen2014stochastic}, and Stochastic Gradient Thermostats \cite{ding2014bayesian}.
+By using small batches of the full data set these methods are able to quickly compute a stochastic approximation to the true gradient, which with a small enough step-size and with an appropriate amount of injected noise, can produce samples from the true posterior.
+The Hamiltonian dynamics avoid random walk behavior by providing momentum to the sampler allowing it to reach far away regions of parameter space.
+Critically for the big data domain, the small step-sizes prevent the trajectories of the sampler from accumulating too much error and therefore they can avoid a Metropolis-Hastings correction step involving the full data set.
+
+Approximate Bayesian Computation (ABC) is a set of inference procedures for so-called {\em likelihood-free} models \cite{abccitations}.
+In contrast to typical Bayesian modeling, where the parameters of a probabilistic model of data are the focus, the primary interest in ABC are the parameters of a simulation model.
+The simulator can generate pseudo or auxiliary data by running the simulator with a set of parameter values, but it cannot compute a likelihood directly.
+Inference involves running simulations, comparing pseudo data with observations, and using a distance between the two as an approximate likelihood.
+Markov chain Monte Carlo (MCMC) or Sequential Monte Carlo (SMC) are the most common inference algorithms.
+It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior.
+
+The approximation to the true posterior can be made arbitrarily small, in principle, by controlling the kernel width parameter $\epsilon$.
+As $\epsilon \rightarrow 0$, however, simulated data must be closer and closer to the observations (or receive zero likelihood).
+The acceptance rate of a Markov chain for small $\epsilon$ can therefore be low.
+This inefficiency of ABC sampling is exacerbated by the random walk proposals.
+As the dimensionality of the parameter space grows,  $\epsilon$ must grow to prevent low acceptance rates.
+This has perhaps prevented application of ABC to high-dimensional simulators.
+Our paper addresses this directly: can we use the sampling efficiency of Hamiltonian dynamics (i.e.
+high acceptance rates) in high-dimensional simulators?
+
+In this paper we adopt the developments in scaling Bayesian inference to ABC.
+To do this we use forward simulations to approximate the likelihood-free gradient.
+The key difference with SG methods mentioned above is that the stochasticity of the gradient does not come from approximating the full data gradient with a batch gradient, but by the stochasticity of the simulator.
+It is therefore not the expense of the simulator (though this could very well be the case for many interesting simulation-based models, and we address this in Future Work) that requires an approximation to the gradient, but the likelihood-free nature of the problem.
+For high-dimensional parameter spaces, computing the gradients naively (i.e.
+by finite differences) can squash the gains brought by the Hamiltonian dynamics; to address this further approximations to the gradient can be used \cite{spall}.
+
+The approximate gradients used by HABC can be noisy and difficult to manage is care is not taken.
+To address this we show how a local Gaussian approximation of the simulator output can improve the stability of the gradient estimates over a direct estimate using the ABC kernel function.
+
+A further innovation of this paper is the use of common random numbers (CRN) to improve the efficiency of the Hamiltonian dynamics.
+By using the same random seed for the finite difference approximation, the first order noise in the gradient is reduced;  by keeping the random seeds fixed for several consecutive steps, the second order stochasticity of one gradient estimate to the next is eliminated.
+We show that this is a valid MCMC procedure.
+
+In Section~\ref{sec:sghd-data} we review three approaches to stochastic gradient inference using Hamiltonian dynamics: SGLD, SGHMC, and SGNHT.
+We briefly review ABC-MCMC in Section~\ref{sec:abcmcmc}, followed by Hamiltonian ABC in Section~\ref{sec:habc}, where  we will show how to improve the stability of the gradient estimates by using CRNs and local density estimators of the simulator.
+Extensions to high-dimensional parameter spaces are also discussed.
+In Section~\ref{sec:toy} we show how HABC behaves on a simple one-dimensional problem, then in Section~\ref{sec:experiments} we compare HABC with ABC-MCMC for two problems: a low-dimensional model of chaotic population dynamics and a high-dimensional problem.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{GRADIENTS FROM FORWARD SIMULATIONS} \label{gradsforward}
+\section{APPROXIMATE BAYESIAN COMPUTATION}\label{sec:abcmcmc}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Consider the Bayesian inference task of either drawing samples from or learning an approximate model of the following (usually intractable) posterior distribution:
+\begin{equation}
+  \pi(\thetav | \y_1, \ldots, \y_N ) \propto \pi(\thetav) \pi( \y_1, \ldots, \y_N  | \thetav )
+\end{equation}
+where $\pi(\thetav)$ is a prior distribution over parameters $\thetav \in {\rm I\!R}^{D}$ and $\pi( \y_1, \ldots, \y_N  | \thetav )$ is the likelihood of $N$ data observations, where $\y_n \in {\rm I\!R}^J$.
+In ABC, the vector of $J$ observations are typically informative statistics of the raw observations.
+  The simulator is treated as generator of random pseudo-observations, i.e.
+$\x \simsim \pi( \x | \thetav )$ is a draw from the simulator.
+Discrepancies between the simulator outputs $\x$ and the observations $\y$ are scaled by a closeness parameter $\eps$ and treated as likelihoods.
+This is the equivalent to putting an $\eps$-kernel around the observations, and using a Monte Carlo estimate of the likelihood using $S$ draws of $\x$:
+\begin{equation}
+  \pi_{\epsvec}( \y | \thetav ) =  \int \pi_{\epsvec}(\y | \x ) \pi( \x | \thetav ) d\x 
+                           \approx  \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}(\y | \x^{(s)} ) \label{eq:abc_mc_approx}
+\end{equation}
+
+In ABC Markov chain Monte Carlo (MCMC) \cite{marjoram2003markov,Wilkinson2013,sisson:2010} the Metropolis-Hastings (MH) proposal distribution is composed of the product of the proposal for the parameters $\thetav$ and the proposal for the simulator outputs:
+\begin{equation}
+  q( \thetavp, \x^{(1)'}, \ldots, \x^{(S)'} | \thetav ) =  q( \thetavp | \thetav ) \prod_s \pi( \x^{(s)'} | \thetavp)
+\end{equation}
+Using this form of the proposal distribution, and using the Monte Carlo approximation eq~\ref{eq:abc_mc_approx}, we arrive at the following Metropolis-Hastings accept-reject probability,
+%
+\begin{equation}
+\alpha = \min \lp 1, \frac{\pi\lp\thetavp\rp \sum_{s=1}^S \pi_{\epsvec}(\y | \xps )  q( \thetav | \thetavp )}{\pi\lp\thetav\rp \sum_{s=1}^S \pi_{\epsvec}(\y | \xs ) q( \thetavp | \thetav )} \rp \label{eq:abc_mh_acceptance_with_s}
+\end{equation}
+%
+If the simulations are part of the Markov chain, the algorithm corresponds to the pseudo-marginal (PM) sampler \cite{delmoral2008,andrieu2009pseudo}, otherwise it is a marginal sampler \cite{marjoram2003markov,sisson:2010}.
+For this paper we will be interested in the PM sampler because this is equivalent to having the random states that generated the simulation outputs in the state of the Markov chain, which we will use within a valid ABC sampling algorithm in Section~\ref{sec:habc}.
+%
+%   If they are discarded, the
+% Two subtly different versions of ABC-MCMC are the pseudo-marginal (PM) \cite{delmoral2008,andrieu2009pseudo} and marginal sampler \cite{marjoram2003markov,sisson:2010}.
+In PM, the simulations are part of the state of the Markov chain whereas in the marginal sampler they are not
+% When only the numerator is re-estimated at every iteration (and the denominator is carried over from the previous iteration), then this algorithm corresponds to pseudo-marginal (PM) sampling \cite{delmoral2008,andrieu2009pseudo}.
+PM sampling is asymptotically correct (taking for granted the approximation introduced by the kernel $\pi_{\epsvec}$) but can display very poor mixing properties.
+By resampling the denominator as well, we improve mixing at the cost of introducing a further approximation.
+This sampler is known as the marginal sampler \cite{marjoram2003markov,sisson:2010}.
+% [TODO: END TAKEN FROM POPE]
+
+An alternative approach to computing the ABC likelihood is to estimate the parameters of a conditional model  $\pi(\x|\thetav)$, e.g.
+kernel density estimate \cite{TurnerGenLik2014} or a Gaussian model \cite{wood2010statistical}.
+While either approach should be adequate and both have their own limits and advantages, for this paper we will use a Gaussian model.
+In ABC, using a conditional Gaussian model for  $\pi(\x|\thetav)$ is called a {\em synthetic likelihood} (SL) model \cite{wood2010statistical}.
+For a SL log-likelihood model, we compute estimators of the first and second moments of $\pi(\x|\thetav)$.
+The advantage is that for a Gaussian $\epsilon$-kernel, we can convolve the two densities
+\begin{eqnarray}
+  \pi_{\epsvec}( \y | \thetav ) & = & \int \mathcal{N}( \y | \x, \eps^2 ) \mathcal{N}( \x | 
+  \mu_{\thetav}, \sigma^2_{\thetav} ) d\x \\
+                          & = & \mathcal{N}( \y | \mu_{\thetav}, \sigma^2_{\thetav} + \eps^2 )
+\end{eqnarray}
+
+Of particular concern to this paper is the behavior of the log-likelihoods for different values of $\epsilon$.
+In the $\eps$-kernel case, the log-likelihood is very sensitive to small values of $\eps$:
+\begin{eqnarray}
+  \log \pi_{\epsvec}( \y | \thetav ) & = & \log \sum_s \mathcal{N}( \y | \x^{(s)}, \eps^2 ) \\
+                          & = & \log \mathcal{N}( \y | \x^{(s)}, \eps^2 ) + \log\lp 1 + H \rp \\
+                          & \approx & -\log \eps -\frac{1}{2 \eps^2} ( \y - \x^{(m)} )^2 
+\end{eqnarray}
+where $m$ is the simulation that is closest to $\y$, $H$ is a sum over terms close to $0$.
+We can see that the log-likelihood can be set arbitrarily small by decreasing $\eps$.
+On the other hand, by using a model of the simulation at $\thetav$
+\begin{eqnarray}
+ \log \pi_{\epsvec}( \y | \thetav ) & \approx & -\frac{1}{2} \log (\sigma^2_{\thetav} + \eps^2 ) -\frac{( \y - \mu_{\thetav} )^2 }{2 (\sigma^2_{\thetav} + \eps^2 )} 
+\end{eqnarray}
+ For the SL model, $\eps$ acts as a smoothing term and can be set to small values with little change to the log-likelihood, as long as the SL estimators are fit appropriately.
+This insensitivity to $\eps$ will be used in Section~\ref{sec:habc} for estimating gradients of the ABC likelihood.
+Before describing HABC in full detail however, we now explain how scaling Hamiltonian dynamics in Bayesian learning can be accomplished using stochastic gradients from batched data.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{SCALING BAYESIAN INFERENCE USING HAMILTONIAN DYNAMICS} \label{sec:sghd-data}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Scaling Bayesian inference algorithms to massive datasets is necessary for their existing relevance in the so-called {\em big data} era.
+We now review the role stochastic gradient methods combined with Hamiltonian dynamics have played in recent advances in scaling Bayesian inference.
+Most importantly, these methods have combined the ability of HMC to explore high-dimensional parameter spaces with the computational efficiency of using stochastic gradients based on small batches of the full dataset.
+After an overview of HMC, we will briefly describe stochastic gradient Hamiltonian dynamics (SGHDs), starting with using  Langevin dynamics \cite{welling2011bayesian}, then HMC with friction \cite{chen2014stochastic}, and finally HMC with thermostats \cite{ding2014bayesian}.
+We will then make the connection between SGHDs and HABC in Section~\ref{sec:habc}.
+
+\subsection{Hamiltonian Monte Carlo}
+
+Hamiltonian dynamics are often necessary to adequately explore the target distribution of high-dimensional parameter spaces.
+By proposing parameters that are far from the current location and yet have high acceptance probability, Hamiltonian Monte Carlo \cite{duane1987hybrid, neal2011mcmc}  can efficiently avoid random walk behavior that can render proposals in high-dimensions inadequate.
+
+HMC simulates the trajectory of a particle along a frictionless surface, using random initial momentum $\rhov$ and position $\thetav$.
+The Hamiltonian function computes the energy of the system and the dynamics govern how the momentum and position change over time.
+The continuous Hamiltonian dynamics can be simulated by discretizing time into small steps $\eta$.
+If $\eta$ is small, the value of $\thetav$ at the end of a simulation can be used as proposals within the Metropolis-Hastings algorithm.
+Hamiltonian dynamics should propose $\thetav$ that are always accepted, but errors due to discretization may require a  Metropolis-Hastings correction.
+It is this correction step that SGHD algorithms want to avoid as it requires computing the log-likelihood over the full data set.
+
+More formally, the Hamiltonian $H\lp\thetav, \rhov \rp = U(\thetav) + K(\rhov)$ is a function of the current potential energy $U(\thetav)$ and kinetic energy $K(\rhov) = \rhov^T M^{-1}\rhov/2$ ($M$ is a diagonal matrix of masses which for presentation are set to $1$).
+The potential energy is defined by negative log joint density of the data and prior:
+\begin{equation}
+  U(\thetav) = - \log \pi(\thetav) - \sum_{i=1}^N \log \pi(\yi | \thetav )
+\end{equation}
+The Hamiltonian dynamics follow 
+\begin{equation}
+  d\thetav = \rhov dt ~~~~~~~~~~~~ d\rhov = -\nabla U(\thetav) dt
+\end{equation}
+in simulation $dt = \eta$. 
+% Actual simulation of the Hamiltonian requires taking discrete steps, and therefore introduces some error.
+The error can be corrected using a MH step, as shown in Algorithm 1 (showing leap-frog algorithm).
+By only taking a single leap-frog step, HMC reduces to Langevin dynamics which can be represented as a gradient descent algorithm with injected noise.
+By using small step-sizes the acceptance rate approaches 1, so that no MH correction steps are necessary and the algorithm can be run to generate a stream of (possibly highly correlated) samples from the posterior.
+
+% Two problems must be addressed.
+The first is loglikelihood computation.
+For correctness, Bayesian inference algorithms require frequent computation of the (deterministic) loglikelihood over the entire observation set.
+Contrast this to optimization, where stochastic gradient techniques are necessary for learning parameters.
+The second problem is that massive parameter sets is synonymous with massive datasets.
+As datasets grow, so too do our models.
+This is particularly true in the area of deep learning, where parameter vectors start in the hundred thousands and for some models are in the billions.
+Hamiltonian Monte Carlo addresses this for Bayesian inference: by using gradients of the log-likelihood (log-prior) it is possible to avoid random walk behavior and more efficiently explore the posterior.
+
+% Recent advances in scaling Bayesian inference to big data has shown that both batches and gradients can be used to correctly sample from the posterior distribution.
+These use Hamiltonian dynamics based on stochastic gradients (where the stochasticity comes from approximating the true gradient with one based on a batch of data) with appropriate step-sizes (and variations, see below).
+We now describe these algorithms and then show how we can adopt them for ABC.
+
+\subsection{Stochastic Gradient Hamiltonian Dynamics}
+If the log-likelihood over the full data set is replaced with a batch estimate, as is done for the following {\em stochastic gradient Hamiltonian dynamics} (SGHDs) algorithms, then the error in simulating the Hamiltonian dynamics comes not only from the discretization, but from the variance of the stochastic gradient.
+As long as this error is controlled, either by using small steps $\eta$ (SGLD), or adding friction terms $A$ (SGHMC), or using a thermostat $\xi$ (SGNHT), the expensive MH correction step can be avoided and values of $\thetav$ from the Hamiltonian dynamics can be used as samples from the posterior.
+
+SGHDs replace the full potential energy and its gradient with a batch approximation:
+\begin{eqnarray}
+  \hat{U}(\thetav)       & = & - \log \pi(\thetav) - \frac{N}{n} \sum_{i=h_1}^{h_n} \log \pi(\yi | \thetav ) \\
+  \nabla\hat{U}(\thetav) & = & - \nabla \log \pi(\thetav) - \frac{N}{n} \sum_{i=h_1}^{h_n} \nabla \log \pi(\yi | \thetav ) 
+\end{eqnarray} 
+where $n$ is the batch size, and $h_i$ are indices chosen randomly without replacement from $[1,N]$ (i.e. it defined a random batch).  
+%The main motivation for these is to avoid the expensive MH correction step while maximizing the benefit of Hamiltonian dynamics.
+For example, we want to take large steps but without introducing errors that cause the dynamics to fail.
+
+{\bf Stochastic gradient Langevin dynamics} \cite{welling2011bayesian} performs one full leap-frog step of HMC.   Starting with a half step for the momentum, the update for $\thetav$ is 
+\begin{eqnarray}
+  \rhov_t & \sim & \mathcal{N}(0,\eye_p) \\
+  \rhov_{t+\frac{1}{2}} & = & \rhov_t - \eta \nabla \hat{U}(\thetav_t) /2 \\
+  \thetav_{t+1} & = & \thetav_t + \eta \rhov_{t+\frac{1}{2}}
+\end{eqnarray}
+Tt is not necessary to include $\rhov$ in the updates since there is only one step:
+\begin{eqnarray}
+  \thetav_{t+1} & = & \thetav_t + \eta \mathcal{N}(0,\eye_p) - \eta^2 \nabla \hat{U}(\thetav_t) /2 
+\end{eqnarray}
+One of the potential drawbacks of SGLD is that the momentum term is {\em refreshed} for every update of the parameters, and since this means the parameter update only uses the current gradient approximation, it limits the benefits of using Hamiltonian dynamics.
+On the other hand, this also prevents SGLD from accumulating errors in the Hamiltonian dynamics.
+
+{\bf Stochastic Gradient HMC} (SGHMC) \cite{chen2014stochastic} avoids $\rhov$ refreshment altogether.
+By applying HMC directly using the stochastic approximation $\hat{U}$ and $\nabla \hat{U}$, which the authors call {\em naive SGHMC}, the variance of the gradient will introduce errors that left unaddressed will result in sampling from the incorrect target distribution.
+Under the assumption that $\nabla \hat{U}(\thetav) = \nabla U(\thetav) + \mathcal{N}\lp \zerov, \varv_{\thetav}\rp$, where $\varv_{\thetav}$ is the covariance of the gradient approximation, and  updates $\rhov_{t+1} = \rhov_{t} + \Delta \rhov_{t}$ and $\thetav_{t+1} = \thetav_t + \eta \rhov_{t+1}$, the change in momenta $\Delta \rhov$ from one full step is
+ \begin{equation}
+   - \eta \lp \nabla U(\thetav) + \mathcal{N}\lp \zerov, \varv_{\thetav}\rp\rp  =  - \eta \nabla U(\thetav) + \mathcal{N}\lp \zerov, \eta^2 \varv_{\thetav}\rp  
+ \end{equation}
+By adding a friction term $\Bv$ to $\Delta \rhov$ proportional to $\varv_{\thetav}$, the correction step can be avoided
+ \begin{eqnarray}
+   \Delta \rhov & = & - \eta \Bv \rhov_t - \eta \nabla U(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta \Bv\rp 
+ \end{eqnarray}
+ where  $\Bv=\frac{1}{2}\eta \varv_{\thetav_t}$.
+In practice, since we can only estimate $\Bv$ by some $\hat{\Bv}$ and can only compute $\hat{U}$, a user defined friction term $\Cv$ is used (with $\Cv - \hat{\Bv}$ is semi-positive definite).
+Thus the updates used for $\Delta \rhov$ for SGHMC:
+ \begin{equation}
+   - \eta \Cv \rhov_t - \eta \nabla \hat{U}(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta (\Cv-\hat{\Bv}) \rp 
+ \end{equation}
+ In our experiments we compute an online estimate $\hat{\varv}$ and set $\Cv = c \eye_p + \hat{\varv}$.  
+ %As mentioned above, the main benefit of using SGHMC over SGLD is the ability of avoiding refreshing $\rhov$ while not accumulating Hamiltonian errors.
+
+ {\bf Stochastic Gradient thermostats} (SGT) \cite{ding2014bayesian} addresses the difficulty of estimating $\hat{\Bv}$ by introducing a scalar variable $\xi$ who's addition to the Hamiltonian dynamics maintains the temperature of the system constant, i.e.
+it acts as a (Nose-Hoovier) thermostat \cite{leimkuhler:2009}.
+The update equations remain simple: initialize $\xi = \Cv$ (or $c$), then for $t=1\ldots$
+ \begin{eqnarray}
+   \rhov_{t+1} & = & \rho_t - \eta \xi_t \rhov - \eta \nabla \hat{U}(\thetav_t) + \mathcal{N}\lp \zerov, 2 \eta_t \Cv \rp \\
+   \thetav_{t+1} & = & \thetav_t + \eta \rhov_{t+1} \\
+   \xi_{t+1} & = & \xi_t + \eta \lp  \rhov_{t+1}^T \rhov_{t+1} / p - 1 \rp
+ \end{eqnarray}
+ In summary, the hyperparameters required for these algorithms are $\eta$ and $\Cv$ (for SGHMC and SGNHT only), and in practice, some way of estimating $\hat{\varv}$ for SGHMC.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{HAMILTONIAN ABC}\label{sec:habc}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+In the application of stochastic gradient Hamiltonian dynamics to ABC, the stochasticity of the gradient due to batch sizes is instead caused by two other factors.
+The first factor, is due to the finite number of forward simulations that are used to approximate the log-likelihood function.
+We will explain below how to perform ABC gradient estimates based on finite difference approximations on two log-likelihood formulations: the $\epsilon$-kernel density model and on a Gaussian conditional model, both are based on $S$ random simulations.
+The second factor is due to the dimension of $D$.
+For large $D$, a finite difference estimate requires $2p$ forward simulations for each random simulation, by using a further stochastic approximation to the gradient, called Simultaneous Perturbation Stochastic Approximation (SPSA) \cite{spall}, an unbiased estimate of the gradient can computed with $2q$ forward simulations, where $ \geq 1$ and usually $Q \ll D$.
+
+To begin, we define $\thetavplusd = \thetav + c \Deltavd$ and $\thetavminusd = \thetav + c \Deltavd$, where $c$ is a small positive scalar, e.g.
+$c=0.001$, and $\Deltavd$ is a vector of size $D$ that, for FD estimates is $1$ for entry $d$ and $0$ otherwise.
+Our gradient estimate require simulating $S$ times at $\thetavplus$ and $\thetavminus$, generating sets of pseudo-data $\{ \xplussd \}$ and $\{ \xminussd \}$.
+These are then plugged into the log-likelihood estimate for $\y$, which we call $\mathcal{L}$:
+\begin{eqnarray}
+  \frac{\partial \mathcal{L}(\thetav)}{\partial \thetavd} & \approx & \frac{\mathcal{L}(\thetavplusd)- \mathcal{L}(\thetavminusd)}{2c}
+\end{eqnarray}
+
+For the $\epsilon$-kernel density model
+% ,
+% \begin{eqnarray}
+%   \mathcal{L}(\thetavpmd) & = & \log \sum_{s=1}^2 \mathcal{N}( \y | f( \thetavpmd, \omega_s ), \eps^2 ) \\
+%                           & = & \log \mathcal{N}( \y | f( \thetavpmd, \omega_{m} ), \eps^2 ) + \log\lp 1 + H_d \rp \\
+%                           & \approx & \log \mathcal{N}( \y | f( \thetavpmd, \omega_{m} ), \eps^2 ) + 0
+% \end{eqnarray}
+% where we replaced the random variable $\x$ with a deterministic function $f( \thetavpmd, \omega_s )$ (i.e.
+the simulator run at $\thetavpmd$ using the random state $\omega_s$, such that $\xpmd = f( \thetavpmd, \omega_s )$).
+In the second line we have pulled out the value from the random state $m$ which has the maximum log-likelihood (i.e.
+is closest to $\y$) and used $H_d$ (sum of terms $<0$ and $\ll 0$ for small $\eps$), and in the last line simplified further by assuming small $\eps$, so that the log-likelihood is dominated by a single simulation closest to $\y$.
+it is easy to show that the gradient estimate is sensitive to $\epsilon$ in this case: 
+\begin{eqnarray}
+  \frac{\partial \mathcal{L}(\thetav)}{\partial \thetavd} & \approx & \frac{1}{2c\eps^2}\lp -\frac{1}{2}(\y-\xplussd)^2 + \frac{1}{2}(\y-\xminussd)^2  \rp
+\end{eqnarray}
+So by making $\epsilon$ arbitrarily small we can make the gradient extremely sensitive to our simulations.
+Intuitively, we want the gradient to be fairly stable no matter what values of $\epsilon$ we use.
+To do this we will replace log-likelihood with a conditional model.
+%
+% A better way of approximating the gradient is to use a conditional model of $\pi(\x|\thetav)$, for example a conditional kernel density estimate \cite{TurnerGenLik2014} or a conditional Gaussian model \cite{wood2010statistical}.
+While either approach should be adequate and both have their own limits and advantages, for this discussion we will use a Gaussian estimate.
+In ABC, using a conditional Gaussian model for  $\pi(\x|\thetav)$ is called a {\em synthetic likelihood} (SL) model \cite{wood2010statistical}.
+For a SL log-likelihood model, we compute estimators of the first and second moments of $\pi(\x|\thetav)$.
+For Gaussian $\epsilon$-kernel, we can convolve the two densities
+% \begin{eqnarray}
+%   \mathcal{L}(\thetavpmd) & = & \log \int \mathcal{N}( \y | \x, \eps^2 ) \mathcal{N}( \x |
+%   \mu_{\thetavpm}, \sigma^2_{\thetavpm} ) d\x \\
+%                           & = & \log \mathcal{N}( \y | \mu_{\thetavpm}, \sigma^2_{\thetavpm} \eps^2 )
+% \end{eqnarray}
+We can see that $\epsilon$ now can be set very small as it now acts as a smoother for our estimate, resulting in a robust gradient.
+
+\subsection{Gradients in High Dimensions}
+
+%\subsection{Spall's method: SPSA}
+In the gradient-free setting, Spall \cite{spall_spsa} provides a stochastic approximate to the true gradient using only 2 forward simulations (function evaluations).
+This is in contrast to multivariate finite-difference stochastic approximation FDSA \cite{kiefer1952} requiring $2p$ evaluations.
+
+The gradient estimate is
+\begin{equation}
+  \gradestimate{t}{\thetav_t} = \begin{bmatrix} 
+                                   \frac{\feval{t}{+} - \feval{t}{-}}{2 \cstep{t}{1} \dstep{t}{1}} \\
+                                   \vdots \\
+                                   \frac{\feval{t}{+} - \feval{t}{-}}{2 \cstep{t}{p} \dstep{t}{p}} \\
+                                \end{bmatrix}
+\end{equation}
+where $\cstep{t}{p}$ is a step-size that is usually constant for all dimensions $p$, but can be different (as shown in this case); $\dstep{t}{p} \in {-1, +1}$ is a {\em perturbation mask} (called symmetric Bernouilli variables by Spall), i.e.
+$\dstep{t}{p} \sim 2*\text{Bernouilli}(0.5) - 1$; and  $\feval{t}{\pm}$ are function evaluations:
+\begin{eqnarray}
+  \feval{t}{+} &= \loglik{\thetav + \cstepconst{t}\dstepconst{t}} \\
+  \feval{t}{-} &= \loglik{\thetav - \cstepconst{t}\dstepconst{t}}
+\end{eqnarray}
+A way of reducing the noise in the gradient is by averaging over $q$ draws of $\dstepconst{t}$:
+\begin{equation}
+  \gradestimate{t}{\thetav_t} = \frac{1}{q} \sum_{j=1}^q \gradestimaterepeat{t}{\thetav_t}{j}
+\end{equation}
+where for each $j$ new perturbation masks are drawn.
+
+For ABC, the gradients we compute are stochastic because we must approximate $\pi(\x | \thetav)$ with forward simulations.
+We can make the approximation arbitrarily good by running more simulations, but this is often unnecessary for sampling, because we can make use of the uncertainty in our current approximation to decide if our sample is correct or not (cite UAI2014?).
+
+\begin{itemize}
+  \item Log likelihood is much noisier in  1, and the 2 there are good reasons why this is a better choice: is no longer depends on $epsilon$ (as the true gradient shouldn't), convolving the conditional density is the right thing to do, if possible (though it might not be necessary to perform ABC)
+  \item In both cases, we rewrite the expression conditioning on a set of random seeds omega.
+Explain the reason for this.
+By observing the simulation at different theta but same seeds, we can eliminate much of the noise in the markov chain.
+By treating omega as part of the state, we can smoothly move along noiseless outputs for several iterations and then step.
+\end{itemize}
+
+\subsection{Stochastic ABC Gradients}
+For low-dimensional problems a finite difference (FD) approximation to $\nabla U$.
+For each dimension $p$, forward simulate for each random seed twice at plus $c$ and minus $c$ (only along dimension $p$): $x_{\pm}^{(s)} \sim \pi( \x | \thetav \pm c \Delta_p, \omega_s)$.
+These simulations are then plugged into the appropriate estimator (logsumsexp or SL).
+$\hat{U}_{K_{\eps}}$:
+\begin{eqnarray}
+  - \log \pi(\thetav \pm c \Delta_p) - \log \sum_s \pi_{\eps}(\y | f( \thetav \pm c \Delta_p, \omega_s ) )
+\end{eqnarray}   
+or $\hat{U}_{SL}$
+\begin{eqnarray}
+  - \log \pi(\thetav \pm c \Delta_p) - \log \mathcal{N}(\y | \mu_{\thetav \pm c \Delta_p}, \sigma^2_{\thetav \pm c \Delta_p} + \eps^2 )
+\end{eqnarray}
+
+\begin{eqnarray}
+  \hat{U}(\thetav)       & = & - \log \pi(\thetav) - \log \pi_{\eps}(\y | \thetav ) \\
+  \nabla\hat{U}(\thetav) & = & - \nabla \log \pi(\thetav) - \frac{N}{n} \sum_{i=h_1}^{h_n} \nabla \log \pi(\yi | \thetav ) 
+\end{eqnarray}
+
+\subsection{Common Random Numbers}
+
+[CITE NEAL How to view an MCMC...]
+Another variance reduction technique called the method of {\em common random numbers} (CRN) \cite{kleinman1999} sets a common seed for the random number generator (RNG) for both calls to the simulator.
+This technique can remove the effect of the simulator noise in the gradient estimate, leaving on the randomness of the perturbation masks as the source of noise.
+Consider using SPSA instead of SGD: the CRN technique is equivalent of using the same batch of data vectors to evaluate the log-likelihood which is a sensible approach.
+
+The ABC likelihood is proportional to the convolution between a kernel around the observation statistics $\y$ and the generator of pseudo-statistics $\x$, i.e.
+the simulator.
+Since pseudo-statistics can only be generated, we resort to a Monte Carlo estimate for the ABC likelihood based on $S$ draws from the simulator using the same parameter vector $\thetav$ but different random seeds:
+\begin{eqnarray}
+  \pi\lp \y | \thetav \rp & = &  \int \pi_{\epsvec}\lp \y | \x \rp \pi\lp \x | \thetav \rp d \x \\
+   & \approx & \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}\lp \y | \x^{(s)} \rp
+\end{eqnarray}
+where $\x^{(s)} = \pi\lp \x | \thetav, \omega_s \rp$.
+Note we have passed a unique random seed $\omega_s$ into the simulator, making $\x^{(s)}$ a deterministic function of the simulator with parameters $\{\thetav, \omega_s\}$.
+It will be useful to rewrite this deterministic simulator function as $f$: $\x^{(s)} = f\lp \thetav, \omega_s \rp$, allowing us to rewrite the likelihood as
+\begin{eqnarray}
+  \pi\lp \y | \thetav \rp & \approx & \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}\lp \y | f\lp \thetav, \omega_s \rp \rp
+\end{eqnarray}
+
+% \subsection{Conditional Density Estimators}
+% [TODO: TAKEN FROM POPE]  When the simulator is fast and stochastic, it can be beneficial to the inference procedure to build a local, conditional model of the distribution $\pi(\y|\thetav)$ using $S$ simulator responses in $\yone, \ldots, \yS \overset{\simulator}{\sim} \pi( \y | \thetav)$.
+% %
+% The simplest local response model is the \emph{conditional Gaussian}, an approach called {\em synthetic likelihood} in ABC \cite{wood2010statistical}.
+It computes estimators of the first and second moments of the responses and uses the Gaussian distribution to analytically compute the likelihood (thus providing an alternative to kernel ABC).
+For our algorithms, this allows the direct computation of the CDF:
+% \begin{eqnarray}
+%  && \muhattheta  =  \frac{1}{S} \sum_{s=1}^{S} \y_s ~~~~~~~~~~~
+%  \hat{\Sigma}_{\thetav}  =  \frac{1}{S-1} \sum_{s=1}^{S} \lp \ys - \muhattheta \rp \lp \ys - \muhattheta \rp^T\\
+%  &&\F_{\y | \thetav}(\ystar; \muhattheta,  \hat{\Sigma}_{\thetav})  =  \int_{-\infty}^{\ystar} \mathcal{N}\lp \y | \muhattheta,  \hat{\Sigma}_{\thetav} \rp d \y
+% \end{eqnarray}
+% % \begin{eqnarray}
+% %   &&\F_{\y | \thetav}(\ystar; \muhattheta,  \hat{\Sigma}_{\thetav})  =  \int_{-\infty}^{\ystar} \mathcal{N}\lp \y | \muhattheta,  \hat{\Sigma}_{\thetav} \rp d \y
+% % \end{eqnarray}
+% where $\muhattheta$ and $\hat{\Sigma}_{\thetav}$ are computed from the $S$ simulations.
+In our experiments we  often use a factorized model: $\mathcal{N}(\y | \muhattheta,\hat{\Sigma}_{\thetav}) \approx \prod_{j=1}^J \mathcal{N}( y_j | \muhatthetaj,  \sigmahatthetaj)$, resulting in a factorized product over CDFs as well.
+% Modeling the response by only the first two moments may be inadequate due to multi-modality, asymmetric noise, etc.
+For such cases a \emph{conditional KDE} (kernel density estimate) response model can by used.
+In \cite{TurnerGenLik2014} this approach is shown to be superior to conditional Gaussians for certain computational psychology models.
+Note that for Gaussian kernels the conditional KDE is very similar to kernel ABC, but has additional flexibility of adaptively choosing bandwidths (rather than the fixed $\epsvec$ in kernel ABC).
+% [TODO: END TAKEN FROM POPE]
+
+\subsection{Algorithms}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Demonstration}\label{sec:toy}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+[FROM UAI] In this illustrative problem we infer the rate of an exponential distribution under a gamma prior with shape $\alpha$ and rate $\beta$, having $N$ observations at the true rate $\theta^{\star}$; this is the {\em exponential example} in \cite{turner2012tutorial}.
+Let $\w$ be a vector of $N$ draws from an exponential distribution, i.e.
+$w_n \sim \expdist(\theta)$.
+The posterior is a gamma distribution with shape $\alpha+N$ and rate $\beta + \sum w_n$.
+To use this problem with ABC, we use the exponential draws as the simulator and the mean of $\w$ as the statistic $y$ and assume that $N$ is known.
+The inference problem is therefore to sample from $p( \theta | y, \alpha, \beta, N )$.
+
+\subsection{First order CRNs}
+Using same seed in a gradient estimate at time $t$ (versus not using them).
+This is makes sense if we compare to SG-data where a single batch is used for computing the gradient.
+If a FD method was used, we would use the same batch.
+
+\subsection{Second order CRNs}
+Using same seed in a gradient estimate at multiple time steps (versus not changing them -- the previous example).  Analogous to keeping the same batch for consecutive steps in SG-data.
+
+\subsection{Comparing Gradient Estimates with SL}
+\begin{itemize}
+  \item Can estimate ABC posterior by CLT of Gammas (sum of exponential).
+For large S, the variance of the SL approximates the ``true'' CLT approximation.
+Thus we can set $\epsilon$ much smaller for the SL case.
+The kernel logsumexp suffers when $\epsilon$ is small, using only the closest simulation to $\y$.
+Instead the SL estimates a density around the simulations and the likelihood is less sensitive to $\epsilon$.
+  \item Better tolerance to low $\epsilon$ using SL.
+\end{itemize}
+
+\subsection{Using variance of gradients in SGHMC}
+Compute online $\hat{B}$ then fix.
+
+\subsection{Visualizing Noise from Gradient versus Injected Noise}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Experiments}\label{sec:experiments}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\subsection{Robbin's Monro: SGD}
+\subsection{Blowfly}
+[FROM UAI] Adult blowfly populations exhibit dynamic behavior for which several competing population models exist.
+In this experiment, we use observational data and a simulation model from Wood \cite{wood2010statistical}, based on their improvement upon previous population dynamics theory.
+Population dynamics are modeled using (discretized) differential equations that can produce chaotic behavior for some parameter settings.
+An example blowfly replicate series is shown in Figure~\ref{fig:blowflyexamples}, along with  times-series generated by a sample from $\p(\thetav|\y)$ using GPS-ABC.
 
-\subsection{Spall's method: SASP}
+In \cite{wood2010statistical} there are several explanations of the population dynamics, corresponding to different simulations and parameters.
+We concentrate on the equation (1) in section 1.2.3 of the supplementary information, considered ``a better alternative model" by the author.
+The population dynamics equation generates  $N_1, \ldots, N_T$ using the following update rule:
+\begin{equation}
+N_{t+1} = P N_{t-\tau} \exp(-N_{t-\tau}/N_0) e_t + N_t \exp(-\delta \epsilon_t) \nonumber
+\end{equation}
+where $e_t \sim  \mathcal{G}( 1/{\sigma_p^2},1/{\sigma_p^2})$ and $\epsilon_t 
+ \sim  \mathcal{G}( 1/{\sigma_d^2},1/{\sigma_d^2})$  
+are sources of noise, and $\tau$ is an integer (not to be confused with the $\tau$ used as the MH acceptance threshold in our algorithms).
+In total, there are 6 parameters $\theta = \{ \log P, \log \delta, \log N_0, \log \sigma_d, \log \sigma_p, \tau\}$.
+See \cite{wood2010statistical} for further details about the significance of the parameters.
+We put Gaussian priors over all the parameters (with Gaussian proposal distributions), except for $\tau$ which has a Poisson prior (and a left/right increment proposal).
+Time-series generated with parameters from this prior distribution produce extremely varied results, some are chaotic, some are degenerate, etc.
+Modeling this simulator is {\em very} challenging.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{LANGEVIN DYNAMICS} \label{ld}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+As with any ABC problem the choice of statistics is important as it directly affects the quality of the results.
+It is also non-trivial and requires careful thought and sometimes trial and error.
+In total there are 10 statistics: the log of the mean of all $25\%$ quantiles of $N/1000$ (4 statistics), the mean of the $25\%$ quantiles of the first-order differences of $N/1000$ (4 statistics), and the maximal peaks of smoothed $N$, with 2 different thresholds (2 statistics).
+With these statistics it is possible to reproduce time-series that appear similar to the observations.
+Note that these statistics are different from Wood's, but they capture similar time-series features and are sufficient to produce credible population dynamics.
+[END FROM UAI]
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{STOCHASTIC-GRADIENT LD} \label{sgld}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Bayesian Autoencoders}
+We perform Bayesian inference on an autoencoder neural network using MNIST images as observations.
+We apply two inference algorithms: SGNHT and HABC.
+For SGNHT, we use the known gradient and for HABC we apply 2SPSA.
+After 1000 iterations we collect parameter vectors every 100 iterations for a total of 100 sets.
+(Do we use SGHMC to estimate Bhat in first 1000 iterations?).
+Use same batches at each iteration for SGNHT and HABC?
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{SGLD-ABC} \label{sgld}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Statistical Tests for Gradients}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{EXPERIMENTS} \label{experiments}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Logistic Regression with Stochastic Gradients}
-
-\subsection{Likelihood-free Inference of Blowfly Dynamics}
-
-\subsection{Other Useful ABC problem}
+results: convergence of training error (batches), convergence in test error using parameters collected (record nbr of forward or backward passes required), show mean and variance of filters (how?), compare with using the last parameter sample only (stochastic optimization).
 
@@
 
+% \begin{itemize}
+%   \item
+%   \item
+% \end{itemize}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\begin{itemize}
+  \item New set of ABC algorithms for efficient exploration of posterior distribution.
+  \item Key: opens the door to high-dimensional ABC inference.
+\end{itemize}
+
+A further complication, which we do not address on this paper, is the forward simulation might be, and for many interesting problems is, expensive.
+We would therefore prefer to resort to some sort of surrogate modeling \cite{Meeds2014GpsUai,wilkinson:2014, michael} to avoid simulations whenever possible.
+We address this scenario in Section~\ref{future}.
+Shown for relatively fast simulators.
+Extensions to surrogates with GPS (Meeds2014GpsUai, rasmussen:2003, wilkinson:2014) can use derivative of GP for likelihood as GP of gradient.
+These could be matched by occasional gradient estimates (Osborne).
+
 \subsubsection*{References}
-\newpage
 {
-\bibliographystyle{abcsgld}
+\bibliographystyle{icml2014}
+\bibliography{abcsgld}
 }

### 2015-02-24 code/demo_blowfly.py

@@ -4,12 +4,15 @@ from sa_algorithms import *
 from scipy import stats as spstats
 import pylab as pp
 
+from working_code import *
+
 problem_params = default_params()
 problem_params["std_log_tau"] = 0.2
 problem_params["tau_is_log_normal"] = True
 problem_params["epsilon"] = 0.9 #0.01 #10*np.array([1.0,1.0,0.5,0.5,0.5,0.1,0.01,0.5,2.0,1])
 #problem_params["epsilon"] = np.array([1.0,1.0,0.5,0.5,0.5,0.1,0.01,0.5,2.0,1])
-problem_params["epsilon"] = 2.0+0*np.array([0.25,0.25,0.25,0.25,0.25,0.05,0.01,0.25,2.0,2.0])
+problem_params["epsilon"] = np.array([0.25,0.25,0.25,0.25,0.25,0.05,0.01,0.25,2.0,2.0])
+#problem_params["epsilon"] = 0.5
 #problem_params["epsilon"] = 0.1*np.array([0.1,0.1,0.1,0.1,0.1,0.01,0.01,0.1,1.0,1.0])
 #problem_params["epsilon"] = 0.01*np.array([  0.91064006,   0.12437852,   1.06735857,   1.70135243,1.10402222,   0.22966667, 0.08973333,   1.28127273, 10.        ,   9.        ])
 
@@ -82,32 +85,22 @@ def run_sgld( problem, T, h, A, c, theta=None, x = None, verbose_rate = 100, use
     x = problem.simulate( theta, omega )
     loglike_x = problem.loglike_x( x )
     
+    if accept_move(log_acceptance):
+      x         = x_proposal
+      theta     = theta_proposal
+      loglike_x = loglike_x_proposal
+    
+    X.append(x)
+    THETAS.append(theta)
+    LL.append(loglike_x)
+    
     # --------------- #
     # samples omegas  #
     # --------------- #
-    if use_omega and np.random.rand()<1.1:
-      omega_proposal = np.random.randint(T)
-      x_proposal     = problem.simulate( theta, omega_proposal )
-      loglike_x_proposal = problem.loglike_x( x_proposal )
-      
+    if use_omega and np.random.rand()<0.01:
+      theta, x, omega, loglike_x = omega_sample(problem, theta, x, omega, loglike_x )
     
-      #log_acceptance =  -0.5*( (x_proposal-x)**2 /(problem.x_epsilon**2) )
[…]

### 2015-02-24 code/demo_exponential.py

@@ -5,14 +5,15 @@ from scipy import stats as spstats
 import pylab as pp
 
 problem_params = default_params()
-problem_params["epsilon"] = 1.0
+
 problem_params["N"] = 5
 problem_params["q_stddev"] = 0.5
 problem_params["theta_star"]      = 0.1
-#problem_params["alpha"]           = 1.0
-#problem_params["beta"]            = 1.0
-problem_params["alpha"]           = 3.75
-problem_params["beta"]            = 1.01
+problem_params["epsilon"] = 0.1*np.sqrt( 1.0 / (problem_params["N"]*problem_params["theta_star"]**2) )
+problem_params["alpha"]           = 1.0
+problem_params["beta"]            = 1.0
+#problem_params["alpha"]           = 3.75
+#problem_params["beta"]            = 1.01
 
 exp_problem = ExponentialProblem( problem_params, force_init = True )
 
@@ -98,7 +99,7 @@ def run_sgld( problem, T, h, A, c, theta=None, x = None, verbose_rate = 100, use
   if use_omega is False:
     omega = 0
   else:
-    omega = np.random.randint(T)
+    omega = 10**6
     
   alpha = 0.0
   OMEGAS = [omega]
@@ -128,9 +129,11 @@ def run_sgld( problem, T, h, A, c, theta=None, x = None, verbose_rate = 100, use
     p = np.random.randn()
     current_p = p
       
-    true_grad = problem.true_abc_gradient( theta, true_gradients )
+    #true_grad = problem.true_abc_gradient( theta, true_gradients )
+    true_grad = problem.true_gradient( theta, true_gradients )
     
-    two_side_grad_U = problem.two_sided_gradient( theta, x, omega, c, two_gradients, S )
+    #two_side_grad_U = problem.two_sided_gradient( theta, x, omega, c, two_gradients, S )
+    two_side_grad_U = problem.two_sided_sl_gradient( theta, x, omega, c, two_gradients, S )
     #one_side_grad_U = problem.one_sided_gradient( theta, x, omega, c, one_gradients )
     
     # divide by theta to transform var to 
[…]

### 2015-02-24 code/heavyside_nn.py

@@ -1,0 +1,39 @@
+
+
+
+class layer( object ):
+  def __init__( self, nbr_input_units, nbr_hidden_units ):
+    self.nbr_input_units  = nbr_input_units
+    self.nbr_hidden_units = nbr_hidden_units
+    
+  def forward( self, inputs ):
+    N,D = inputs.shape
+    
+    A = np.dot( inputs, self.W ) + self.b
+    
+    I = pp.find( A >= 0 )
+    
+    H = np.zeros( (N,n) )
+    
+    H[I] = 1
+    return H
+    
+    
+class heavyside_nn(object):
+  def __init__( self ):
+    self.layers = []
+    self.nbr_parameters = 0
+    
+  def add_layer( self, layer ):
+    self.layers.append( layer )
+    self.nbr_parameters += layer.nbr_input_units*layer.nbr_hidden_units + layer.nbr_hidden_units
+    
+  def forward( self, inputs ):
+    
+    for layer in self.layers:
+      outputs = layer.forward( inputs )
+      inputs = outputs
+      
+    return outputs
+    
+if __name__ == "__main__":

### 2015-02-24 code/lokta_volterra.py

@@ -1,0 +1,179 @@
+# 2 species
+h = function(x, pars) {
+   hazs = numeric(length(pars))
+   hazs[1] = pars[1]*x[1]
+   hazs[2] = pars[2]*x[1]*x[2]
+   hazs[3] = pars[3]*x[2]
+   return(hazs)
+}
+
+smat = matrix(0,nrow=2,ncol=3)
+smat[1,1] = 1; smat[1,2] = -1
+smat[2,2] = 1; smat[2,3] = -1
+rownames(smat) = c("Prey", "Predator")
+
+initial = c(100, 100)
+
+pars = c(0.5,0.0025,0.3)
+
+model = create_model(smat, h, initial, pars)
+
+
+f = get_f = function(x, pars)
+{
+   fmat = matrix(0, nrow=3, ncol=2)
+   fmat[1,1] = pars[1]
+   fmat[2,1] = pars[2]*x[2]
+   fmat[2,2] = pars[2]*x[1]
+   fmat[3,2] = pars[3]
+   return(fmat)
+}
+model = create_model(smat, h, initial, pars, f)
+l2 = lna(model, maxtime=50, ddt=0.1, TRUE)
+plot(g[,1], g[,2], type="l")
+
+
+# 3 species
+h = function(x, pars) {
+   hazs = numeric(length(pars))
+   hazs[1] = pars[1]*x[1]
+   hazs[2] = pars[2]*x[1]*x[2]
+   hazs[3] = pars[3]*x[2]*x[3]
+   hazs[4] = pars[4]*x[3]
+   return(hazs)
+}
+
+smat = matrix(0,nrow=3,ncol=4)
+smat[1,1] = 1; smat[1,2] = -1
+smat[2,2] = 1; smat[2,3] = -1
+smat[3,3] = 1; smat[3,4] = 1
+rownames(smat) = c("Prey", "Predator","Other")
+
+initial = c(10, 100,10)
+
+pars = c(0.5,0.0025,0.3, 0.6)
+
+model = create_model(smat, h, initial, pars)
+g = gillespie(model, maxtime=50)
+
+f = get_f = function(x, pars)
+{
+   fmat = matrix(0, nrow=3, ncol=2)
+   fmat[1,1] = pars[1]
+   fmat[2,1] = pars[2]*x[2]
+   fmat[2,2] = pars[2]*x[1]
+   fmat[3,2] = pars[3]
+   return(fmat)
+}
[…]

### 2015-02-24 code/working_code.py

@@ -1,0 +1,228 @@
+import numpy as np
+import scipy as sp
+from scipy import stats as spstats
+import pylab as pp
+
+
+def accept_move(log_acceptance):
+  accept = False
+  if log_acceptance < 0:
+    if np.random.rand() < np.exp( log_acceptance ):
+      # accept downward move
+      accept=True
+  else:
+    # accept upward move
+    accept = True
+  return accept
+
+def hamiltonian_accept( problem, theta, theta_proposal, x, x_proposal, p, p_proposal ):
+  # using kernel_epsilon( observations | x )
+  loglike_x          = problem.loglike_x( x )
+  loglike_x_proposal = problem.loglike_x( x_proposal )
+  
+  # a log-normal proposal, so we need to compute this
+  loglike_q_from_proposal_to_theta = -0.5*np.dot(p_proposal.T, p_proposal)
+  loglike_q_from_theta_to_proposal = -0.5*np.dot(p.T,p)
+  
+  # loglike_prior_theta
+  loglike_prior_theta           = problem.loglike_prior( theta )
+  loglike_prior_theta_proposal  = problem.loglike_prior( theta_proposal )
+  
+  log_acceptance =  loglike_x_proposal + loglike_prior_theta_proposal + loglike_q_from_proposal_to_theta \
+                  - loglike_x          - loglike_prior_theta          - loglike_q_from_theta_to_proposal
+                  
+  if accept_move(log_acceptance):
+    x         = x_proposal
+    theta     = theta_proposal
+    loglike_x = loglike_x_proposal
+    p         = p_proposal
+    
+  return theta, x, loglike_x 
+
+def omega_flip(problem, theta, x, omega, loglike_x ):
+  
+  omega     = np.random.randint(10**6)
+  x         = problem.simulate( theta, omega )
+  loglike_x = problem.loglike_x( x )
+    
+  return theta, x, omega, loglike_x
+    
+def omega_sample(problem, theta, x, omega, loglike_x ):
+  
+  omega_proposal     = np.random.randint(10**6)
+  x_proposal         = problem.simulate( theta, omega_proposal )
+  loglike_x_proposal = problem.loglike_x( x_proposal )
+
+  log_acceptance =  loglike_x_proposal  \
+                  - loglike_x        
+  
+  if accept_move(log_acceptance):
+    x         = x_proposal
[…]

### 2015-02-25 papers/uai-2015/draft.tex
working thermo and draft edits

@@
 However, due to the difficulty in scaling likelihood estimates, ABC remains useful for relatively low-dimensional problems.
-[REASONS] We introduce Hamiltonian ABC (HABC), a set of likelihood-free algorithms that apply recent advances in scaling Bayesian learning using stochastic gradients with Hamiltonian dynamics to simulation models.
-[WHY GOOD: low nbr sims and non-random walk]  We find that a small number forward simulations can effectively approximate the true ABC gradient, allowing Hamiltonian dynamics to efficiently traverse parameter spaces.
-
-  Models in simulation-based science domains are considered likelihood-free (LF) since they are not defined by a probabilistic model, but by a simulator that maps parameters to pseudo-observations.
+We introduce Hamiltonian ABC (HABC), a set of likelihood-free algorithms that apply recent advances in scaling Bayesian learning using Hamiltonian Monte Carlo (HMC) and stochastic gradients.
+We find that a small number forward simulations can effectively approximate the ABC gradient, allowing Hamiltonian dynamics to efficiently traverse parameter spaces.
+We also describe a new simple yet general approach of incorporating random seeds into the state of the Markov chain, further reducing the random walk behavior of HABC.
+We demonstrate HABC on several typical ABC problems, and show that HABC performs comparably to regular Bayesian inference on a high-dimensional problem from machine learning.
+  %
+  %
+  % Models in simulation-based science domains are considered likelihood-free (LF) since they are not defined by a probabilistic model, but by a simulator that maps parameters to pseudo-observations.
 To perform parameter inference in the models, a powerful set of statistical procedures called Approximate Bayesian Computation (ABC) can be applied.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-In simulation-based science, models are defined by simulator and parameters.
-These are likelihood-free models because, in contrast to probabilistic models, their likelihoods are either intractable or must be approximated by simulations.
+In simulation-based science, models are defined by a simulator and its parameters.
+These are called {\em likelihood-free} models because, in contrast to probabilistic models, their likelihoods are either intractable to compute or must be approximated by simulations.
+To perform inference in likelihood-free models, a broad class of algorithms called Approximate Bayesian Computation \cite{beaumont2002approximate,marjoram2003markov,sisson2007sequential,sisson:2010,marin:2012,fan:2013} are employed.
 
-Recent advances in Bayesian inference within the {\em big data} domain have combined the efficiency of stochastic gradient algorithms used for optimization with Hamiltonian dynamics;  these include Stochastic Gradient Langevin Dynamics (SGLD) \cite{welling2011bayesian}, Stochastic Gradient HMC \cite{chen2014stochastic}, and Stochastic Gradient Thermostats \cite{ding2014bayesian}.
-By using small batches of the full data set these methods are able to quickly compute a stochastic approximation to the true gradient, which with a small enough step-size and with an appropriate amount of injected noise, can produce samples from the true posterior.
+At the core of every ABC algorithm is simulation.
+To evaluate the quality of a parameter vector $\thetav$, a simulation is run using $\thetav$ as inputs and producing outputs $\x$.
+If the pseudo-data $\x$ is ``close'' to observations $\y$, then $\thetav$ is kept as a sample from the approximate posterior.
+Parameters $thetav$ are then adjusted, depending upon the algorithm, to obtain the next sample.
+
+In ABC, there is a fundamental trade-off between the computation required to obtain independent samples and the approximation to the true posterior.
+If the parameter measuring closeness is too small, then samplers ``mix'' poorly; on the other hand, if it is too large, then the approximation is poor.
+As the dimension of the parameters grows, the problem worsens, just as it does for general Bayesian inference with probabilistic models, but it is more acute for ABC due to its simulation requirement.
+There is therefore a deep interest in improving the efficiency of ABC samplers (in terms of computation per independent sample).
+In this paper we address this issue directly by using Hamiltonian dynamics to approximately sample from likelihood-free models with high-dimensional parameters.
+
+Hamiltonian Monte Carlo (HMC) \cite{duane1987hybrid, neal2011mcmc} is perhaps the only Bayesian inference algorithm that scales to high-dimensional parameter spaces.
+The core computation of HMC is the gradient of the log-likelihood.
+Two problems arise if we consider HMC for ABC: one, how can the gradients be computed for high-dimensional likelihood-free models, and two, given a stochastic approximation to the gradient, can a valid HMC algorithm be derived?
+
+ To answer the latter, we turn to recent developments in scaling Bayesian inference using HMC and stochastic gradients \cite{welling2011bayesian,chen2014stochastic,ding2014bayesian}.
+We call these {\em stochastic gradient Hamiltonian dynamics} (SGHD) algorithms.
+SGHD are computationally efficient for two reasons.
+First, they avoid computing the gradient of the log-likelihood over the entire data set, instead approximating it using small batches of data, i.e.
+computing stochastic gradients.
+Second, they can maintain reasonable approximations to the Hamiltonian dynamics and therefore avoid a Metropolis-Hastings correction step involving the full data set.
+Different strategies are employed to do this: small step-sizes combined with  Langevin dynamics \cite{welling2011bayesian}, using friction to prevent accumulation of errors in the Hamiltonian \cite{chen2014stochastic}, and using a thermostat to control the temperature of the Hamiltonian \cite{ding2014bayesian}.
+Each of these strategies can be used by HABC.
+
+%Recent advances in Bayesian inference within the {\em big data} domain have combined the efficiency of stochastic gradient algorithms used for optimization with Hamiltonian dynamics;  these include Stochastic Gradient Langevin Dynamics (SGLD) \cite{welling2011bayesian}, Stochastic Gradient HMC \cite{chen2014stochastic}, and Stochastic Gradient Thermostats \cite{ding2014bayesian}.
+%By using small batches of the full data set these methods are able to quickly compute a stochastic approximation to the true gradient, which with a small enough step-size and with an appropriate amount of injected noise, can produce samples from the true posterior.
 The Hamiltonian dynamics avoid random walk behavior by providing momentum to the sampler allowing it to reach far away regions of parameter space.
@@
 
-Approximate Bayesian Computation (ABC) is a set of inference procedures for so-called {\em likelihood-free} models \cite{abccitations}.
+In HABC, we use forward simulations to approximate the likelihood-free gradient.
+The key difference between SGHD methods and HABC is that the stochasticity of the gradient does not come from approximating the full data gradient with a batch gradient, but by the stochasticity of the simulator.
+It is therefore not the expense of the simulator (though this could very well be the case for many interesting simulation-based models -- see Section~\ref{sec:conclusion}) that requires an approximation to the gradient, but the likelihood-free nature of the problem.
+
+There are several difficulties in estimating gradients of likelihood-free models that we address with HABC.
+The first is do to the form of the ABC log-likelihood.
+As we show in Section~\ref{sec:abc}, using a conditional model for $\pi( \x | \thetav )$ provides an estimate of the ABC likelihood that is less sensitive to $\eps$ and therefore is more conducive to stochastic gradient computations.
+The second difficulty is that for high-dimensional parameter spaces, computing the gradients naively (i.e.
+by finite differences \cite{kiefer1952stochastic}) can squash the gains brought by the Hamiltonian dynamics.
+Fortunately, we can use existing stochastic approximation algorithms \cite{spall1992multivariate,spall2000adaptive} that can be used to compute unbiased estimators of the gradient with a small number of forward simulations that is {\em independent} of the parameter dimension.
+The {\em stochastic perturbation stochastic approximation} (SPSA) \cite{spall1992multivariate} is described in Section~\ref{sec:habc}
+
+%Approximate Bayesian Computation (ABC) is a set of inference procedures for so-called {\em likelihood-free} models \cite{abccitations}.
 In contrast to typical Bayesian modeling, where the parameters of a probabilistic model of data are the focus, the primary interest in ABC are the parameters of a simulation model.
@@
 
-The approximation to the true posterior can be made arbitrarily small, in principle, by controlling the kernel width parameter $\epsilon$.
+%The approximation to the true posterior can be made arbitrarily small, in principle, by controlling the kernel width parameter $\epsilon$.
 As $\epsilon \rightarrow 0$, however, simulated data must be closer and closer to the observations (or receive zero likelihood).
@@
 
-In this paper we adopt the developments in scaling Bayesian inference to ABC.
+%In this paper we adopt the developments in scaling Bayesian inference to ABC.
 To do this we use forward simulations to approximate the likelihood-free gradient.
@@
 
-The approximate gradients used by HABC can be noisy and difficult to manage is care is not taken.
+%The approximate gradients used by HABC can be noisy and difficult to manage is care is not taken.
 To address this we show how a local Gaussian approximation of the simulator output can improve the stability of the gradient estimates over a direct estimate using the ABC kernel function.
@@
 A further innovation of this paper is the use of common random numbers (CRN) to improve the efficiency of the Hamiltonian dynamics.
-By using the same random seed for the finite difference approximation, the first order noise in the gradient is reduced;  by keeping the random seeds fixed for several consecutive steps, the second order stochasticity of one gradient estimate to the next is eliminated.
-We show that this is a valid MCMC procedure.
+The idea behind CRNs is to use the same set of random seeds for estimating a gradient by FD or SPSA, i.e.
+when simulating $\pi(\x| \theta + d\theta)$ and $\pi(\x| \theta - d\theta)$ use the same random seeds.
+This was applied successfully to SPSA \cite{kleinman1999simulation} (and is analogous to using the same mini-batch in stochastic gradient methods).
+We extend and simplify this approach by including the random seeds $\omega$ into the state of the Markov chain;  by keeping the random seeds fixed for several consecutive steps, the second order gradient stochasticity is greatly reduced.
+We show that this doing this produces is a valid MCMC procedure.
+This approach is not exclusive to HABC; we apply successfully to ABC-MCMC as well.
 
-In Section~\ref{sec:sghd-data} we review three approaches to stochastic gradient inference using Hamiltonian dynamics: SGLD, SGHMC, and SGNHT.
-We briefly review ABC-MCMC in Section~\ref{sec:abcmcmc}, followed by Hamiltonian ABC in Section~\ref{sec:habc}, where  we will show how to improve the stability of the gradient estimates by using CRNs and local density estimators of the simulator.
+We briefly review ABC in Section~\ref{sec:abc}.
+In Section~\ref{sec:scaling} we review three approaches to stochastic gradient inference using Hamiltonian dynamics: SGLD, SGHMC, and SGNHT.
+We then introduce Hamiltonian ABC in Section~\ref{sec:habc}, where  we will show how to improve the stability of the gradient estimates by using CRNs and local density estimators of the simulator.
 Extensions to high-dimensional parameter spaces are also discussed.
-In Section~\ref{sec:toy} we show how HABC behaves on a simple one-dimensional problem, then in Section~\ref{sec:experiments} we compare HABC with ABC-MCMC for two problems: a low-dimensional model of chaotic population dynamics and a high-dimensional problem.
+In Section~\ref{sec:demo} we show how HABC behaves on a simple one-dimensional problem, then in Section~\ref{sec:experiments} we compare HABC with ABC-MCMC for two problems: a low-dimensional model of chaotic population dynamics and a high-dimensional problem.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{APPROXIMATE BAYESIAN COMPUTATION}\label{sec:abcmcmc}
+\section{APPROXIMATE BAYESIAN COMPUTATION}\label{sec:abc}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{equation}
-where $\pi(\thetav)$ is a prior distribution over parameters $\thetav \in {\rm I\!R}^{D}$ and $\pi( \y_1, \ldots, \y_N  | \thetav )$ is the likelihood of $N$ data observations, where $\y_n \in {\rm I\!R}^J$.
+where $\pi(\thetav)$ is a prior distribution over parameters $\thetav \in {\rm I\!R}^{D}$ and $\pi( \y_1, \ldots, \y_N  | \thetav )$ is the likelihood of $N$ data observations, where $\y_i \in {\rm I\!R}^J$.
 In ABC, the vector of $J$ observations are typically informative statistics of the raw observations.
+It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior ([KEEP?]).
   The simulator is treated as generator of random pseudo-observations, i.e.
@@
 %
-If the simulations are part of the Markov chain, the algorithm corresponds to the pseudo-marginal (PM) sampler \cite{delmoral2008,andrieu2009pseudo}, otherwise it is a marginal sampler \cite{marjoram2003markov,sisson:2010}.
+If the simulations are part of the Markov chain, the algorithm corresponds to the pseudo-marginal (PM) sampler \cite{andrieu2009pseudo}, otherwise it is a marginal sampler \cite{marjoram2003markov,sisson:2010}.
 For this paper we will be interested in the PM sampler because this is equivalent to having the random states that generated the simulation outputs in the state of the Markov chain, which we will use within a valid ABC sampling algorithm in Section~\ref{sec:habc}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{SCALING BAYESIAN INFERENCE USING HAMILTONIAN DYNAMICS} \label{sec:sghd-data}
+\section{SCALING BAYESIAN INFERENCE USING HAMILTONIAN DYNAMICS} \label{sec:scaling}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{Hamiltonian Monte Carlo}
+\subsection{Hamiltonian Monte Carlo}\label{sec:hmc}
 
 Hamiltonian dynamics are often necessary to adequately explore the target distribution of high-dimensional parameter spaces.
-By proposing parameters that are far from the current location and yet have high acceptance probability, Hamiltonian Monte Carlo \cite{duane1987hybrid, neal2011mcmc}  can efficiently avoid random walk behavior that can render proposals in high-dimensions inadequate.
+By proposing parameters that are far from the current location and yet have high acceptance probability, Hamiltonian Monte Carlo \cite{duane1987hybrid, neal2011mcmc}  can efficiently avoid random walk behavior that can render proposals in high-dimensions painfully slow to mix.
 
@@
 
-\subsection{Stochastic Gradient Hamiltonian Dynamics}
+\subsection{Stochastic Gradient Hamiltonian Dynamics}\label{se:sghd}
 If the log-likelihood over the full data set is replaced with a batch estimate, as is done for the following {\em stochastic gradient Hamiltonian dynamics} (SGHDs) algorithms, then the error in simulating the Hamiltonian dynamics comes not only from the discretization, but from the variance of the stochastic gradient.
@@
 %\subsection{Spall's method: SPSA}
-In the gradient-free setting, Spall \cite{spall_spsa} provides a stochastic approximate to the true gradient using only 2 forward simulations (function evaluations).
-This is in contrast to multivariate finite-difference stochastic approximation FDSA \cite{kiefer1952} requiring $2p$ evaluations.
+In the gradient-free setting, Spall \cite{spall1992multivariate} provides a stochastic approximate to the true gradient using only 2 forward simulations (function evaluations).
+This is in contrast to multivariate finite-difference stochastic approximation FDSA \cite{kiefer1952stochastic} requiring $2p$ evaluations.
 
@@
 \begin{itemize}
-  \item Log likelihood is much noisier in  1, and the 2 there are good reasons why this is a better choice: is no longer depends on $epsilon$ (as the true gradient shouldn't), convolving the conditional density is the right thing to do, if possible (though it might not be necessary to perform ABC)
   \item In both cases, we rewrite the expression conditioning on a set of random seeds omega.
@@
 [CITE NEAL How to view an MCMC...]
-Another variance reduction technique called the method of {\em common random numbers} (CRN) \cite{kleinman1999} sets a common seed for the random number generator (RNG) for both calls to the simulator.
+Another variance reduction technique called the method of {\em common random numbers} (CRN) \cite{kleinman1999simulation} sets a common seed for the random number generator (RNG) for both calls to the simulator.
 This technique can remove the effect of the simulator noise in the gradient estimate, leaving on the randomness of the perturbation masks as the source of noise.
@@
 
-The ABC likelihood is proportional to the convolution between a kernel around the observation statistics $\y$ and the generator of pseudo-statistics $\x$, i.e.
-the simulator.
-Since pseudo-statistics can only be generated, we resort to a Monte Carlo estimate for the ABC likelihood based on $S$ draws from the simulator using the same parameter vector $\thetav$ but different random seeds:
+Using the same parameter vector $\thetav$ but different random seeds:
 \begin{eqnarray}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Demonstration}\label{sec:toy}
+\section{Demonstration}\label{sec:demo}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{Blowfly}
+\subsection{Blowfly}\label{sec:bf}
 [FROM UAI] Adult blowfly populations exhibit dynamic behavior for which several competing population models exist.
@@
 
-\subsection{Bayesian Autoencoders}
+\subsection{Bayesian Autoencoders}\label{sec:auto}
 We perform Bayesian inference on an autoencoder neural network using MNIST images as observations.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{DISCUSSION} \label{discussion}
+\section{DISCUSSION} \label{sec:discussion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{CONCLUSION} \label{conclusion}
+\section{CONCLUSION} \label{sec:conclusion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-02-25 code/exponential_problem/exp_wrapper.py

@@ -1,0 +1,373 @@
+from abcpy.factories import *
+from abcpy.problems.exponential import *
+#from sa_algorithms import *
+from scipy import stats as spstats
+import pylab as pp
+
+problem_params = default_params()
+
+problem_params["N"] = 5
+problem_params["q_stddev"] = 0.5
+problem_params["theta_star"]      = 0.1
+problem_params["epsilon"] = 0.1*np.sqrt( 1.0 / (problem_params["N"]*problem_params["theta_star"]**2) )
+problem_params["alpha"]           = 1.0
+problem_params["beta"]            = 1.0
+#problem_params["alpha"]           = 3.75
+#problem_params["beta"]            = 1.01
+
+exp_problem = ExponentialProblem( problem_params, force_init = True )
+
+state_params = state_params_factory.scrape_params_from_problem( exp_problem, S=1 )
+
+# set is_marginal to true so using only "current" state will force a re-run
+mcmc_params  = mcmc_params_factory.scrape_params_from_problem( exp_problem, type="mh", is_marginal = True, nbr_samples = 1 )
+
+# ignore algo, just doing this to get state object
+algo_params = { "modeling_approach"  : "kernel",
+                "observation_groups" : exp_problem.get_obs_groups(),
+                "state_params"       : state_params,
+                "mcmc_params"        : mcmc_params,
+                "algorithm"          : "model_mcmc"
+              }
+algo, model, state  = algo_factory.create_algo_and_state( algo_params )
+recorder     = recorder_factory.create_recorder( {} )
+
+
+class generate_exponential( object ):
+  def __init__( self, abcpy_problem ):
+    self.p = abcpy_problem
+    self.kernel = spstats.norm( self.p.obs_statistics, self.p.epsilon )
+    self.y = self.p.obs_statistics[0]
+    self.x_epsilon = None
+    
+    self.posterior = spstats.gamma( self.p.alpha+self.p.N, 0, 1.0/(self.p.beta+self.p.obs_sum ))
+    
+  def prior_rand(self):
+    return self.p.theta_prior_rand()
+    
+  def propose( self, theta ):
[…]

### 2015-02-25 code/exponential_problem/thermo_sample.py

@@ -1,0 +1,76 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+from scipy import stats as spstats
+from exp_wrapper import *
+from abc_sgld.code.working_code import *
+
+keep_x        = True 
+init_seed     = 1
+T             = 5000 # nbr of samples
+verbose_rate  = 1000
+C             = 5.01    # injected noise variance parameter
+eta           = 0.01 # step size for Hamiltoniam dynamics
+#h = 0.0005
+
+# params for gradients
+d_theta = 0.001  # step size for gradient estimate
+S       = 5 
+grad_params = {}
+  
+# keep theta within bounds
+lower_bounds = np.array([0.001])
+upper_bounds = np.array([np.inf])
+
+
+# ----------------------- #
+# common ransom seeds     #
+# ----------------------- #
+use_omega    = True    # use fixed random seeds for simulations
+omega_rate   = 0.01    # probabilty of changing omegas
+omega_switch = True    # if true, randomly change omegas
+omega_sample = False   # sample omegas instead
+assert omega_switch and omega_sample is False, "only get to do one type of omega update"
+omega_params = {"use_omega":use_omega, \
+                "omega_rate":omega_rate, \
+                "omega_switch":omega_switch,\
+                "omega_sample":omega_sample}  
+
+if __name__ == "__main__": 
+  pp.close('all')
+  
+  chain_id = 1
+  params = {}
+  params["chain_id"]  = chain_id
+  params["init_seed"] = init_seed
+  params["T"]       = T
+  params["S"]       = S
+  params["d_theta"] = d_theta
+  params["eta"]     = eta
+  params["C"]       = C
+  params["omega_params"] = omega_params
+  params["verbose_rate"] = verbose_rate
+  params["grad_params"]  = {"logs":{"true":[],"true_abc":[],"2side_keps":[],"2side_sl":[]},\
+                            "record_2side_sl_grad":False, "record_2side_keps_grad":False,"record_true_abc_grad":True,"record_true_grad":False}
+  params["lower_bounds"] = lower_bounds
+  params["upper_bounds"] = upper_bounds
+  params["keep_x"]       = keep_x
+  
+  theta0 = np.array( [0.2])
+  x0     = None
+  
+  # initialize the simulator
+  np.random.seed(init_seed)
+  problem = generate_exponential( exp_problem )
+  params["grad_func"] = problem.two_sided_sl_gradient
+
[…]

### 2015-02-25 code/working_code.py

@@ -3,7 +3,23 @@ import scipy as sp
 from scipy import stats as spstats
 import pylab as pp
 
+max_int_for_omega = 10**8
 
+def bounce_off_boundaries( theta, p, lower_bounds, upper_bounds ):
+  D = len(theta)
+  
+  for d in range(D):
+    if theta[d] < lower_bounds[d]:
+      theta[d] = lower_bounds[d] + (lower_bounds[d]-theta[d]) #from Neal Handbook, p37
+      p[d] *= -1
+
+    if theta[d] > upper_bounds[d]:
+      theta[d] = upper_bounds[d] - (theta[d]-upper_bounds[d]) #from Neal Handbook, p37
+      p[d] *= -1
+  return theta, p
+  
+
+    
 def accept_move(log_acceptance):
   accept = False
   if log_acceptance < 0:
@@ -39,6 +55,13 @@ def hamiltonian_accept( problem, theta, theta_proposal, x, x_proposal, p, p_prop
     
   return theta, x, loglike_x 
 
+def init_omega( use_omega ):
+  if use_omega is False:
+    omega = None
+  else:
+    omega = np.random.randint(max_int_for_omega)
+  return omega
+  
 def omega_flip(problem, theta, x, omega, loglike_x ):
   
   omega     = np.random.randint(10**6)
@@ -65,14 +88,10 @@ def omega_sample(problem, theta, x, omega, loglike_x ):
   
 def run_mcmc( problem, params, theta, x = None ):
   T         = params["T"]
+  S         = params["S"]
   use_omega = params["use_omega"]
 
-  # random seed init
-  if use_omega is False:
-    omega = None
-  else:
-    omega = np.random.randint(10**6)
-    
+  omega = init_omega( use_omega )
   
   # pseudo-data
   if x is None:
@@ -92,7 +111,7 @@ def run_mcmc( problem, params, theta, x = None ):
     # sample (theta, x)    #
     # -------------------- #
     theta_proposal = problem.propose( theta )
-    x_proposal     = problem.simulate( theta_proposal, omega )
+    x_proposal     = problem.simulate( theta_proposal, omega, S )
     
     # using kernel_epsilon( observations | x )
     loglike_x_proposal = problem.loglike_x( x_proposal )
@@ -121,7 +140,7 @@ def run_mcmc( problem, params, theta, x = None ):
     # samples omegas  #
     # --------------- #
     if use_omega and np.random.rand() < omega_rate:
[…]

### 2015-02-25 code/exponential_problem/thermo_sample.py
added sghmc to working

@@ -9,12 +9,12 @@ keep_x        = True
 init_seed     = 1
 T             = 5000 # nbr of samples
 verbose_rate  = 1000
-C             = 5.01    # injected noise variance parameter
-eta           = 0.01 # step size for Hamiltoniam dynamics
+C             = 10.01    # injected noise variance parameter
+eta           = 0.001 # step size for Hamiltoniam dynamics
 #h = 0.0005
 
 # params for gradients
-d_theta = 0.001  # step size for gradient estimate
+d_theta = 0.01  # step size for gradient estimate
 S       = 5 
 grad_params = {}
   
@@ -68,7 +68,8 @@ if __name__ == "__main__":
   np.random.seed(init_seed + 1000*chain_id)
   
   # run algorithm
-  outs = run_thermostats( problem, params, theta0, x0 )
+  #outs = run_thermostats( problem, params, theta0, x0 )
+  outs = run_sghmc( problem, params, theta0, x0 )
   
   # view results of single chain
   problem.view_single_chain( outs )

### 2015-02-25 code/working_code.py

@@ -148,7 +148,126 @@ def run_mcmc( problem, params, theta, x = None ):
     if np.mod(t+1,verbose_rate)==0:
       print "t = %04d    loglik = %3.3f    theta = %3.3f    x = %3.3f"%(t+1,loglike_x,theta,x)
   return np.squeeze(np.array( THETAS)), np.squeeze(np.array(X)), np.squeeze(np.array(LL)), np.squeeze(np.array(OMEGAS))
+
+def run_sghmc( problem, params, theta, x = None ):
   
+  T             = params["T"]
+  S             = params["S"]
+  verbose_rate  = params["verbose_rate"]
+  omega_params  = params["omega_params"]
+  deltaC        = params["C"]  # injected noise added to Variance
+  eta           = params["eta"] # Hamiltonain step size
+  d_theta       = params["d_theta"] # step used for estimating gradients
+  grad_U_func   = params["grad_func"]
+  grad_U_params = params["grad_params"]
+  
+  keep_x        = params["keep_x"] # the HABC methods do not necessarily have x in the state space
+  
+  # keep theta within bounds
+  lower_bounds = params["lower_bounds"]
+  upper_bounds = params["upper_bounds"]
+  
+  D   = len(theta)
+  
+  outputs = {}
+  # two_gradients = []
+  # one_gradients = []
+  # true_gradients = []
+  
+  omega = init_omega( omega_params )
+  
+  # pseudo-data
+  if keep_x:
+    if x is None:
+      x = problem.simulate( theta, omega, S )
+    assert len(x.shape) == 2, "x should be S by dim"
+    loglike_x = problem.loglike_x( x )
+    X      = [x]
+    LL     = [loglike_x]
+  else:
+    x         = None
+    loglike_x = None
+    
+  # initialize momentum  
+  p = np.random.randn(D)
+  
+  THETAS = [theta]  
+  OMEGAS = [omega]
+  
+  grads = np.zeros( (T,D) )
+  
+  # sample...
+  for t in xrange(T):
+    
+    # ----------------------------- # 
+    # simulate trajectory for theta #
+    # ----------------------------- #
+    
+    # estimate stochastic gradient
+    grad_U = grad_U_func( theta, d_theta, omega, S, grad_U_params)
+
+    #grad_U = problem.one_sided_gradient( theta, x, omega, c )
[…]

### 2015-02-25 code/exponential_problem/exp_wrapper.py
added sold to working code

@@ -9,7 +9,7 @@ problem_params = default_params()
 problem_params["N"] = 5
 problem_params["q_stddev"] = 0.5
 problem_params["theta_star"]      = 0.1
-problem_params["epsilon"] = 0.1*np.sqrt( 1.0 / (problem_params["N"]*problem_params["theta_star"]**2) )
+problem_params["epsilon"] = 0.25*np.sqrt( 1.0 / (problem_params["N"]*problem_params["theta_star"]**2) )
 problem_params["alpha"]           = 1.0
 problem_params["beta"]            = 1.0
 #problem_params["alpha"]           = 3.75

### 2015-02-25 code/exponential_problem/thermo_sample.py

@@ -7,15 +7,15 @@ from abc_sgld.code.working_code import *
 
 keep_x        = True 
 init_seed     = 1
-T             = 5000 # nbr of samples
+T             = 15000 # nbr of samples
 verbose_rate  = 1000
 C             = 10.01    # injected noise variance parameter
-eta           = 0.001 # step size for Hamiltoniam dynamics
+eta           = 0.01 # step size for Hamiltoniam dynamics
 #h = 0.0005
 
 # params for gradients
 d_theta = 0.01  # step size for gradient estimate
-S       = 5 
+S       = 10 
 grad_params = {}
   
 # keep theta within bounds
@@ -69,7 +69,8 @@ if __name__ == "__main__":
   
   # run algorithm
   #outs = run_thermostats( problem, params, theta0, x0 )
-  outs = run_sghmc( problem, params, theta0, x0 )
+  #outs = run_sghmc( problem, params, theta0, x0 )
+  outs = run_sgld( problem, params, theta0, x0 )
   
   # view results of single chain
   problem.view_single_chain( outs )

### 2015-02-25 code/working_code.py

@@ -149,6 +149,119 @@ def run_mcmc( problem, params, theta, x = None ):
       print "t = %04d    loglik = %3.3f    theta = %3.3f    x = %3.3f"%(t+1,loglike_x,theta,x)
   return np.squeeze(np.array( THETAS)), np.squeeze(np.array(X)), np.squeeze(np.array(LL)), np.squeeze(np.array(OMEGAS))
 
+def run_sgld( problem, params, theta, x = None ):
+  
+  T             = params["T"]
+  S             = params["S"]
+  verbose_rate  = params["verbose_rate"]
+  omega_params  = params["omega_params"]
+  eta           = params["eta"] # Hamiltonain step size
+  d_theta       = params["d_theta"] # step used for estimating gradients
+  grad_U_func   = params["grad_func"]
+  grad_U_params = params["grad_params"]
+  
+  keep_x        = params["keep_x"] # the HABC methods do not necessarily have x in the state space
+  
+  # keep theta within bounds
+  lower_bounds = params["lower_bounds"]
+  upper_bounds = params["upper_bounds"]
+  
+  D   = len(theta)
+  
+  outputs = {}
+  # two_gradients = []
+  # one_gradients = []
+  # true_gradients = []
+  
+  omega = init_omega( omega_params )
+  
+  # pseudo-data
+  if keep_x:
+    if x is None:
+      x = problem.simulate( theta, omega, S )
+    assert len(x.shape) == 2, "x should be S by dim"
+    loglike_x = problem.loglike_x( x )
+    X      = [x]
+    LL     = [loglike_x]
+  else:
+    x         = None
+    loglike_x = None
+    
+  
+  
+  THETAS = [theta]  
+  OMEGAS = [omega]
+  
+  grads = np.zeros( (T,D) )
+  
+  # sample...
+  for t in xrange(T):
+    
+    # ----------------------------- # 
+    # simulate trajectory for theta #
+    # ----------------------------- #
+    
+    # estimate stochastic gradient
+    grad_U = grad_U_func( theta, d_theta, omega, S, grad_U_params)
+    #
+    if grad_U_params["record_2side_sl_grad"]:
+      grad_U_dummy = problem.two_sided_sl_gradient( theta, d_theta, omega, S, grad_U_params )
+    if grad_U_params["record_2side_keps_grad"]:
+      grad_U_dummy = problem.two_sided_keps_gradient( theta, d_theta, omega, S, grad_U_params )
[…]

### 2015-02-25 code/exponential_problem/exp_wrapper.py
added mcmc

@@ -6,10 +6,10 @@ import pylab as pp
 
 problem_params = default_params()
 
-problem_params["N"] = 5
+problem_params["N"] = 50
 problem_params["q_stddev"] = 0.5
-problem_params["theta_star"]      = 0.1
-problem_params["epsilon"] = 0.25*np.sqrt( 1.0 / (problem_params["N"]*problem_params["theta_star"]**2) )
+problem_params["theta_star"]      = 0.2
+problem_params["epsilon"] = 0.1*np.sqrt( 1.0 / (problem_params["N"]*problem_params["theta_star"]**2) )
 problem_params["alpha"]           = 1.0
 problem_params["beta"]            = 1.0
 #problem_params["alpha"]           = 3.75
@@ -49,7 +49,9 @@ class generate_exponential( object ):
     return self.p.theta_proposal_rand( theta )
   
   def loglike_x( self, x ):
-    return self.kernel.logpdf( x )
+    S,J = x.shape
+    L = logsumexp( self.kernel.logpdf( x ), 0 ) - np.log(S)
+    return L
     
   def loglike_prior( self, theta ):
     return self.p.theta_prior_logpdf( theta )

### 2015-02-25 code/exponential_problem/thermo_sample.py

@@ -7,15 +7,15 @@ from abc_sgld.code.working_code import *
 
 keep_x        = True 
 init_seed     = 1
-T             = 15000 # nbr of samples
+T             = 5000 # nbr of samples
 verbose_rate  = 1000
-C             = 10.01    # injected noise variance parameter
-eta           = 0.01 # step size for Hamiltoniam dynamics
+C             = 20.01    # injected noise variance parameter
+eta           = 0.001 # step size for Hamiltoniam dynamics
 #h = 0.0005
 
 # params for gradients
 d_theta = 0.01  # step size for gradient estimate
-S       = 10 
+S       = 5 
 grad_params = {}
   
 # keep theta within bounds
@@ -70,7 +70,8 @@ if __name__ == "__main__":
   # run algorithm
   #outs = run_thermostats( problem, params, theta0, x0 )
   #outs = run_sghmc( problem, params, theta0, x0 )
-  outs = run_sgld( problem, params, theta0, x0 )
+  #outs = run_sgld( problem, params, theta0, x0 )
+  outs = run_mcmc( problem, params, theta0, x0 )
   
   # view results of single chain
   problem.view_single_chain( outs )

### 2015-02-25 code/working_code.py

@@ -87,22 +87,34 @@ def omega_sample(problem, theta, x, omega, loglike_x ):
   return theta, x, omega, loglike_x
   
 def run_mcmc( problem, params, theta, x = None ):
-  T         = params["T"]
-  S         = params["S"]
-  use_omega = params["use_omega"]
-
-  omega = init_omega( use_omega )
+  T             = params["T"]
+  S             = params["S"]
+  verbose_rate  = params["verbose_rate"]
+  omega_params  = params["omega_params"]
+  # keep theta within bounds
+  lower_bounds = params["lower_bounds"]
+  upper_bounds = params["upper_bounds"]
+  
+  keep_x        = True
+  omega = init_omega( omega_params )
+  outputs = {}
   
   # pseudo-data
-  if x is None:
-    x = problem.simulate( theta, omega )
+  if keep_x:
+    if x is None:
+      x = problem.simulate( theta, omega, S )
+    assert len(x.shape) == 2, "x should be S by dim"
+    loglike_x = problem.loglike_x( x )
+    X      = [x]
+    LL     = [loglike_x]
+  else:
+    x         = None
+    loglike_x = None
   
-  loglike_x = problem.loglike_x( x )
+  p_dummy = np.random.randn(len(theta))
   
-  X      = [x]
   THETAS = [theta]
   OMEGAS = [omega]
-  LL     = [loglike_x]
   
   # sample...
   for t in xrange(T):
@@ -111,6 +123,9 @@ def run_mcmc( problem, params, theta, x = None ):
     # sample (theta, x)    #
     # -------------------- #
     theta_proposal = problem.propose( theta )
+    # bounce position off parameter boundaries
+    theta_proposal, p_dummy = bounce_off_boundaries( theta_proposal, p_dummy, lower_bounds, upper_bounds )
+    
     x_proposal     = problem.simulate( theta_proposal, omega, S )
     
     # using kernel_epsilon( observations | x )
@@ -132,22 +147,39 @@ def run_mcmc( problem, params, theta, x = None ):
       theta     = theta_proposal
       loglike_x = loglike_x_proposal
     
-    X.append(x)
-    THETAS.append(theta)
-    LL.append(loglike_x)
+    
     
     # --------------- #
     # samples omegas  #
     # --------------- #
-    if use_omega and np.random.rand() < omega_rate:
[…]

### 2015-02-26 code/exponential_problem/exp_wrapper.py
some fixes

@@ -1,5 +1,6 @@
 from abcpy.factories import *
 from abcpy.problems.exponential import *
+from abc_sgld.code.working_code import *
 #from sa_algorithms import *
 from scipy import stats as spstats
 import pylab as pp
@@ -7,7 +8,7 @@ import pylab as pp
 problem_params = default_params()
 
 problem_params["N"] = 50
-problem_params["q_stddev"] = 0.5
+problem_params["q_stddev"] = 0.01
 problem_params["theta_star"]      = 0.2
 problem_params["epsilon"] = 0.1*np.sqrt( 1.0 / (problem_params["N"]*problem_params["theta_star"]**2) )
 problem_params["alpha"]           = 1.0
@@ -260,7 +261,10 @@ class generate_exponential( object ):
     
     for time_id in [1,5,10,25,50,75,100,200,300,400,500,750,1000,1500,2000,3000,4000,5000,7500,10000,12500,15000,17500,20000,25000,30000,35000,40000,45000,50000]:
       if time_id <= len(thetas):
-        errs.append( bin_errors_1d(self.p.coarse_theta_range, self.p.posterior_cdf_bins, thetas[:time_id]) )
+        #errs.append( bin_errors_1d(self.p.coarse_theta_range, self.p.posterior_cdf_bins, thetas[:time_id]) )
+        er = cramer_vonMises_criterion( self.p.posterior_cdf_bins, thetas[:time_id], self.p.coarse_theta_range )
+        #errs.append( bin_sq_errors_1d(self.p.coarse_theta_range, self.p.posterior_cdf_bins, thetas[:time_id]) )
+        errs.append( er  )
         time_ids.append(time_id)
         nbr_sims.append(nsims[:time_id].sum()+at_burnin)
         
@@ -271,24 +275,27 @@ class generate_exponential( object ):
     
     f2 = pp.figure()
     sp1 = f2.add_subplot(1,3,1)
-    pp.semilogx( time_ids, errs, "bo-", lw=2)
+    pp.loglog( time_ids, errs, "bo-", lw=2)
     pp.xlabel( "nbr samples")
     pp.ylabel( "err")
     pp.grid('on')
     sp2 = f2.add_subplot(1,3,2)
-    pp.semilogx( nbr_sims, errs, "ro-", lw=2)
+    pp.loglog( nbr_sims, errs, "ro-", lw=2)
     pp.xlabel( "nbr sims")
[…]

### 2015-02-26 code/exponential_problem/make_illustrations.py

@@ -7,10 +7,10 @@ from abc_sgld.code.working_code import *
 
 keep_x        = True 
 init_seed     = 1
-T             = 15000 # nbr of samples
+T             = 1000 # nbr of samples
 verbose_rate  = 1000
-C             = 20.01    # injected noise variance parameter
-eta           = 0.001 # step size for Hamiltoniam dynamics
+C             = 1.01    # injected noise variance parameter
+eta           = 0.01 # step size for Hamiltoniam dynamics
 #h = 0.0005
 
 # params for gradients
@@ -26,7 +26,7 @@ upper_bounds = np.array([np.inf])
 # ----------------------- #
 # common ransom seeds     #
 # ----------------------- #
-use_omega    = True    # use fixed random seeds for simulations
+use_omega    = False    # use fixed random seeds for simulations
 omega_rate   = 1.1    # probabilty of changing omegas
 omega_switch = False    # if true, randomly change omegas
 omega_sample = True   # sample omegas instead
@@ -68,12 +68,26 @@ if __name__ == "__main__":
   np.random.seed(init_seed + 1000*chain_id)
   
   # run algorithm
-  outs = run_thermostats( problem, params, theta0, x0 )
-  #outs = run_sghmc( problem, params, theta0, x0 )
-  #outs = run_sgld( problem, params, theta0, x0 )
-  #outs = run_mcmc( problem, params, theta0, x0 )
+  np.random.seed(init_seed + 1000*chain_id)
+  sgnht = run_thermostats( problem, params, theta0, x0 )
   
+  np.random.seed(init_seed + 1000*chain_id)
+  sghmc = run_sghmc( problem, params, theta0, x0 )
+  
+  np.random.seed(init_seed + 1000*chain_id)
+  sgld = run_sgld( problem, params, theta0, x0 )
+  
+  np.random.seed(init_seed + 1000*chain_id)
+  mcmc = run_mcmc( problem, params, theta0, x0 )
+  
+  pp.figure(1)
+  pp.clf()
+  pp.plot( mcmc["THETA"][-1000:])
+  pp.plot( sgld["THETA"][-1000:])
+  pp.plot( sghmc["THETA"][-1000:])
+  pp.plot( sgnht["THETA"][-1000:])
+  pp.legend( ["MCMC","SGLD","SGHMC","SGNHT"])
   # view results of single chain
-  problem.view_single_chain( outs )
+  #problem.view_single_chain( sgnht )
     
   pp.show()

### 2015-02-26 code/working_code.py

@@ -5,6 +5,76 @@ import pylab as pp
 
 max_int_for_omega = 10**8
 
+def bin_errors_1d( bins, true_centered_probability, samples ):
+  N = len(samples)
+  cnts, bins = np.histogram( samples ,bins=bins)
+  
+  n = cnts.sum()
+  
+  probability = cnts / (float(N))
+  
+  missed = N - n
+  
+  double_error = np.sum(np.abs( true_centered_probability - probability )) + missed/float(N)
+  
+  #pdb.set_trace()
+  return double_error/2.0
+
+def cramer_vonMises_criterion( true_P, samples, bins ):
+  # see wiki cramer_von_Mises
+  N = len(samples)
+  cnts, bins = np.histogram( samples ,bins=bins)
+  
+  n = cnts.sum()
+  
+  est_P = cnts / (float(N))
+  
+  # for samples outside bins, we assume true_P is zeros
+  missed = N - n
+  
+  discrete_cvm_error = np.sum( ( true_P - est_P)**2) + (missed/float(N))**2
+  
+  #pdb.set_trace()
+  return discrete_cvm_error
+  
+def bin_sq_errors_1d( bins, true_centered_probability, samples ):
+  N = len(samples)
+  cnts, bins = np.histogram( samples ,bins=bins)
+  
+  n = cnts.sum()
+  
+  probability = cnts / (float(N))
+  
+  missed = N - n
+  
+  double_error = np.sqrt( np.sum( ( true_centered_probability - probability)**2) + (missed/float(N))**2 )
+  
+  #pdb.set_trace()
+  return double_error/2.0
+  
+
+        
+def convergence_to_scalar( psi, psi_estimate, times ):
+  nbrEstimate = len( psi_estimate )
+  
+  T     = len(times)
+  means = np.zeros( T )
+  vrs   = np.zeros( T )
+  for time_id,t in zip(range(T),times):
+    print time_id, t
+    n = t
+    if time_id <= nbrEstimate:
+      mean_estimate = psi_estimate[:t+1].mean()
+      var_estimate  = float(1.0/(t-1))*(psi - mean_estimate)**2
+    
+      means[time_id] = mean_estimate
+      vrs[time_id]   = var_estimate
+      
+  means = means[:time_id+1]
+  vrs = vrs[:time_id+1]  
+  used_times = times[:time_id+1]  
+  return used_times, means, vrs
+
 def bounce_off_boundaries( theta, p, lower_bounds, upper_bounds ):
   D = len(theta)
   
[…]

### 2015-02-26 papers/uai-2015/draft.tex
added figures, including variance of gradients

@@
 %\include{cdefs}
+
+% For algorithms
+\usepackage{algorithm}
+\usepackage{algorithmic}
+
+% As of 2011, we use the hyperref package to produce hyperlinks in the
+% resulting PDF.  If this breaks your system, please commend out the
+% following usepackage line and replace \usepackage{icml2014} with
+% \usepackage[nohyperref]{icml2014} above.
+%\usepackage{hyperref}
+% Packages hyperref and algorithmic misbehave sometimes.  We can fix
+% this with the following command.
+%\newcommand{\theHalgorithm}{\arabic{algorithm}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-In the application of stochastic gradient Hamiltonian dynamics to ABC, the stochasticity of the gradient due to batch sizes is instead caused by two other factors.
+The general approach of applying Hamiltonian dynamics to ABC requires choosing one of the SGHD algorithms and then plugging in the ABC gradient approximation $\nabla \hat{U}(\thetav)$.
+With this in mind we leave the details of the Hamiltonian updates to previous work \cite{welling2011bayesian,chen2014stochastic,ding2014bayesian} and focus on the details of how stochastic gradients are computed in the likelihood-free setting.
+
+\subsection{Deterministic Representations of Simulations}
+Implicit in each simulation run $\x \simsim \pi(\x | \thetav)$ is a sequence if internally generated random numbers that are used to produce random draws from $\pi(\x | \thetav)$.
+These random numbers are important to HABC because we wish to control the stochasticity of the simulator when computing its gradient.
+Furthermore, we will control the random numbers over multiple time steps.
+Instead of keeping track of random numbers, we can equivalently keep a vector of $S$ random seeds $\omegav$.
+This allows HABC to treat the simulation function $\pi(\x|\thetav)$ as a blackbox, outside of which we can control the random number generator (RNG), and represent $\x^{(s)}$ as the output of a deterministic function; i.e.
+$\x^{(s)} = f( \thetav, \omega_s)$ instead of $\x^{(s)} \simsim \pi(\x | \thetav)$.
+We include $\omegav$ as part of the state of our Markov chain.
+
+\subsection{Kernel-$\eps$ versus Synthetic-likelihood -based Gradients}
+In Section~\ref{sec:abc} we showed that the synthetic-likelihood representation of $\lleps(\thetav)$ is less sensitive to small choices of $\epsvec$.
+This is particularly important to HABC as our gradient approximations are proportional to differences in $\lleps(\thetav)$; if the variance of the stochastic gradients is too high, then we must choose a very small step-size $\eta$, eliminating the usefulness of HMC for ABC.
+Under the deterministic representation of $\x^{(s)}$, we can write the loglikelihoods
+\begin{eqnarray}
+ \lleps(\thetav) & \propto & \log \sum_s \mathcal{N}( \y | f( \thetav, \omega_s), \eps^2 )\\
+                 & \approx & -\log \eps -\frac{1}{2 \eps^2} ( \y - f( \thetav, \omega_m) )^2 
+\end{eqnarray}
+In the second line we have again assumes $\eps$ is very small and $m$ is the index of the random seed producing the closest simulation to $\y$.
+For a finite difference approximation, $\partial \lleps(\thetav)/\partial\thetav$ is
+\begin{equation}
+ \frac{1}{4\dtheta \eps^2} \lp ( \y - f( \thetav-\dtheta, \omega_m^-) )^2 -( \y - f( \thetav + \dtheta, \omega_m^+) )^2 \rp
+\end{equation}
+
+On the other hand, the synthetic-likelihood is stable; using deterministic representation, 
+\begin{equation}
+\mu_{\thetav} = \frac{1}{S} \sum_s f( \thetav, \omega_s ) ~~~~~~~~~~~~~ \sigma^s_{\thetav} = \frac{1}{S-1} \sum_s ( \mu_{\thetav} - f( \thetav, \omega_s ) )^2
+\end{equation}
+the gradients (for a 1-dim problem) use $\eps$ as a smoothness prior in $\partial \lleps(\thetav)/\partial\thetav$:
+\begin{equation}
+  -\frac{1}{2}\log\lp\frac{\sigma_{\theta^+}^2 + \eps^2}{\sigma_{\theta^-}^2 + \eps^2}\rp -\frac{( \y - \mu_{\thetav^+} )^2 }{2 (\sigma^2_{\thetav^+} + \eps^2 )} +\frac{( \y - \mu_{\thetav^-} )^2 }{2 (\sigma^2_{\thetav^-} + \eps^2 )}
+\end{equation}
+
+\begin{figure}[t]
+\vskip 0.2in
+\begin{center}
+\includegraphics[width=0.95\columnwidth]{./images/exp_varg_figure.pdf}
+\caption{\small{Variance of gradient estimation using kernel-$\eps$ and SL for different values of $S\in\{5,50\}$ and fixed $\eps = 0.37$ (the same used in the other results).
+For kernel-$\eps$, $\varv(\thetav_{\text{MAP}}) = 6.7 \pm 131.5$ and $\varv(\thetav_{\text{MAP}}) = 6.9 \pm 18.9$ (for $S=5$ and $S=50$, respectively).
+For SL,$\varv(\thetav_{\text{MAP}}) = 6.35 \pm 42.4$ and $\varv(\thetav_{\text{MAP}}) = 0.62 \pm 5.2$ (for $S=5$ and $S=50$, respectively).
+The ``true'' ABC gradient at $\thetav_{\text{MAP}}$ is $0.26$.
+The ``true'' ABC loglikelihood is based on a Gaussian approximation (smoothed by $\eps$), with mean and variance set by moment matching with the true Gamma distribution (the sum of $N$ exponentials).
+Not only does kernel-$\eps$ keep a high variance for even large $S$, it seems to have a bias in its gradient estimate.
+[look into this?] }}
+\label{fig:exp-varg}
+\end{center}
+\vskip -0.2in
+\end{figure}
+
+\subsection{From Finite Differences to Simultaneous Perturbations}
+\input{algo_fdsa_abc}
+Algorithm~\ref{algo:fdsa} shows the {\em finite difference stochastic approximation} (FDSA) \cite{kiefer1952stochastic} to $\nabla U(\thetav)$ as a function of random seeds $\omegav$.
+Note we have deliberately shown the deterministic simulations ($f$) outside of $\lleps$ to emphasize its dependence on $\x$.
+The number of simulations required for FDSA is $2 S D$, which may be acceptable for some small ABC problems.
+Our goal is to scale ABC to high-dimensions and for that we need an alternative stochastic approximation of $\nabla U(\thetav)$.
+
+In the gradient-free setting, Spall \cite{spall1992multivariate, spall2000adaptive} provides a stochastic approximate to the true gradient using only $2$ forward simulations for any dimension $D$ (though the approximation can be improved by averaging $R$ estimates).
+Spall's {\em simultaneous perturbation stochastic approximation} (SPSA) algorithm works as follows.
+Let $L$ be the gradient-free function we wish to optimize.
+Each approximation randomly generates a {\em perturbation mask} (our name) $\Deltav$ of dimension $D=|\thetav|$ where entry $\deltad \sim \text{Bernouilli}(1/2)$ (actually, values $\pm 1$, not $0/1$).
+Then $L$ is evaluated at $\thetav+\dtheta \Deltav$ and $\thetav-\dtheta \Deltav$, giving the gradient approximation $\gradvest(\thetav) \approx \partial L(\thetav) / \partial \thetav$:
+\begin{equation}
+  \gradvest(\thetav) = \frac{L\lp\thetav+\dtheta \Deltav\rp - L\lp\thetav-\dtheta \Deltav\rp}{2 \dtheta} \begin{bmatrix} 
+                                   1/\Delta_1 \\
+                                   1/\Delta_2 \\
+                                   \vdots \\
+                                   1/\Delta_D \\
+                                \end{bmatrix}
+\end{equation}
+If we let $\gradvest_r(\thetav)$ be the estimate using perturbation mask $\Deltav_r$, the estimate $\gradvest(\thetav)$ can be improved by averaging $\gradvest(\thetav) = 1/R \sum_r \gradvest_r(\thetav)$.
+Algorithm~\ref{algo:spsa} shows SPSA to estimate $\nabla U(\thetav)$.  The number of simulations required for SPSA is $2 S R$, where $R \geq 1$.
+\input{algo_spsa_abc}
+
+Variations of SPSA include {\em one-sided} SPSA \cite{spall2000adaptive} (we use what Spall calls 2SPSA) and an algorithm for estimating the Hessian based on the same principal as SPSA \cite{spall2005monte}.
+The one-sided version is attractive computationally, but for HABC, the updates for $\thetav$ require simulating two-sides anyway, so using 2SPSA makes more sense.
+SPSA has also been used within a procedure for maximum-likelihood estimation for hidden Markov models using ABC \cite{Ehrlich2013}.
+
+\subsection{Sticky Random Numbers}
+An obvious use of common random numbers, as previously demonstrated for SPSA \cite{kleinman1999simulation}, is within the gradient approximation so that the noise from the simulator has less of an influence that $\dtheta$.
+In addition to this use, we have found that using the same random seeds over multiple time-steps improves the performance of HABC (TBD).
+This is very similar to using dependent random streams to drive MCMC \cite{Murray2012,Neal2012}, the main difference we believe is that we are using the Hamiltonian dynamics to drive proposal in $\thetav$ and using $\omegav$ to give persistent simulations, reducing the variance from the unknown simulator noise function.
+
+Our updates for $\omegav$ are simple.
+At each time step $t$, each seed is refreshed to a new random value with probability $\gamma$, i.e.
+$\omega_s \leftarrow \omega_s$ with probability $\gamma$, and $\omega_s \leftarrow \mathcal{U}(1,W)$, where $W=10^8$ in our algorithms.
+By observing the simulation at different $\thetav$ but using same seeds, we can eliminate much of the simulator noise from the Markov chain.
+By treating $\omegav$ as part of the state of the Markov chain, we can smoothly move along noiseless outputs for several iterations and then propose new seeds.
+Though proposing new seeds produces a valid Metropolis-Hastings algorithm (keeping $\thetav$ fixed and changing $\omegav$), we found that using {\em sticky} random seeds works well.
+These seeds can be precomputed before the chain is run and can be seen as apart from the state of the Markov chain.
+[CONFUSING]
+
+[We may want anti-thetic instead of CRNs, ie overrelax versus underrelax, etc.]
+% [CITE NEAL How to view an MCMC...]
+% Another variance reduction technique called the method of {\em common random numbers} (CRN) \cite{kleinman1999simulation} sets a common seed for the random number generator (RNG) for both calls to the simulator.
+This technique can remove the effect of the simulator noise in the gradient estimate, leaving on the randomness of the perturbation masks as the source of noise.
+Consider using SPSA instead of SGD: the CRN technique is equivalent of using the same batch of data vectors to evaluate the log-likelihood which is a sensible approach.
+%
+%
+% [By observing the simulation at different theta but same seeds, we can eliminate much of the noise in the markov chain.
+By treating omega as part of the state, we can smoothly move along noiseless outputs for several iterations and then step.]
+
+% In the application of stochastic gradient Hamiltonian dynamics to ABC, the stochasticity of the gradient due to batch sizes is instead caused by two other factors.
 The first factor, is due to the finite number of forward simulations that are used to approximate the log-likelihood function.
@@
 
-To begin, we define $\thetavplusd = \thetav + c \Deltavd$ and $\thetavminusd = \thetav + c \Deltavd$, where $c$ is a small positive scalar, e.g.
+% To begin, we define $\thetavplusd = \thetav + c \Deltavd$ and $\thetavminusd = \thetav + c \Deltavd$, where $c$ is a small positive scalar, e.g.
 $c=0.001$, and $\Deltavd$ is a vector of size $D$ that, for FD estimates is $1$ for entry $d$ and $0$ otherwise.
@@
 These are then plugged into the log-likelihood estimate for $\y$, which we call $\mathcal{L}$:
-\begin{eqnarray}
-  \frac{\partial \mathcal{L}(\thetav)}{\partial \thetavd} & \approx & \frac{\mathcal{L}(\thetavplusd)- \mathcal{L}(\thetavminusd)}{2c}
-\end{eqnarray}
+% \begin{eqnarray}
+%   \frac{\partial \mathcal{L}(\thetav)}{\partial \thetavd} & \approx & \frac{\mathcal{L}(\thetavplusd)- \mathcal{L}(\thetavminusd)}{2c}
+% \end{eqnarray}
 
-For the $\epsilon$-kernel density model
-% ,
-% \begin{eqnarray}
-%   \mathcal{L}(\thetavpmd) & = & \log \sum_{s=1}^2 \mathcal{N}( \y | f( \thetavpmd, \omega_s ), \eps^2 ) \\
-%                           & = & \log \mathcal{N}( \y | f( \thetavpmd, \omega_{m} ), \eps^2 ) + \log\lp 1 + H_d \rp \\
-%                           & \approx & \log \mathcal{N}( \y | f( \thetavpmd, \omega_{m} ), \eps^2 ) + 0
-% \end{eqnarray}
-% where we replaced the random variable $\x$ with a deterministic function $f( \thetavpmd, \omega_s )$ (i.e.
+% For the $\epsilon$-kernel density model
+% % ,
+% % \begin{eqnarray}
+% %   \mathcal{L}(\thetavpmd) & = & \log \sum_{s=1}^2 \mathcal{N}( \y | f( \thetavpmd, \omega_s ), \eps^2 ) \\
+% %                           & = & \log \mathcal{N}( \y | f( \thetavpmd, \omega_{m} ), \eps^2 ) + \log\lp 1 + H_d \rp \\
+% %                           & \approx & \log \mathcal{N}( \y | f( \thetavpmd, \omega_{m} ), \eps^2 ) + 0
+% % \end{eqnarray}
+% % where we replaced the random variable $\x$ with a deterministic function $f( \thetavpmd, \omega_s )$ (i.e.
 the simulator run at $\thetavpmd$ using the random state $\omega_s$, such that $\xpmd = f( \thetavpmd, \omega_s )$).
@@
 is closest to $\y$) and used $H_d$ (sum of terms $<0$ and $\ll 0$ for small $\eps$), and in the last line simplified further by assuming small $\eps$, so that the log-likelihood is dominated by a single simulation closest to $\y$.
-it is easy to show that the gradient estimate is sensitive to $\epsilon$ in this case: 
-\begin{eqnarray}
-  \frac{\partial \mathcal{L}(\thetav)}{\partial \thetavd} & \approx & \frac{1}{2c\eps^2}\lp -\frac{1}{2}(\y-\xplussd)^2 + \frac{1}{2}(\y-\xminussd)^2  \rp
-\end{eqnarray}
-So by making $\epsilon$ arbitrarily small we can make the gradient extremely sensitive to our simulations.
+% it is easy to show that the gradient estimate is sensitive to $\epsilon$ in this case:
+% \begin{eqnarray}
+%   \frac{\partial \mathcal{L}(\thetav)}{\partial \thetavd} & \approx & \frac{1}{2c\eps^2}\lp -\frac{1}{2}(\y-\xplussd)^2 + \frac{1}{2}(\y-\xminussd)^2  \rp
+% \end{eqnarray}
+% So by making $\epsilon$ arbitrarily small we can make the gradient extremely sensitive to our simulations.
 Intuitively, we want the gradient to be fairly stable no matter what values of $\epsilon$ we use.
@@
 % \end{eqnarray}
-We can see that $\epsilon$ now can be set very small as it now acts as a smoother for our estimate, resulting in a robust gradient.
+% We can see that $\epsilon$ now can be set very small as it now acts as a smoother for our estimate, resulting in a robust gradient.
 
-\subsection{Gradients in High Dimensions}
+%\subsection{Gradients in High Dimensions}
 
 %\subsection{Spall's method: SPSA}
-In the gradient-free setting, Spall \cite{spall1992multivariate} provides a stochastic approximate to the true gradient using only 2 forward simulations (function evaluations).
+% In the gradient-free setting, Spall \cite{spall1992multivariate} provides a stochastic approximate to the true gradient using only 2 forward simulations (function evaluations).
 This is in contrast to multivariate finite-difference stochastic approximation FDSA \cite{kiefer1952stochastic} requiring $2p$ evaluations.
+%
+% The gradient estimate is
+% \begin{equation}
+%   \gradestimate{t}{\thetav_t} = \begin{bmatrix}
+%                                    \frac{\feval{t}{+} - \feval{t}{-}}{2 \cstep{t}{1} \dstep{t}{1}} \\
+%                                    \vdots \\
+%                                    \frac{\feval{t}{+} - \feval{t}{-}}{2 \cstep{t}{p} \dstep{t}{p}} \\
+%                                 \end{bmatrix}
+% \end{equation}
+% where $\cstep{t}{p}$ is a step-size that is usually constant for all dimensions $p$, but can be different (as shown in this case); $\dstep{t}{p} \in {-1, +1}$ is a {\em perturbation mask} (called symmetric Bernouilli variables by Spall), i.e.
+$\dstep{t}{p} \sim 2*\text{Bernouilli}(0.5) - 1$; and  $\feval{t}{\pm}$ are function evaluations:
+% \begin{eqnarray}
+%   \feval{t}{+} &= \loglik{\thetav + \cstepconst{t}\dstepconst{t}} \\
+%   \feval{t}{-} &= \loglik{\thetav - \cstepconst{t}\dstepconst{t}}
+% \end{eqnarray}
+% A way of reducing the noise in the gradient is by averaging over $q$ draws of $\dstepconst{t}$:
+% \begin{equation}
+%   \gradestimate{t}{\thetav_t} = \frac{1}{q} \sum_{j=1}^q \gradestimaterepeat{t}{\thetav_t}{j}
+% \end{equation}
+% where for each $j$ new perturbation masks are drawn.
+%
 
-The gradient estimate is
-\begin{equation}
-  \gradestimate{t}{\thetav_t} = \begin{bmatrix} 
-                                   \frac{\feval{t}{+} - \feval{t}{-}}{2 \cstep{t}{1} \dstep{t}{1}} \\
-                                   \vdots \\
-                                   \frac{\feval{t}{+} - \feval{t}{-}}{2 \cstep{t}{p} \dstep{t}{p}} \\
-                                \end{bmatrix}
-\end{equation}
-where $\cstep{t}{p}$ is a step-size that is usually constant for all dimensions $p$, but can be different (as shown in this case); $\dstep{t}{p} \in {-1, +1}$ is a {\em perturbation mask} (called symmetric Bernouilli variables by Spall), i.e.
-$\dstep{t}{p} \sim 2*\text{Bernouilli}(0.5) - 1$; and  $\feval{t}{\pm}$ are function evaluations:
-\begin{eqnarray}
-  \feval{t}{+} &= \loglik{\thetav + \cstepconst{t}\dstepconst{t}} \\
-  \feval{t}{-} &= \loglik{\thetav - \cstepconst{t}\dstepconst{t}}
-\end{eqnarray}
-A way of reducing the noise in the gradient is by averaging over $q$ draws of $\dstepconst{t}$:
-\begin{equation}
-  \gradestimate{t}{\thetav_t} = \frac{1}{q} \sum_{j=1}^q \gradestimaterepeat{t}{\thetav_t}{j}
-\end{equation}
-where for each $j$ new perturbation masks are drawn.
-
-For ABC, the gradients we compute are stochastic because we must approximate $\pi(\x | \thetav)$ with forward simulations.
-We can make the approximation arbitrarily good by running more simulations, but this is often unnecessary for sampling, because we can make use of the uncertainty in our current approximation to decide if our sample is correct or not (cite UAI2014?).
-
-\begin{itemize}
-  \item In both cases, we rewrite the expression conditioning on a set of random seeds omega.
-Explain the reason for this.
-By observing the simulation at different theta but same seeds, we can eliminate much of the noise in the markov chain.
-By treating omega as part of the state, we can smoothly move along noiseless outputs for several iterations and then step.
-\end{itemize}
-
-\subsection{Stochastic ABC Gradients}
-For low-dimensional problems a finite difference (FD) approximation to $\nabla U$.
+% \subsection{Stochastic ABC Gradients}
+% For low-dimensional problems a finite difference (FD) approximation to $\nabla U$.
 For each dimension $p$, forward simulate for each random seed twice at plus $c$ and minus $c$ (only along dimension $p$): $x_{\pm}^{(s)} \sim \pi( \x | \thetav \pm c \Delta_p, \omega_s)$.
@@
 $\hat{U}_{K_{\eps}}$:
-\begin{eqnarray}
-  - \log \pi(\thetav \pm c \Delta_p) - \log \sum_s \pi_{\eps}(\y | f( \thetav \pm c \Delta_p, \omega_s ) )
-\end{eqnarray}   
-or $\hat{U}_{SL}$
-\begin{eqnarray}
-  - \log \pi(\thetav \pm c \Delta_p) - \log \mathcal{N}(\y | \mu_{\thetav \pm c \Delta_p}, \sigma^2_{\thetav \pm c \Delta_p} + \eps^2 )
-\end{eqnarray}
+% \begin{eqnarray}
+%   - \log \pi(\thetav \pm c \Delta_p) - \log \sum_s \pi_{\eps}(\y | f( \thetav \pm c \Delta_p, \omega_s ) )
+% \end{eqnarray}
+% or $\hat{U}_{SL}$
+% \begin{eqnarray}
+%   - \log \pi(\thetav \pm c \Delta_p) - \log \mathcal{N}(\y | \mu_{\thetav \pm c \Delta_p}, \sigma^2_{\thetav \pm c \Delta_p} + \eps^2 )
+% \end{eqnarray}
+%
+%
+% \begin{eqnarray}
+%   \hat{U}(\thetav)       & = & - \log \pi(\thetav) - \log \pi_{\eps}(\y | \thetav ) \\
+%   \nabla\hat{U}(\thetav) & = & - \nabla \log \pi(\thetav) - \frac{N}{n} \sum_{i=h_1}^{h_n} \nabla \log \pi(\yi | \thetav )
+% \end{eqnarray}
 
-\begin{eqnarray}
-  \hat{U}(\thetav)       & = & - \log \pi(\thetav) - \log \pi_{\eps}(\y | \thetav ) \\
-  \nabla\hat{U}(\thetav) & = & - \nabla \log \pi(\thetav) - \frac{N}{n} \sum_{i=h_1}^{h_n} \nabla \log \pi(\yi | \thetav ) 
-\end{eqnarray}
+%\subsection{Common Random Numbers}
 
-\subsection{Common Random Numbers}
+% Using the same parameter vector $\thetav$ but different random seeds:
+% \begin{eqnarray}
+%   \pi\lp \y | \thetav \rp & = &  \int \pi_{\epsvec}\lp \y | \x \rp \pi\lp \x | \thetav \rp d \x \\
+%    & \approx & \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}\lp \y | \x^{(s)} \rp
+% \end{eqnarray}
+% where $\x^{(s)} = \pi\lp \x | \thetav, \omega_s \rp$.
 
-[CITE NEAL How to view an MCMC...]
-Another variance reduction technique called the method of {\em common random numbers} (CRN) \cite{kleinman1999simulation} sets a common seed for the random number generator (RNG) for both calls to the simulator.
-This technique can remove the effect of the simulator noise in the gradient estimate, leaving on the randomness of the perturbation masks as the source of noise.
-Consider using SPSA instead of SGD: the CRN technique is equivalent of using the same batch of data vectors to evaluate the log-likelihood which is a sensible approach.
-
-Using the same parameter vector $\thetav$ but different random seeds:
-\begin{eqnarray}
-  \pi\lp \y | \thetav \rp & = &  \int \pi_{\epsvec}\lp \y | \x \rp \pi\lp \x | \thetav \rp d \x \\
-   & \approx & \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}\lp \y | \x^{(s)} \rp
-\end{eqnarray}
-where $\x^{(s)} = \pi\lp \x | \thetav, \omega_s \rp$.
-Note we have passed a unique random seed $\omega_s$ into the simulator, making $\x^{(s)}$ a deterministic function of the simulator with parameters $\{\thetav, \omega_s\}$.
+% Note we have passed a unique random seed $\omega_s$ into the simulator, making $\x^{(s)}$ a deterministic function of the simulator with parameters $\{\thetav, \omega_s\}$.
 It will be useful to rewrite this deterministic simulator function as $f$: $\x^{(s)} = f\lp \thetav, \omega_s \rp$, allowing us to rewrite the likelihood as
-\begin{eqnarray}
-  \pi\lp \y | \thetav \rp & \approx & \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}\lp \y | f\lp \thetav, \omega_s \rp \rp
-\end{eqnarray}
+% \begin{eqnarray}
+%   \pi\lp \y | \thetav \rp & \approx & \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}\lp \y | f\lp \thetav, \omega_s \rp \rp
+% \end{eqnarray}
 
@@
 
-\subsection{Algorithms}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 The inference problem is therefore to sample from $p( \theta | y, \alpha, \beta, N )$.
+
+\begin{figure}[t]
+\vskip 0.2in
+\begin{center}
+\includegraphics[width=0.95\columnwidth]{./images/exp_crn_figure.pdf}
+\caption{\small{A view of a simulator in terms of common random numbers.
+The horizontal line represents $\y$ and red shading $\pm 2\eps$.
+The shaded curved region represents $2\sigma$ of $\pi(\x|\thetav)$.
+The dashed lines are $f(\thetav, \omega_s)$ for several values of $\omega$.
+The blue circles are potential random samples from $\pi(\x|\thetav)$.
+For a fixed value $\omega_s$, the simulator produces deterministic outputs that change smoothly, even though the simulator itself is quite noisy.}}
+\label{fig:exp-crns}
+\end{center}
+\vskip -0.2in
+\end{figure}
+
+\begin{figure*}[t]
+\vskip 0.2in
+\setlength{\linewidth}{\textwidth}
+\setlength{\hsize}{\textwidth}
+\begin{center}
+\includegraphics[width=0.5\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
+\includegraphics[width=1.2\columnwidth]{./images/exp-ABC-MCMC-theta-timeseries.pdf}
+\includegraphics[width=0.5\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
+\includegraphics[width=1.2\columnwidth]{./images/exp-SG-Langevin-theta-timeseries.pdf}
+\includegraphics[width=0.5\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
+\includegraphics[width=1.2\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries.pdf}
+\caption{\small{}}
+\label{fig:exp-posteriors}
+\end{center}
+\vskip -0.2in
+\end{figure*}
 
@@
   \item Key: opens the door to high-dimensional ABC inference.
+  \item [Test for more accurate gradients] For ABC, the gradients we compute are stochastic because we must approximate $\pi(\x | \thetav)$ with forward simulations.
+We can make the approximation arbitrarily good by running more simulations, but this is often unnecessary for sampling, because we can make use of the uncertainty in our current approximation to decide if our sample is correct or not (cite UAI2014?).
 \end{itemize}

### 2015-02-26 code/exponential_problem/exp_wrapper.py

@@ -7,12 +7,14 @@ import pylab as pp
 
 problem_params = default_params()
 
-problem_params["N"] = 50
-problem_params["q_stddev"] = 0.01
-problem_params["theta_star"]      = 0.2
-problem_params["epsilon"] = 0.1*np.sqrt( 1.0 / (problem_params["N"]*problem_params["theta_star"]**2) )
-problem_params["alpha"]           = 1.0
-problem_params["beta"]            = 1.0
+problem_params["N"]               = 20
+problem_params["q_stddev"]        = 0.2
+problem_params["theta_star"]      = 0.15
+problem_params["epsilon"]         = 0.25*np.sqrt( 1.0 / (problem_params["N"]*problem_params["theta_star"]**2) )
+problem_params["alpha"]           = 0.1
+problem_params["beta"]            = 0.1
+problem_params["min_range"]       = 0.005
+problem_params["max_range"]       = 0.3
 #problem_params["alpha"]           = 3.75
 #problem_params["beta"]            = 1.01
 
@@ -70,7 +72,7 @@ class generate_exponential( object ):
       if seed is not None:
         # save current state
         current_state = np.random.get_state()
-        np.random.seed(seed+s)
+        np.random.seed(seed[s])
     
     
       raw_outputs = self.p.simulation_function( theta )
@@ -162,17 +164,17 @@ class generate_exponential( object ):
       theta_plus = theta_minus + 2*d_theta
       
       if omega is not None:
-        seed = omega+s
+        seed = omega[s]
        
       if seed is None:
         state = np.random.get_state()
       
-      X_plus[s] = self.simulate( theta_plus, seed=seed )[0]
+      X_plus[s] = self.simulate( theta_plus, seed=[seed] )[0]
       
       if seed is None:
         np.random.set_state(state)
         
-      X_minus[s] = self.simulate( theta_minus, seed=seed )[0]
+      X_minus[s] = self.simulate( theta_minus, seed=[seed] )[0]
       
       seeds.append(seed)
       
@@ -190,7 +192,7 @@ class generate_exponential( object ):
     
     grad = (f_plus-f_minus)/(theta_plus-theta_minus) + (self.p.alpha-1)/theta - self.p.beta
     
[…]

### 2015-02-26 code/exponential_problem/figs_run_and_view_crns.py

@@ -1,0 +1,160 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+from scipy import stats as spstats
+from exp_wrapper import *
+from abc_sgld.code.working_code import *
+
+keep_x        = True 
+init_seed     = 1
+T             = 10000 # nbr of samples
+verbose_rate  = 1000
+C             = 1.01    # injected noise variance parameter
+eta           = 0.01 # step size for Hamiltoniam dynamics
+#h = 0.0005
+
+# params for gradients
+d_theta = 0.01  # step size for gradient estimate
+S       = 5 
+grad_params = {}
+  
+# keep theta within bounds
+lower_bounds = np.array([0.001])
+upper_bounds = np.array([np.inf])
+
+
+# ----------------------- #
+# common ransom seeds     #
+# ----------------------- #
+use_omega    = False    # use fixed random seeds for simulations
+omega_rate   = 1.1    # probabilty of changing omegas
+omega_switch = False    # if true, randomly change omegas
+omega_sample = True   # sample omegas instead
+assert (omega_switch * omega_sample) ==0, "only get to do one type of omega update"
+omega_params = {"use_omega":use_omega, \
+                "omega_rate":omega_rate, \
+                "omega_switch":omega_switch,\
+                "omega_sample":omega_sample}  
+
+
+def make_figure(  problem, seeds, theta_range ):
+  figsize = (12,8)
+
+  pp.rc('text', usetex=True)
+  pp.rc('font', family='times')
+  f = pp.figure( figsize=figsize )
+  sp=pp.subplot(1,1,1)
+      
+  theta0=theta_range[0]
+  thetaN=theta_range[-1]
+  
+  #pp.plot( THETA[-5000:], X[-5000:], 'b.-', alpha=0.5 )
+  pp.hlines( problem.p.obs_statistics, theta0, thetaN,lw=4,alpha=0.5)
+  
+  #pp.fill_between( [theta0, thetaN], problem.y-problem.p.epsilon, problem.y+problem.p.epsilon, alpha=0.5,color='r')
+  pp.fill_between( [theta0, thetaN], problem.y-2*problem.p.epsilon, problem.y+2*problem.p.epsilon, alpha=0.25,color='r')
+
+  mn = 1.0/theta_range
[…]

### 2015-02-26 code/exponential_problem/figs_run_gradient_distributions.py

@@ -1,0 +1,184 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+from scipy import stats as spstats
+from exp_wrapper import *
+from abc_sgld.code.working_code import *
+
+keep_x        = True 
+init_seed     = 1
+T             = 10000 # nbr of samples
+verbose_rate  = 1000
+C             = 1.01    # injected noise variance parameter
+eta           = 0.01 # step size for Hamiltoniam dynamics
+#h = 0.0005
+
+# params for gradients
+d_theta = 0.01  # step size for gradient estimate
+S       = 5 
+grad_params = {}
+  
+# keep theta within bounds
+lower_bounds = np.array([0.001])
+upper_bounds = np.array([np.inf])
+
+
+# ----------------------- #
+# common ransom seeds     #
+# ----------------------- #
+use_omega    = False    # use fixed random seeds for simulations
+omega_rate   = 1.1    # probabilty of changing omegas
+omega_switch = False    # if true, randomly change omegas
+omega_sample = True   # sample omegas instead
+assert (omega_switch * omega_sample) ==0, "only get to do one type of omega update"
+omega_params = {"use_omega":use_omega, \
+                "omega_rate":omega_rate, \
+                "omega_switch":omega_switch,\
+                "omega_sample":omega_sample}  
+
+
+def make_figure(  problem, seeds, theta_range ):
+  figsize = (12,8)
+
+  pp.rc('text', usetex=True)
+  pp.rc('font', family='times')
+  f = pp.figure( figsize=figsize )
+  sp=pp.subplot(1,1,1)
+      
+  theta0=theta_range[0]
+  thetaN=theta_range[-1]
+  
+  #pp.plot( THETA[-5000:], X[-5000:], 'b.-', alpha=0.5 )
+  pp.hlines( problem.p.obs_statistics, theta0, thetaN,lw=4,alpha=0.5)
+  
+  #pp.fill_between( [theta0, thetaN], problem.y-problem.p.epsilon, problem.y+problem.p.epsilon, alpha=0.5,color='r')
[…]

### 2015-02-26 code/exponential_problem/figs_runs_posteriors.py

@@ -1,0 +1,190 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+from scipy import stats as spstats
+from exp_wrapper import *
+from abc_sgld.code.working_code import *
+
+def view_theta_timeseries( problem, theta_range, samples, algoname, howmany = 1000 ):
+  figsize = (16,6)
+  alpha=0.75
+  pp.rc('text', usetex=True)
+  pp.rc('font', family='times')
+  f = pp.figure( figsize=figsize )
+  sp=pp.subplot(1,1,1)
+      
+  theta0=theta_range[0]
+  thetaN=theta_range[-1]
+  
+  #pp.hist( samples, problem.p.nbins_coarse, range=problem.p.range,normed = True, alpha = alpha )
+  pp.plot( samples[-howmany:], 'k-', lw=2, alpha=0.5)
+  pp.plot( samples[-howmany:], 'ko')
+  
+  pp.ylim(theta0,thetaN)
+  #pp.xlabel( "theta")
+  #pp.ylabel( "x")
+  
+  pp.xlabel( r'time' )
+  pp.ylabel( r'${\bm \theta}$', rotation='horizontal' )
+  
+  pp.legend( [algoname], loc=1,fancybox=True,prop={'size':16} )
+  
+    
+  #pp.title( r"Simulation with Common Random Numbers", fontsize=22) 
+  set_tick_fonsize( sp, 16 )
+  set_title_fonsize( sp, 32 )
+  set_label_fonsize( sp, 24 )
+
+def view_posterior( problem, theta_range, samples, algoname, burnin = 1000 ):
+  figsize = (6,6)
+  alpha=0.75
+  pp.rc('text', usetex=True)
+  pp.rc('font', family='times')
+  f = pp.figure( figsize=figsize )
+  sp=pp.subplot(1,1,1)
+      
+  theta0=theta_range[0]
+  thetaN=theta_range[-1]
+  
+  pp.hist( samples[1000:], problem.p.nbins_coarse, range=problem.p.range,normed = True, alpha = alpha )
+  pp.plot( theta_range, np.exp( problem.loglike_posterior(theta_range)), 'k--', lw=3)
+  
+  pp.xlim(theta0,thetaN)
+  #pp.xlabel( "theta")
+  #pp.ylabel( "x")
+  
+  pp.ylabel( r'$\pi( {\bm \theta } | {\bf y} )$', rotation='vertical' )
[…]

### 2015-02-26 code/exponential_problem/figs_view_gradient_disitributions.py

@@ -1,0 +1,185 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+from scipy import stats as spstats
+from exp_wrapper import *
+from abc_sgld.code.working_code import *
+
+keep_x        = True 
+init_seed     = 1
+T             = 10000 # nbr of samples
+verbose_rate  = 1000
+C             = 1.01    # injected noise variance parameter
+eta           = 0.01 # step size for Hamiltoniam dynamics
+#h = 0.0005
+
+# params for gradients
+d_theta = 0.01  # step size for gradient estimate
+S       = 5 
+grad_params = {}
+  
+# keep theta within bounds
+lower_bounds = np.array([0.001])
+upper_bounds = np.array([np.inf])
+
+
+# ----------------------- #
+# common ransom seeds     #
+# ----------------------- #
+use_omega    = False    # use fixed random seeds for simulations
+omega_rate   = 1.1    # probabilty of changing omegas
+omega_switch = False    # if true, randomly change omegas
+omega_sample = True   # sample omegas instead
+assert (omega_switch * omega_sample) ==0, "only get to do one type of omega update"
+omega_params = {"use_omega":use_omega, \
+                "omega_rate":omega_rate, \
+                "omega_switch":omega_switch,\
+                "omega_sample":omega_sample}  
+
+
+def view_grads(G,names,lines):
+  Ss = G.keys()
+  xr = np.linspace(-100,100,1000)
+  pp.rc('text', usetex=True)
+  pp.rc('font', family='times')
+  f = pp.figure( figsize=(12,8) )
+  lgs = []
+  sp=pp.subplot(1,1,1)
+  for line,S in zip(lines,Ss):
+    g=G[S]
+    g1 = spstats.kde.gaussian_kde(g[:,1])
+    g2 = spstats.kde.gaussian_kde(g[:,2])
+    
+    #pp.hist( g[:,1], 50, normed=True, histtype='step', color="b",linewidth=2)
+    #pp.hist( g[:,2], 50, normed=True, histtype='step', color="g",linewidth=2)
+    
[…]

### 2015-02-26 code/exponential_problem/make_illustrations.py

@@ -1,93 +1,0 @@
-import numpy as np
-import pylab as pp
-import scipy as sp
-from scipy import stats as spstats
-from exp_wrapper import *
-from abc_sgld.code.working_code import *
-
-keep_x        = True 
-init_seed     = 1
-T             = 1000 # nbr of samples
-verbose_rate  = 1000
-C             = 1.01    # injected noise variance parameter
-eta           = 0.01 # step size for Hamiltoniam dynamics
-#h = 0.0005
-
-# params for gradients
-d_theta = 0.01  # step size for gradient estimate
-S       = 5 
-grad_params = {}
-  
-# keep theta within bounds
-lower_bounds = np.array([0.001])
-upper_bounds = np.array([np.inf])
-
-
-# ----------------------- #
-# common ransom seeds     #
-# ----------------------- #
-use_omega    = False    # use fixed random seeds for simulations
-omega_rate   = 1.1    # probabilty of changing omegas
-omega_switch = False    # if true, randomly change omegas
-omega_sample = True   # sample omegas instead
-assert (omega_switch * omega_sample) ==0, "only get to do one type of omega update"
-omega_params = {"use_omega":use_omega, \
-                "omega_rate":omega_rate, \
-                "omega_switch":omega_switch,\
-                "omega_sample":omega_sample}  
-
-if __name__ == "__main__": 
-  pp.close('all')
-  
-  chain_id = 1
-  params = {}
-  params["chain_id"]  = chain_id
-  params["init_seed"] = init_seed
-  params["T"]       = T
-  params["S"]       = S
-  params["d_theta"] = d_theta
-  params["eta"]     = eta
-  params["C"]       = C
-  params["omega_params"] = omega_params
-  params["verbose_rate"] = verbose_rate
-  params["grad_params"]  = {"logs":{"true":[],"true_abc":[],"2side_keps":[],"2side_sl":[]},\
-                            "record_2side_sl_grad":False, "record_2side_keps_grad":False,"record_true_abc_grad":True,"record_true_grad":False}
-  params["lower_bounds"] = lower_bounds
-  params["upper_bounds"] = upper_bounds
-  params["keep_x"]       = keep_x
-  
-  theta0 = np.array( [0.2])
-  x0     = None
-  
-  # initialize the simulator
[…]

### 2015-02-26 code/exponential_problem/temp_scripts.py

@@ -1,0 +1,143 @@
+from abcpy.problems.blowfly.blowfly    import BlowflyProblem   as Problem
+from abcpy.problems.blowfly.blowfly    import default_params   as load_default_params
+from abcpy.plotting import *
+
+import pdb
+import numpy as np
+import pylab as pp
+
+# exponential distributed observations with Gamma(alpha,beta) prior over lambda
+problem_params = load_default_params()
+problem_params["q_factor"] = 0.1
+problem = Problem( problem_params, force_init = True )
+
+if __name__ == "__main__":
+  figsize = (16,12)
+  #dpi = 800
+  
+  
+  
+  
+  params = load_default_params()
+  p = Problem( problem_params, force_init = True )
+  
+  # log_P       = theta[0]
+  # log_delta   = theta[1]
+  # log_N0      = theta[2]
+  # log_sigma_d = theta[3]
+  # log_sigma_p = theta[4]
+  # tau         = theta[5]
+  
+  bin_ranges = [(-10,3),(-10,3),(2,32)]
+  
+  bin_ranges = [(-3,7),(-6,6),(2,10)]
+  theta_names = [r'$\log P$', r'$\log \delta$', r'$\log N_0$', r'$\log \sigma_d$',r'\log \sigma_p',r'$\tau']
+  #theta_names = ["A","B","C","D","E","F"]
+  
+  colors = ["b","g","r","b","g","r"]
+  theta_ids =   [0,1,2]
+  #theta_ids =   [3,4,5]
+  #names = [r"REJ $\epsilon 5$",r"p-SL S50",r"GPS $\xi 0.3$"]
+  names = ["REJ","SL","GPS"]
+  #names = ["REJ 10","REJ 7.5","REJ 5"]
+  #names = [r"m-SL 2 $\xi$ 0.1",r"ASL 2 $\xi$ 0.2",r"ASL 2 $\xi$ 0.3"]
+  #names = [r"m-SL 2",r"m-SL 10",r"m-SL 50"]
+  
+  thetas1 = np.loadtxt("./uai2014/runs/blowfly/rejection/eps5p0_repeat1_thetas.txt")
+  thetas2 = np.loadtxt("./uai2014/runs/blowfly_eps/sl_pseudo_s10/just_gaussian_repeat2_thetas.txt")
[…]

### 2015-02-26 code/exponential_problem/thermo_sample.py

@@ -7,7 +7,7 @@ from abc_sgld.code.working_code import *
 
 keep_x        = True 
 init_seed     = 1
-T             = 15000 # nbr of samples
+T             = 5000 # nbr of samples
 verbose_rate  = 1000
 C             = 20.01    # injected noise variance parameter
 eta           = 0.001 # step size for Hamiltoniam dynamics

### 2015-02-26 code/working_code.py

@@ -125,18 +125,25 @@ def hamiltonian_accept( problem, theta, theta_proposal, x, x_proposal, p, p_prop
     
   return theta, x, loglike_x 
 
-def init_omega( omega_params ):
+def init_omega( omega_params, S ):
   if omega_params["use_omega"] is False:
     omega = None
   else:
-    omega = np.random.randint(max_int_for_omega)
+    omega = np.zeros(S,dtype=int)
+    for s in range(S):
+      omega[s] = np.random.randint(max_int_for_omega)
   return omega
   
-def omega_flip(problem, theta, x, omega, loglike_x ):
+def omega_switch(problem, theta, x, omega, loglike_x, omega_params ):
+  any_changes = False
+  for s in range(len(omega)):
+    if np.random.rand() < omega_params["omega_rate"]:
+      omega[s]     = np.random.randint(max_int_for_omega)
+      any_changes = True
   
-  omega     = np.random.randint(max_int_for_omega)
-  x         = problem.simulate( theta, omega )
-  loglike_x = problem.loglike_x( x )
+  if any_changes:
+    x         = problem.simulate( theta, omega )
+    loglike_x = problem.loglike_x( x )
     
   return theta, x, omega, loglike_x
     
@@ -166,7 +173,7 @@ def run_mcmc( problem, params, theta, x = None ):
   upper_bounds = params["upper_bounds"]
   
   keep_x        = True
-  omega = init_omega( omega_params )
+  omega = init_omega( omega_params, S )
   outputs = {}
   
   # pseudo-data
@@ -222,17 +229,17 @@ def run_mcmc( problem, params, theta, x = None ):
     # --------------- #
     # samples omegas  #
     # --------------- #
-    if omega_params["use_omega"] and np.random.rand() < omega_params["omega_rate"]:
+    if omega_params["use_omega"]:
       if omega_params["omega_sample"]:
         # propose new omega and accept/reject using MH
         theta, x, omega, loglike_x = omega_sample(problem, theta, x, omega, loglike_x )
       elif omega_params["omega_switch"]:
         # randomly switch to new omega
-        omega = init_omega( omega_params )
+        theta, x, omega, loglike_x = omega_switch(problem, theta, x, omega, loglike_x, omega_params )
     
     if keep_x:
-      x = problem.simulate( theta, omega, S )
-      loglike_x = problem.loglike_x( x )
+      #x = problem.simulate( theta, omega, S )
+      #loglike_x = problem.loglike_x( x )
[…]

### 2015-02-27 code/blowfly_problem/blowfly_mcmc_get_epsilons.py
blowfly code

@@ -1,0 +1,64 @@
+from abcpy.factories import *
+from abcpy.problems.blowfly.blowfly import *
+
+import numpy as np
+import pylab as pp
+
+# exponential distributed observations with Gamma(alpha,beta) prior over lambda
+problem_params = default_params()
+problem_params["epsilon"] = 2.5*np.ones(10)
+problem_params["min_epsilon"] = 0.5*np.ones(10)
+problem_params["q_factor"] = 0.2
+problem_params["blowfly_filename"] = "./data/blowfly.txt"
+problem = BlowflyProblem( problem_params, force_init = True )
+
+
+# # since we are running abc_rejection, use a distance epsilon state
+# state_params = {}
+# state_params["S"]                          = 1
+# state_params["obs_statistics"]             = problem.get_obs_statistics()
+# state_params["theta_prior_rand_func"]      = problem.theta_prior_rand
+# state_params["theta_prior_logpdf_func"]    = problem.theta_prior_logpdf
+# state_params["theta_proposal_rand_func"]   = problem.theta_proposal_rand
+# state_params["theta_proposal_logpdf_func"] = problem.theta_proposal_logpdf
+# state_params["simulation_function"]        = problem.simulation_function
+# state_params["statistics_function"]        = problem.statistics_function
+# state_params["log_kernel_func"]            = log_gaussian_kernel
+# state_params["is_marginal"]                = True
+# state_params["epsilon"]                    = 1.0 #np.array([0.1,0.1,0.1,1.0])
+
+nbr_samples = 1000
+#epsilon     = 0.5
+state_params = state_params_factory.scrape_params_from_problem( problem, S = 5 )
+mcmc_params  = mcmc_params_factory.scrape_params_from_problem( problem, type="mh", is_marginal = True, nbr_samples = nbr_samples )
+algo_params = { "modeling_approach"  : "kernel",
+                "observation_groups" : problem.get_obs_groups(),
+                "state_params"       : state_params,
+                "mcmc_params"        : mcmc_params,
+                "algorithm"          : "model_mcmc"
+              }
+og = algo_params["observation_groups"][0]
+og.params["response_params"]['min_epsilon'] = problem_params["min_epsilon"]
+og.params["response_params"]["epsilon_update_rule"]=True
+og.params["response_params"]["epsilon_update_params"]={"decay":0.995,"quantile":0.25}
+recorder_params = {}  
+algo, model, state  = algo_factory.create_algo_and_state( algo_params )
[…]

### 2015-02-27 code/blowfly_problem/blowfly_runs.py

@@ -1,0 +1,266 @@
+import numpy as np
+import pylab as pp
+import scipy as sp
+from scipy import stats as spstats
+from blowfly_wrapper import *
+from abc_sgld.code.working_code import *
+
+def view_results( problem, thetas, all_stats, burnin = 1 ):
+  stats = all_stats.mean(1)
+  self=problem
+  pp.rc('text', usetex=True)
+  pp.rc('font', family='times')
+  f = pp.figure( figsize=(12,8) )
+  sp=pp.subplot(1,1,1)
+  
+  # plotting params
+  nbins       = 20
+  alpha       = 0.5
+  label_size  = 8
+  linewidth   = 3
+  linecolor   = "r"
+  
+  f=pp.figure()
+  for i in range(6):
+    sp=f.add_subplot(2,10,i+1)
+    pp.hist( thetas[:,i], 10, normed=True, alpha = 0.5)
+    #pp.title( self.theta_names[i])
+    set_label_fonsize( sp, 6 )
+    set_tick_fonsize( sp, 6 )
+    set_title_fonsize( sp, 8 )
+  for i in range(10):
+    sp=f.add_subplot(2,10,10+i+1)
+    pp.hist( stats[:,i], 10, normed=True, alpha = 0.5)
+    ax=pp.axis()
+    pp.vlines( self.obs_statistics[i], 0, ax[3], color="r", linewidths=2)
+    # if self.obs_statistics[i] < ax[0]:
+    #   ax[0] = self.obs_statistics[i]
+    # elif self.obs_statistics[i] > ax[1]:
+    #   ax[1] = self.obs_statistics[i]
+    pp.axis( [ min(ax[0],self.obs_statistics[i]), max(ax[1],self.obs_statistics[i]), ax[2],ax[3]] )
+    #pp.title( self.stats_names[i])
+    #set_label_fonsize( sp, 6 )
+    #set_tick_fonsize( sp, 6 )
+    #set_title_fonsize( sp, 8 )
+  #pp.suptitle( "top: posterior, bottom: post pred with true")
+  
+  # f = pp.figure()
+  # I = np.random.permutation( len(thetas) )
+  # for i in range(16):
+  #   sp=pp.subplot(4,4,i+1)
+  #   theta = thetas[ I[i],:]
+  #   test_obs = self.simulation_function( theta )
+  #   test_stats = self.statistics_function( test_obs )
+  #   err = np.sum( np.abs( self.obs_statistics - test_stats ) )
[…]

### 2015-02-27 code/blowfly_problem/blowfly_wrapper.py

@@ -1,0 +1,393 @@
+from abcpy.factories import *
+from abcpy.problems.blowfly.blowfly import *
+
+import numpy as np
+import pylab as pp
+
+
+from abc_sgld.code.working_code import *
+#from sa_algorithms import *
+from scipy import stats as spstats
+import pylab as pp
+
+default_epsilons = 0.25+0*np.array([ 0.08257003,  0.01013485,  0.01,  0.01, 0.01 , 0.01,  0.01,0.01,0.53027013,0.13338243] )
+
+# exponential distributed observations with Gamma(alpha,beta) prior over lambda
+problem_params = default_params()
+problem_params["epsilon"] = default_epsilons
+#problem_params["min_epsilon"] = 0.1*np.ones(10)
+problem_params["q_factor"] = 0.2
+problem_params["blowfly_filename"] = "./data/blowfly.txt"
+problem_params["tau_is_log_normal"] = True
+bf_problem = BlowflyProblem( problem_params, force_init = True )
+
+state_params = state_params_factory.scrape_params_from_problem( bf_problem, S=1 )
+
+# set is_marginal to true so using only "current" state will force a re-run
+mcmc_params  = mcmc_params_factory.scrape_params_from_problem( bf_problem, type="mh", is_marginal = False, nbr_samples = 1 )
+
+# ignore algo, just doing this to get state object
+algo_params = { "modeling_approach"  : "kernel",
+                "observation_groups" : bf_problem.get_obs_groups(),
+                "state_params"       : state_params,
+                "mcmc_params"        : mcmc_params,
+                "algorithm"          : "model_mcmc"
+              }
+algo, model, state  = algo_factory.create_algo_and_state( algo_params )
+recorder     = recorder_factory.create_recorder( {} )
+
+
+class generate_blowfly( object ):
+  def __init__( self, abcpy_problem ):
+    self.p = abcpy_problem
+    self.y = np.squeeze( self.p.obs_statistics )
+    self.x_epsilon = None
+    self.J=len(self.y)
+    
+
+    # multivariate uses variance (norm uses std dev)
+    self.kernel = spstats.multivariate_normal( self.p.obs_statistics, self.p.epsilon**2 )
+    
[…]

### 2015-02-27 code/exponential_problem/figs_runs_posteriors.py

@@ -68,7 +68,7 @@ def view_posterior( problem, theta_range, samples, algoname, burnin = 1000 ):
 
 keep_x        = True 
 init_seed     = 4
-T             = 20000 # nbr of samples
+T             = 50000 # nbr of samples
 verbose_rate  = 1000
 C             = 1.01    # injected noise variance parameter
 eta           = 0.01 # step size for Hamiltoniam dynamics
@@ -152,7 +152,7 @@ if __name__ == "__main__":
   # sghmc = run_sghmc( problem, params, theta0, x0 )
   
   #
-  # np.random.seed(init_seed + 1000*chain_id)
+  np.random.seed(init_seed + 1000*chain_id)
   sgld = run_sgld( problem, params, theta0, x0 )
   algoname = "SG-Langevin"
   view_posterior( problem, theta_range, sgld["THETA"], algoname, burnin=1000 )
@@ -163,9 +163,9 @@ if __name__ == "__main__":
   if saveit:
     pp.savefig("exp-%s-theta-timeseries.pdf"%(algoname), format="pdf", dpi=600,bbox_inches="tight")
     pp.savefig("../../papers/uai-2015/images/exp-%s-theta-timeseries.pdf"%(algoname), format="pdf", dpi=600,bbox_inches="tight")
-  
   #
-  # np.random.seed(init_seed + 1000*chain_id)
+  #
+  np.random.seed(init_seed + 1000*chain_id)
   mcmc = run_mcmc( problem, params, theta0, x0 )
   algoname = "ABC-MCMC"
   view_posterior( problem, theta_range, mcmc["THETA"], algoname, burnin=1000 )
@@ -177,6 +177,10 @@ if __name__ == "__main__":
     pp.savefig("exp-%s-theta-timeseries.pdf"%(algoname), format="pdf", dpi=600,bbox_inches="tight")
     pp.savefig("../../papers/uai-2015/images/exp-%s-theta-timeseries.pdf"%(algoname), format="pdf", dpi=600,bbox_inches="tight")
   
+  results = {}
+  results["mcmc"] = mcmc
+  results["sgld"] = sgld
+  results["sgnht"] = sgnht
   # pp.figure(1)
   # pp.clf()
   # pp.plot( mcmc["THETA"][-1000:])

### 2015-02-27 code/working_code.py

@@ -2,6 +2,7 @@ import numpy as np
 import scipy as sp
 from scipy import stats as spstats
 import pylab as pp
+import pdb
 
 max_int_for_omega = 10**8
 
@@ -79,13 +80,15 @@ def bounce_off_boundaries( theta, p, lower_bounds, upper_bounds ):
   D = len(theta)
   
   for d in range(D):
-    if theta[d] < lower_bounds[d]:
-      theta[d] = lower_bounds[d] + (lower_bounds[d]-theta[d]) #from Neal Handbook, p37
-      p[d] *= -1
+    if lower_bounds is not None:
+      if theta[d] < lower_bounds[d]:
+        theta[d] = lower_bounds[d] + (lower_bounds[d]-theta[d]) #from Neal Handbook, p37
+        p[d] *= -1
 
-    if theta[d] > upper_bounds[d]:
-      theta[d] = upper_bounds[d] - (theta[d]-upper_bounds[d]) #from Neal Handbook, p37
-      p[d] *= -1
+    if upper_bounds is not None:
+      if theta[d] > upper_bounds[d]:
+        theta[d] = upper_bounds[d] - (theta[d]-upper_bounds[d]) #from Neal Handbook, p37
+        p[d] *= -1
   return theta, p
   
 
@@ -142,7 +145,7 @@ def omega_switch(problem, theta, x, omega, loglike_x, omega_params ):
       any_changes = True
   
   if any_changes:
-    x         = problem.simulate( theta, omega )
+    x         = problem.simulate( theta, omega, len(omega) )
     loglike_x = problem.loglike_x( x )
     
   return theta, x, omega, loglike_x
@@ -150,7 +153,7 @@ def omega_switch(problem, theta, x, omega, loglike_x, omega_params ):
 def omega_sample(problem, theta, x, omega, loglike_x ):
   
   omega_proposal     = np.random.randint(max_int_for_omega)
-  x_proposal         = problem.simulate( theta, omega_proposal )
+  x_proposal         = problem.simulate( theta, omega_proposal, len(omega) )
   loglike_x_proposal = problem.loglike_x( x_proposal )
 
   log_acceptance =  loglike_x_proposal  \
@@ -182,7 +185,7 @@ def run_mcmc( problem, params, theta, x = None ):
       x = problem.simulate( theta, omega, S )
     assert len(x.shape) == 2, "x should be S by dim"
     loglike_x = problem.loglike_x( x )
[…]

### 2015-02-28 papers/uai-2015/draft.tex
draft -- exp problem text, code exa-variance

@@
 [We may want anti-thetic instead of CRNs, ie overrelax versus underrelax, etc.]
-% [CITE NEAL How to view an MCMC...]
-% Another variance reduction technique called the method of {\em common random numbers} (CRN) \cite{kleinman1999simulation} sets a common seed for the random number generator (RNG) for both calls to the simulator.
-This technique can remove the effect of the simulator noise in the gradient estimate, leaving on the randomness of the perturbation masks as the source of noise.
-Consider using SPSA instead of SGD: the CRN technique is equivalent of using the same batch of data vectors to evaluate the log-likelihood which is a sensible approach.
-%
-%
-% [By observing the simulation at different theta but same seeds, we can eliminate much of the noise in the markov chain.
-By treating omega as part of the state, we can smoothly move along noiseless outputs for several iterations and then step.]
-
-% In the application of stochastic gradient Hamiltonian dynamics to ABC, the stochasticity of the gradient due to batch sizes is instead caused by two other factors.
-The first factor, is due to the finite number of forward simulations that are used to approximate the log-likelihood function.
-We will explain below how to perform ABC gradient estimates based on finite difference approximations on two log-likelihood formulations: the $\epsilon$-kernel density model and on a Gaussian conditional model, both are based on $S$ random simulations.
-The second factor is due to the dimension of $D$.
-For large $D$, a finite difference estimate requires $2p$ forward simulations for each random simulation, by using a further stochastic approximation to the gradient, called Simultaneous Perturbation Stochastic Approximation (SPSA) \cite{spall}, an unbiased estimate of the gradient can computed with $2q$ forward simulations, where $ \geq 1$ and usually $Q \ll D$.
-
-% To begin, we define $\thetavplusd = \thetav + c \Deltavd$ and $\thetavminusd = \thetav + c \Deltavd$, where $c$ is a small positive scalar, e.g.
-$c=0.001$, and $\Deltavd$ is a vector of size $D$ that, for FD estimates is $1$ for entry $d$ and $0$ otherwise.
-Our gradient estimate require simulating $S$ times at $\thetavplus$ and $\thetavminus$, generating sets of pseudo-data $\{ \xplussd \}$ and $\{ \xminussd \}$.
-These are then plugged into the log-likelihood estimate for $\y$, which we call $\mathcal{L}$:
-% \begin{eqnarray}
-%   \frac{\partial \mathcal{L}(\thetav)}{\partial \thetavd} & \approx & \frac{\mathcal{L}(\thetavplusd)- \mathcal{L}(\thetavminusd)}{2c}
-% \end{eqnarray}
-
-% For the $\epsilon$-kernel density model
-% % ,
-% % \begin{eqnarray}
-% %   \mathcal{L}(\thetavpmd) & = & \log \sum_{s=1}^2 \mathcal{N}( \y | f( \thetavpmd, \omega_s ), \eps^2 ) \\
-% %                           & = & \log \mathcal{N}( \y | f( \thetavpmd, \omega_{m} ), \eps^2 ) + \log\lp 1 + H_d \rp \\
-% %                           & \approx & \log \mathcal{N}( \y | f( \thetavpmd, \omega_{m} ), \eps^2 ) + 0
-% % \end{eqnarray}
-% % where we replaced the random variable $\x$ with a deterministic function $f( \thetavpmd, \omega_s )$ (i.e.
-the simulator run at $\thetavpmd$ using the random state $\omega_s$, such that $\xpmd = f( \thetavpmd, \omega_s )$).
-In the second line we have pulled out the value from the random state $m$ which has the maximum log-likelihood (i.e.
-is closest to $\y$) and used $H_d$ (sum of terms $<0$ and $\ll 0$ for small $\eps$), and in the last line simplified further by assuming small $\eps$, so that the log-likelihood is dominated by a single simulation closest to $\y$.
-% it is easy to show that the gradient estimate is sensitive to $\epsilon$ in this case:
-% \begin{eqnarray}
-%   \frac{\partial \mathcal{L}(\thetav)}{\partial \thetavd} & \approx & \frac{1}{2c\eps^2}\lp -\frac{1}{2}(\y-\xplussd)^2 + \frac{1}{2}(\y-\xminussd)^2  \rp
-% \end{eqnarray}
-% So by making $\epsilon$ arbitrarily small we can make the gradient extremely sensitive to our simulations.
-Intuitively, we want the gradient to be fairly stable no matter what values of $\epsilon$ we use.
-To do this we will replace log-likelihood with a conditional model.
-%
-% A better way of approximating the gradient is to use a conditional model of $\pi(\x|\thetav)$, for example a conditional kernel density estimate \cite{TurnerGenLik2014} or a conditional Gaussian model \cite{wood2010statistical}.
-While either approach should be adequate and both have their own limits and advantages, for this discussion we will use a Gaussian estimate.
-In ABC, using a conditional Gaussian model for  $\pi(\x|\thetav)$ is called a {\em synthetic likelihood} (SL) model \cite{wood2010statistical}.
-For a SL log-likelihood model, we compute estimators of the first and second moments of $\pi(\x|\thetav)$.
-For Gaussian $\epsilon$-kernel, we can convolve the two densities
-% \begin{eqnarray}
-%   \mathcal{L}(\thetavpmd) & = & \log \int \mathcal{N}( \y | \x, \eps^2 ) \mathcal{N}( \x |
-%   \mu_{\thetavpm}, \sigma^2_{\thetavpm} ) d\x \\
-%                           & = & \log \mathcal{N}( \y | \mu_{\thetavpm}, \sigma^2_{\thetavpm} \eps^2 )
-% \end{eqnarray}
-% We can see that $\epsilon$ now can be set very small as it now acts as a smoother for our estimate, resulting in a robust gradient.
-
-%\subsection{Gradients in High Dimensions}
-
-%\subsection{Spall's method: SPSA}
-% In the gradient-free setting, Spall \cite{spall1992multivariate} provides a stochastic approximate to the true gradient using only 2 forward simulations (function evaluations).
-This is in contrast to multivariate finite-difference stochastic approximation FDSA \cite{kiefer1952stochastic} requiring $2p$ evaluations.
-%
-% The gradient estimate is
-% \begin{equation}
-%   \gradestimate{t}{\thetav_t} = \begin{bmatrix}
-%                                    \frac{\feval{t}{+} - \feval{t}{-}}{2 \cstep{t}{1} \dstep{t}{1}} \\
-%                                    \vdots \\
-%                                    \frac{\feval{t}{+} - \feval{t}{-}}{2 \cstep{t}{p} \dstep{t}{p}} \\
-%                                 \end{bmatrix}
-% \end{equation}
-% where $\cstep{t}{p}$ is a step-size that is usually constant for all dimensions $p$, but can be different (as shown in this case); $\dstep{t}{p} \in {-1, +1}$ is a {\em perturbation mask} (called symmetric Bernouilli variables by Spall), i.e.
-$\dstep{t}{p} \sim 2*\text{Bernouilli}(0.5) - 1$; and  $\feval{t}{\pm}$ are function evaluations:
-% \begin{eqnarray}
-%   \feval{t}{+} &= \loglik{\thetav + \cstepconst{t}\dstepconst{t}} \\
-%   \feval{t}{-} &= \loglik{\thetav - \cstepconst{t}\dstepconst{t}}
-% \end{eqnarray}
-% A way of reducing the noise in the gradient is by averaging over $q$ draws of $\dstepconst{t}$:
-% \begin{equation}
-%   \gradestimate{t}{\thetav_t} = \frac{1}{q} \sum_{j=1}^q \gradestimaterepeat{t}{\thetav_t}{j}
-% \end{equation}
-% where for each $j$ new perturbation masks are drawn.
-%
-
-% \subsection{Stochastic ABC Gradients}
-% For low-dimensional problems a finite difference (FD) approximation to $\nabla U$.
-For each dimension $p$, forward simulate for each random seed twice at plus $c$ and minus $c$ (only along dimension $p$): $x_{\pm}^{(s)} \sim \pi( \x | \thetav \pm c \Delta_p, \omega_s)$.
-These simulations are then plugged into the appropriate estimator (logsumsexp or SL).
-$\hat{U}_{K_{\eps}}$:
-% \begin{eqnarray}
-%   - \log \pi(\thetav \pm c \Delta_p) - \log \sum_s \pi_{\eps}(\y | f( \thetav \pm c \Delta_p, \omega_s ) )
-% \end{eqnarray}
-% or $\hat{U}_{SL}$
-% \begin{eqnarray}
-%   - \log \pi(\thetav \pm c \Delta_p) - \log \mathcal{N}(\y | \mu_{\thetav \pm c \Delta_p}, \sigma^2_{\thetav \pm c \Delta_p} + \eps^2 )
-% \end{eqnarray}
-%
-%
-% \begin{eqnarray}
-%   \hat{U}(\thetav)       & = & - \log \pi(\thetav) - \log \pi_{\eps}(\y | \thetav ) \\
-%   \nabla\hat{U}(\thetav) & = & - \nabla \log \pi(\thetav) - \frac{N}{n} \sum_{i=h_1}^{h_n} \nabla \log \pi(\yi | \thetav )
-% \end{eqnarray}
-
-%\subsection{Common Random Numbers}
-
-% Using the same parameter vector $\thetav$ but different random seeds:
-% \begin{eqnarray}
-%   \pi\lp \y | \thetav \rp & = &  \int \pi_{\epsvec}\lp \y | \x \rp \pi\lp \x | \thetav \rp d \x \\
-%    & \approx & \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}\lp \y | \x^{(s)} \rp
-% \end{eqnarray}
-% where $\x^{(s)} = \pi\lp \x | \thetav, \omega_s \rp$.
-
-% Note we have passed a unique random seed $\omega_s$ into the simulator, making $\x^{(s)}$ a deterministic function of the simulator with parameters $\{\thetav, \omega_s\}$.
-It will be useful to rewrite this deterministic simulator function as $f$: $\x^{(s)} = f\lp \thetav, \omega_s \rp$, allowing us to rewrite the likelihood as
-% \begin{eqnarray}
-%   \pi\lp \y | \thetav \rp & \approx & \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}\lp \y | f\lp \thetav, \omega_s \rp \rp
-% \end{eqnarray}
-
-% \subsection{Conditional Density Estimators}
-% [TODO: TAKEN FROM POPE]  When the simulator is fast and stochastic, it can be beneficial to the inference procedure to build a local, conditional model of the distribution $\pi(\y|\thetav)$ using $S$ simulator responses in $\yone, \ldots, \yS \overset{\simulator}{\sim} \pi( \y | \thetav)$.
-% %
-% The simplest local response model is the \emph{conditional Gaussian}, an approach called {\em synthetic likelihood} in ABC \cite{wood2010statistical}.
-It computes estimators of the first and second moments of the responses and uses the Gaussian distribution to analytically compute the likelihood (thus providing an alternative to kernel ABC).
-For our algorithms, this allows the direct computation of the CDF:
-% \begin{eqnarray}
-%  && \muhattheta  =  \frac{1}{S} \sum_{s=1}^{S} \y_s ~~~~~~~~~~~
-%  \hat{\Sigma}_{\thetav}  =  \frac{1}{S-1} \sum_{s=1}^{S} \lp \ys - \muhattheta \rp \lp \ys - \muhattheta \rp^T\\
-%  &&\F_{\y | \thetav}(\ystar; \muhattheta,  \hat{\Sigma}_{\thetav})  =  \int_{-\infty}^{\ystar} \mathcal{N}\lp \y | \muhattheta,  \hat{\Sigma}_{\thetav} \rp d \y
-% \end{eqnarray}
-% % \begin{eqnarray}
-% %   &&\F_{\y | \thetav}(\ystar; \muhattheta,  \hat{\Sigma}_{\thetav})  =  \int_{-\infty}^{\ystar} \mathcal{N}\lp \y | \muhattheta,  \hat{\Sigma}_{\thetav} \rp d \y
-% % \end{eqnarray}
-% where $\muhattheta$ and $\hat{\Sigma}_{\thetav}$ are computed from the $S$ simulations.
-In our experiments we  often use a factorized model: $\mathcal{N}(\y | \muhattheta,\hat{\Sigma}_{\thetav}) \approx \prod_{j=1}^J \mathcal{N}( y_j | \muhatthetaj,  \sigmahatthetaj)$, resulting in a factorized product over CDFs as well.
-% Modeling the response by only the first two moments may be inadequate due to multi-modality, asymmetric noise, etc.
-For such cases a \emph{conditional KDE} (kernel density estimate) response model can by used.
-In \cite{TurnerGenLik2014} this approach is shown to be superior to conditional Gaussians for certain computational psychology models.
-Note that for Gaussian kernels the conditional KDE is very similar to kernel ABC, but has additional flexibility of adaptively choosing bandwidths (rather than the fixed $\epsvec$ in kernel ABC).
-% [TODO: END TAKEN FROM POPE]
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-[FROM UAI] In this illustrative problem we infer the rate of an exponential distribution under a gamma prior with shape $\alpha$ and rate $\beta$, having $N$ observations at the true rate $\theta^{\star}$; this is the {\em exponential example} in \cite{turner2012tutorial}.
-Let $\w$ be a vector of $N$ draws from an exponential distribution, i.e.
-$w_n \sim \expdist(\theta)$.
-The posterior is a gamma distribution with shape $\alpha+N$ and rate $\beta + \sum w_n$.
-To use this problem with ABC, we use the exponential draws as the simulator and the mean of $\w$ as the statistic $y$ and assume that $N$ is known.
-The inference problem is therefore to sample from $p( \theta | y, \alpha, \beta, N )$.
+We use a simple $D=1$ problem to demonstrate HABC.
+Let $y= \frac{1}{N} \sum_{i} e_i$, where $e_i \sim \text{Exp}(1/\thetastar)$; $\thetastar = 0.15$, $N=20$, and $y=7.74$ in our concrete example.
+Assuming $\pi(\theta ) = \text{Gamma}(\alpha, \beta)$, the true posterior is a gamma distribution with shape $\alpha+N$ and rate $\beta + N y$.
+Our simulator therefore generates the average of $N$ exponential random variates with rate $\lambda = 1/\theta$.
+Data $x \simsim \pi(x|\theta)$ are shown in Figure~\ref{fig:exp-crns}.
+We have explicitly shown the smoothness of the simulator by generating data along trajectories of fixed seeds $\omega_s$; i.e.
+for several $\omega_s$ we vary $\theta$ (dashed lines are function $f(\theta, \omega_s)$) and randomly reveal data (blue circles).
+The horizontal line with shading indicates $y \pm 2 \eps$, where $\eps = 0.37$ is used throughout the demonstration.
 
@@
 
+\subsection{Bias and Variance of $\nabla \hat{U}(\thetav)$}
+At $\theta_{\text{MAP}}$ we estimated the gradient $\nabla \hat{U}(\theta)$ using finite differences for both the kernel-$epsilon$ and synthetic-likelihood representation of $\pi_{eps}(y | x )$.
+For $S=5$ and $S=50$, $10K$ repetitions of computing the gradients was performed; there histograms are shown in Figure~\ref{fig:exp-varg}.
+At $S=5$ both methods reveal a small bias (values X and Y for kernel-$\eps$ and SL, respectively).
+At $S=50$, the bias for kernel-$\eps$ is much smaller, but its variance remains high; for SL, the bias remains unchanged but the variance has decreased significantly (values X pm S and Y pm T).
+The bias in the SL representation comes from the Gaussian approximation $\pi(y|\theta)$ which we can compute analytically using the CLT as $\text{N}( 1/\theta, 1/N \theta^2 + \eps^2)$.
+(We believe this bias is due to the normal approximation assuming a symmetric mode as the bias goes to zero at the mean of the true posterior).
+Note that this bias is directional signal only appears nears the mode.
+For the $S=5$ setting in our experiment both approaches contain similar bias.
+[REWRITE END]
+
+\subsection{Posterior Inference using HABC}
+In Figures~\ref{fig:exp-theta-traces} we show the posterior trace plots for ABC-MCMC, SGLD, SGHMC, and SGNHT versions of HABC using SL gradient estimates ($S=5$).
+The behavior of the trajectories is quite different, though they all produce the same correct posterior estimates (Figure~\ref{fig:exp-posteriors}).
+In Table~\ref{tab:exp-posteriors-nosticky} we report the {\em total variational distance} between the true posterior and the ABC posteriors after different numbers of samples (repeated 5 times).
+
 \begin{figure*}[t]
@@
 \begin{center}
-\includegraphics[width=0.5\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
-\includegraphics[width=1.2\columnwidth]{./images/exp-ABC-MCMC-theta-timeseries.pdf}
-\includegraphics[width=0.5\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
-\includegraphics[width=1.2\columnwidth]{./images/exp-SG-Langevin-theta-timeseries.pdf}
-\includegraphics[width=0.5\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
-\includegraphics[width=1.2\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries.pdf}
+\includegraphics[width=1.0\columnwidth]{./images/exp-ABC-MCMC-theta-timeseries.pdf}
+\includegraphics[width=1.0\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries.pdf}
+\includegraphics[width=1.0\columnwidth]{./images/exp-SG-Langevin-theta-timeseries.pdf}
+\includegraphics[width=1.0\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries.pdf}
+\caption{\small{}}
+\label{fig:exp-theta-traces}
+\end{center}
+\vskip -0.2in
+\end{figure*}
+
+\begin{figure}[t]
+\vskip 0.2in
+\begin{center}
+\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
+\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
+\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
+\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
 \caption{\small{}}
@@
 \vskip -0.2in
-\end{figure*}
+\end{figure}
 
-\subsection{First order CRNs}
-Using same seed in a gradient estimate at time $t$ (versus not using them).
-This is makes sense if we compare to SG-data where a single batch is used for computing the gradient.
-If a FD method was used, we would use the same batch.
+\subsection{Sticky Random Numbers}
+Finally we compare posterior estimates using sticky CRNs.
+Using a flip rate of $1\%$, i.e.
+each $\omega_s$ has $1\%$ chance of changing randomly at each time-step.
+In Figure~\ref{exp-traces-x-sticky} we show the traces plots of $<x^{(s)>}$ for the different methods.
+While the Hamiltonian dynamics provide persistence in $\thetav$-space, the sticky random numbers do the same for the ABC statistics.
+The Table~\ref{tab:exp-tvd-sticky} we report the same {\em tvd} error using sticky numbers.
+[DISCUSS difference with no-sticky].
+The potential benefit of using persistent random seeds is shown for all methods, including ABC-MCMC.
 
-\subsection{Second order CRNs}
-Using same seed in a gradient estimate at multiple time steps (versus not changing them -- the previous example).  Analogous to keeping the same batch for consecutive steps in SG-data.
-
-\subsection{Comparing Gradient Estimates with SL}
-\begin{itemize}
-  \item Can estimate ABC posterior by CLT of Gammas (sum of exponential).
-For large S, the variance of the SL approximates the ``true'' CLT approximation.
-Thus we can set $\epsilon$ much smaller for the SL case.
-The kernel logsumexp suffers when $\epsilon$ is small, using only the closest simulation to $\y$.
-Instead the SL estimates a density around the simulations and the likelihood is less sensitive to $\epsilon$.
-  \item Better tolerance to low $\epsilon$ using SL.
-\end{itemize}
-
-\subsection{Using variance of gradients in SGHMC}
-Compute online $\hat{B}$ then fix.
-
-\subsection{Visualizing Noise from Gradient versus Injected Noise}
+%\subsection{Visualizing Noise from Gradient versus Injected Noise}
 
@@
 We can make the approximation arbitrarily good by running more simulations, but this is often unnecessary for sampling, because we can make use of the uncertainty in our current approximation to decide if our sample is correct or not (cite UAI2014?).
+  \item Unbiased gradients with low noise.  SL bias might be intolerable; use KDE to get stable unbiased gradients?
 \end{itemize}
@@
 
-\subsubsection*{References}
-{
+%\subsubsection*{References}
+%{
+\clearpage
 \bibliographystyle{icml2014}
 \bibliography{abcsgld}
-}
+%}
 

### 2015-02-28 code/blowfly_problem/blowfly_mcmc_get_epsilons.py

@@ -6,7 +6,7 @@ import pylab as pp
 
 # exponential distributed observations with Gamma(alpha,beta) prior over lambda
 problem_params = default_params()
-problem_params["epsilon"] = 2.5*np.ones(10)
+problem_params["epsilon"] = 0.5*np.ones(10)
 problem_params["min_epsilon"] = 0.5*np.ones(10)
 problem_params["q_factor"] = 0.2
 problem_params["blowfly_filename"] = "./data/blowfly.txt"
@@ -37,10 +37,10 @@ algo_params = { "modeling_approach"  : "kernel",
                 "mcmc_params"        : mcmc_params,
                 "algorithm"          : "model_mcmc"
               }
-og = algo_params["observation_groups"][0]
-og.params["response_params"]['min_epsilon'] = problem_params["min_epsilon"]
-og.params["response_params"]["epsilon_update_rule"]=True
-og.params["response_params"]["epsilon_update_params"]={"decay":0.995,"quantile":0.25}
+#og = algo_params["observation_groups"][0]
+#og.params["response_params"]['min_epsilon'] = problem_params["min_epsilon"]
+#og.params["response_params"]["epsilon_update_rule"]=True
+#og.params["response_params"]["epsilon_update_params"]={"decay":0.995,"quantile":0.25}
 recorder_params = {}  
 algo, model, state  = algo_factory.create_algo_and_state( algo_params )
 recorder     = recorder_factory.create_recorder( recorder_params )

### 2015-02-28 code/blowfly_problem/blowfly_runs.py

@@ -3,8 +3,10 @@ import pylab as pp
 import scipy as sp
 from scipy import stats as spstats
 from blowfly_wrapper import *
-from abc_sgld.code.working_code import *
+from habc_algos import *
 
+#def pp_vector( x ):
+  
 def view_results( problem, thetas, all_stats, burnin = 1 ):
   stats = all_stats.mean(1)
   self=problem
@@ -60,7 +62,78 @@ def view_results( problem, thetas, all_stats, burnin = 1 ):
   #   set_tick_fonsize( sp, 6 )
   #   set_title_fonsize( sp, 8 )
   # pp.suptitle( "time-series from random draws of posterior")
-    
+
+def view_X_timeseries( problem, X, algoname, howmany = 1000 ):
+  muX = X.mean(1) # average over S
+  stdX = X.std(1)
+  x_names = problem.p.stats_names
+  
+  figsize = (16,6)
+  alpha=0.75
+  pp.rc('text', usetex=True)
+  pp.rc('font', family='times')
+  f = pp.figure( figsize=figsize )
+  sp=pp.subplot(1,1,1)
+      
+  for i in range(10):
+    n = len(muX[-howmany:,i])
+    sp=pp.subplot(5,2,i+1)
+    pp.plot( muX[-howmany:,i], 'k-', lw=2, alpha=0.5)
+    pp.fill_between( np.arange(n), muX[-howmany:,i]-2*stdX[-howmany:,i], muX[-howmany:,i]+2*stdX[-howmany:,i],color='b', alpha=0.5)
+    y = problem.y[i]
+    yplus = y + 2*problem.p.epsilon[i]
+    yminus = y - 2*problem.p.epsilon[i]
+    #pp.hlines( y,0,n-1,color='g',linewidth=2)
+    pp.fill_between( np.arange(n),yminus,yplus,color='r',alpha=0.25)
+    pp.hlines( y,0,n-1,color='r',linewidth=2)
+    pp.hlines( yplus,0,n-1,color='r',linewidth=1)
+    pp.hlines( yminus,0,n-1,color='r',linewidth=1)
+    pp.ylabel( x_names[i], rotation='vertical')
+    print "i ",x_names[i], y,yplus,yminus
+    #pdb.set_trace()
+    # set_tick_fonsize( sx_names[i]p, 16 )
+    # set_title_fonsize( sp, 32 )
[…]

### 2015-02-28 code/blowfly_problem/blowfly_wrapper.py

@@ -10,7 +10,7 @@ from abc_sgld.code.working_code import *
 from scipy import stats as spstats
 import pylab as pp
 
-default_epsilons = 0.25+0*np.array([ 0.08257003,  0.01013485,  0.01,  0.01, 0.01 , 0.01,  0.01,0.01,0.53027013,0.13338243] )
+default_epsilons = 0.5 + 0*np.array([ 0.08257003,  0.01013485,  0.01,  0.01, 0.01 , 0.01,  0.01,0.01,0.53027013,0.13338243] )
 
 # exponential distributed observations with Gamma(alpha,beta) prior over lambda
 problem_params = default_params()
@@ -18,7 +18,7 @@ problem_params["epsilon"] = default_epsilons
 #problem_params["min_epsilon"] = 0.1*np.ones(10)
 problem_params["q_factor"] = 0.2
 problem_params["blowfly_filename"] = "./data/blowfly.txt"
-problem_params["tau_is_log_normal"] = True
+#problem_params["tau_is_log_normal"] = True
 bf_problem = BlowflyProblem( problem_params, force_init = True )
 
 state_params = state_params_factory.scrape_params_from_problem( bf_problem, S=1 )

### 2015-02-28 code/blowfly_problem/habc_algos.py

@@ -1,0 +1,622 @@
+import numpy as np
+import scipy as sp
+from scipy import stats as spstats
+import pylab as pp
+import pdb
+
+max_int_for_omega = 10**8
+
+def bin_errors_1d( bins, true_centered_probability, samples ):
+  N = len(samples)
+  cnts, bins = np.histogram( samples ,bins=bins)
+  
+  n = cnts.sum()
+  
+  probability = cnts / (float(N))
+  
+  missed = N - n
+  
+  double_error = np.sum(np.abs( true_centered_probability - probability )) + missed/float(N)
+  
+  #pdb.set_trace()
+  return double_error/2.0
+
+def cramer_vonMises_criterion( true_P, samples, bins ):
+  # see wiki cramer_von_Mises
+  N = len(samples)
+  cnts, bins = np.histogram( samples ,bins=bins)
+  
+  n = cnts.sum()
+  
+  est_P = cnts / (float(N))
+  
+  # for samples outside bins, we assume true_P is zeros
+  missed = N - n
+  
+  discrete_cvm_error = np.sum( ( true_P - est_P)**2) + (missed/float(N))**2
+  
+  #pdb.set_trace()
+  return discrete_cvm_error
+  
+def bin_sq_errors_1d( bins, true_centered_probability, samples ):
+  N = len(samples)
+  cnts, bins = np.histogram( samples ,bins=bins)
+  
+  n = cnts.sum()
+  
+  probability = cnts / (float(N))
+  
+  missed = N - n
+  
+  double_error = np.sqrt( np.sum( ( true_centered_probability - probability)**2) + (missed/float(N))**2 )
+  
+  #pdb.set_trace()
+  return double_error/2.0
+  
+
+        
+def convergence_to_scalar( psi, psi_estimate, times ):
+  nbrEstimate = len( psi_estimate )
+  
+  T     = len(times)
+  means = np.zeros( T )
+  vrs   = np.zeros( T )
+  for time_id,t in zip(range(T),times):
+    print time_id, t
+    n = t
+    if time_id <= nbrEstimate:
+      mean_estimate = psi_estimate[:t+1].mean()
+      var_estimate  = float(1.0/(t-1))*(psi - mean_estimate)**2
+    
+      means[time_id] = mean_estimate
+      vrs[time_id]   = var_estimate
+      
+  means = means[:time_id+1]
+  vrs = vrs[:time_id+1]  
+  used_times = times[:time_id+1]  
[…]

### 2015-02-28 code/exponential_problem/exp_wrapper.py

@@ -84,11 +84,11 @@ class generate_exponential( object ):
     
     return X
   
-  def true_gradient( self, theta, gradients ):
+  def true_gradient( self, theta, gradients=None ):
     
     grad = (self.p.alpha+self.p.N-1)/theta - (self.p.beta+self.p.obs_sum)
     
-    params["logs"]["true"].append( np.squeeze( np.array([grad,theta])) )
+    #params["logs"]["true"].append( np.squeeze( np.array([grad,theta])) )
     return -grad 
 
   def true_abc_gradient( self, theta, d_theta, S, params ):

### 2015-02-28 code/exponential_problem/figs_run_gradient_distributions.py

@@ -118,9 +118,10 @@ if __name__ == "__main__":
   np.random.seed(4)
   problem = generate_exponential( exp_problem )
   
-  theta_map = problem.p.theta_posterior
+  theta_map = problem.p.posterior_mode
+  #theta_map = 1.0/problem.y #problem.p.posterior_mode
   
-  S = 50
+  S = 10
   N = 10000
   G = np.zeros( (N,3))
   for i in range(N):

### 2015-02-28 papers/uai-2015/draft.tex
bf description

@@
 \input{algo_fdsa_abc}
+\input{algo_spsa_abc}
 Algorithm~\ref{algo:fdsa} shows the {\em finite difference stochastic approximation} (FDSA) \cite{kiefer1952stochastic} to $\nabla U(\thetav)$ as a function of random seeds $\omegav$.
@@
 Algorithm~\ref{algo:spsa} shows SPSA to estimate $\nabla U(\thetav)$.  The number of simulations required for SPSA is $2 S R$, where $R \geq 1$.
-\input{algo_spsa_abc}
 
@@
 SPSA has also been used within a procedure for maximum-likelihood estimation for hidden Markov models using ABC \cite{Ehrlich2013}.
+
+\begin{figure}[t]
+\vskip 0.2in
+\begin{center}
+\includegraphics[width=0.95\columnwidth]{./images/exp_crn_figure.pdf}
+\caption{\small{A view of a simulator in terms of common random numbers.
+The horizontal line represents $\y$ and red shading $\pm 2\eps$.
+The shaded curved region represents $2\sigma$ of $\pi(\x|\thetav)$.
+The dashed lines are $f(\thetav, \omega_s)$ for several values of $\omega$.
+The blue circles are potential random samples from $\pi(\x|\thetav)$.
+For a fixed value $\omega_s$, the simulator produces deterministic outputs that change smoothly, even though the simulator itself is quite noisy.}}
+\label{fig:exp-crns}
+\end{center}
+\vskip -0.2in
+\end{figure}
 
@@
 
-[We may want anti-thetic instead of CRNs, ie overrelax versus underrelax, etc.]
+%[We may want anti-thetic instead of CRNs, ie overrelax versus underrelax, etc.]
+
+\begin{figure}[t]
+\vskip 0.2in
+\begin{center}
+\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
+\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
+\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
+\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
+\caption{\small{}}
+\label{fig:exp-posteriors}
+\end{center}
+\vskip -0.2in
+\end{figure}
 
@@
 
-\begin{figure}[t]
-\vskip 0.2in
+\begin{figure*}[t]
+%\vskip 0.2in
+\setlength{\linewidth}{\textwidth}
+\setlength{\hsize}{\textwidth}
 \begin{center}
-\includegraphics[width=0.95\columnwidth]{./images/exp_crn_figure.pdf}
-\caption{\small{A view of a simulator in terms of common random numbers.
-The horizontal line represents $\y$ and red shading $\pm 2\eps$.
-The shaded curved region represents $2\sigma$ of $\pi(\x|\thetav)$.
-The dashed lines are $f(\thetav, \omega_s)$ for several values of $\omega$.
-The blue circles are potential random samples from $\pi(\x|\thetav)$.
-For a fixed value $\omega_s$, the simulator produces deterministic outputs that change smoothly, even though the simulator itself is quite noisy.}}
-\label{fig:exp-crns}
+\includegraphics[width=0.95\columnwidth]{./images/exp-ABC-MCMC-theta-timeseries.pdf}
+\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries.pdf}
+\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Langevin-theta-timeseries.pdf}
+\includegraphics[width=0.95\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries.pdf}
+\caption{\small{}}
+\label{fig:exp-theta-traces}
 \end{center}
 \vskip -0.2in
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure*}[t]
-\vskip 0.2in
-\setlength{\linewidth}{\textwidth}
-\setlength{\hsize}{\textwidth}
-\begin{center}
-\includegraphics[width=1.0\columnwidth]{./images/exp-ABC-MCMC-theta-timeseries.pdf}
-\includegraphics[width=1.0\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries.pdf}
-\includegraphics[width=1.0\columnwidth]{./images/exp-SG-Langevin-theta-timeseries.pdf}
-\includegraphics[width=1.0\columnwidth]{./images/exp-SG-Thermostats-theta-timeseries.pdf}
-\caption{\small{}}
-\label{fig:exp-theta-traces}
-\end{center}
-\vskip -0.2in
-\end{figure*}
-
-\begin{figure}[t]
-\vskip 0.2in
-\begin{center}
-\includegraphics[width=0.45\columnwidth]{./images/exp-ABC-MCMC-posterior_hist.pdf}
-\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
-\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Langevin-posterior_hist.pdf}
-\includegraphics[width=0.45\columnwidth]{./images/exp-SG-Thermostats-posterior_hist.pdf}
-\caption{\small{}}
-\label{fig:exp-posteriors}
-\end{center}
-\vskip -0.2in
-\end{figure}
-
 \subsection{Sticky Random Numbers}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+We present experimental results comparing HABC with standard ABC-MCMC for two challenging simulators.
+The first is low-dimensional ($D=6$), but exhibits chaotic behavior.
+For the second problem we apply the techniques of HABC to a Bayesian image classifier and autoencoder.
+The benefit of this problem is we can compare directly with the SGHD algorithms using the analytic gradients.
+This problem is has very high-dimension ($D>15000$), yet we are able to successfully apply SPSA-based gradients to HABC.
 
 \subsection{Blowfly}\label{sec:bf}
-[FROM UAI] Adult blowfly populations exhibit dynamic behavior for which several competing population models exist.
-In this experiment, we use observational data and a simulation model from Wood \cite{wood2010statistical}, based on their improvement upon previous population dynamics theory.
-Population dynamics are modeled using (discretized) differential equations that can produce chaotic behavior for some parameter settings.
-An example blowfly replicate series is shown in Figure~\ref{fig:blowflyexamples}, along with  times-series generated by a sample from $\p(\thetav|\y)$ using GPS-ABC.
-
-In \cite{wood2010statistical} there are several explanations of the population dynamics, corresponding to different simulations and parameters.
-We concentrate on the equation (1) in section 1.2.3 of the supplementary information, considered ``a better alternative model" by the author.
-The population dynamics equation generates  $N_1, \ldots, N_T$ using the following update rule:
+For these experiments, a simulator of adult sheep blowfly populations \cite{wood2010statistical} is used with statistics set to those from \cite{Meeds2014GpsUai}.
+The observational vector $\y$ is a time-series of a fly population counted daily.
+The population dynamics are modeled using a stochastic differential equation\footnote{Equation~1 in Section 1.2.3 of the supplementary information in \cite{wood2010statistical}.}
 \begin{equation}
@@
  \sim  \mathcal{G}( 1/{\sigma_d^2},1/{\sigma_d^2})$  
-are sources of noise, and $\tau$ is an integer (not to be confused with the $\tau$ used as the MH acceptance threshold in our algorithms).
-In total, there are 6 parameters $\theta = \{ \log P, \log \delta, \log N_0, \log \sigma_d, \log \sigma_p, \tau\}$.
-See \cite{wood2010statistical} for further details about the significance of the parameters.
-We put Gaussian priors over all the parameters (with Gaussian proposal distributions), except for $\tau$ which has a Poisson prior (and a left/right increment proposal).
-Time-series generated with parameters from this prior distribution produce extremely varied results, some are chaotic, some are degenerate, etc.
-Modeling this simulator is {\em very} challenging.
+are sources of noise, and $\tau$ is an integer.
+In total, there are $D=6$ parameters $\theta = \{ \log P, \log \delta, \log N_0, \log \sigma_d, \log \sigma_p, \tau\}$.
+As \cite{Meeds2014GpsUai} we place broad log-normal priors over $\theta_{1\ldots 5}$ and a Poisson prior over $\tau$.
+This is considered a challenging problem because slight changes to some parameter settings can produce degenerate $\x$, while others settings can be very noisy due to the chaotic nature of the equations.
+The statistics from \cite{Meeds2014GpsUai} are used ($J=10$): the log average of $4$ quantiles of $N/1000$, the average of $4$ quantiles of the first-order differences in $N/1000$, and the number of maximal population peaks under two different thresholds.
 
-As with any ABC problem the choice of statistics is important as it directly affects the quality of the results.
-It is also non-trivial and requires careful thought and sometimes trial and error.
-In total there are 10 statistics: the log of the mean of all $25\%$ quantiles of $N/1000$ (4 statistics), the mean of the $25\%$ quantiles of the first-order differences of $N/1000$ (4 statistics), and the maximal peaks of smoothed $N$, with 2 different thresholds (2 statistics).
-With these statistics it is possible to reproduce time-series that appear similar to the observations.
-Note that these statistics are different from Wood's, but they capture similar time-series features and are sufficient to produce credible population dynamics.
-[END FROM UAI]
+%\subsubsection{Posterior Inference using HABC}
+We compare difference HABC algorithms with ABC-MCMC for the blowfly population problem.
+We use $\epsvec = \{ 1/2,1/2,1/2,1/2, 1/4,1/4,1/4,1/4,3/4,3/4 \}$ (slightly different $\epsvec$ from \cite{Meeds2014GpsUai}) and $S=10$ for all experiments.
+We use SPSA with $R=1$ using SL log-likelihoods for all HABC gradient estimates.
+Note that this requires $2S$ forward simulations per time-step, the same as marginal ABC and twice pseudo-marginal ABC.
+
+Figure~\ref{fig:bf-posterior-histograms} show the posterior distributions for $\log P, \log \delta$ and the posterior predictive distributions of [FILL IN].
+Traces of $\thetav$ are shown for SGHMC and SGNHT in Figure~\ref{}.
+Finally, we compare the convergence to $\y$ using the online posterior predictive distribution in Figure~\ref{}.
+By using stick random numbers, the convergence is [FILL IN].
 

### 2015-02-28 code/blowfly_problem/blowfly_runs.py

@@ -201,7 +201,7 @@ def view_posterior( problem, theta_range, samples, algoname, burnin = 1000 ):
   
 keep_x        = True 
 init_seed     = 4
-T             = 50000 # nbr of samples
+T             = 2000 # nbr of samples
 verbose_rate  = 10
 C             = 10.01    # injected noise variance parameter
 eta           = 0.01 # step size for Hamiltoniam dynamics
@@ -209,7 +209,7 @@ eta           = 0.01 # step size for Hamiltoniam dynamics
 
 # params for gradients
 d_theta = 0.5  # step size for gradient estimate
-S       = 5 
+S       = 10 
 grad_params = {}
   
 
@@ -218,7 +218,7 @@ grad_params = {}
 # common ransom seeds     #
 # ----------------------- #
 use_omega    = True    # use fixed random seeds for simulations
-omega_rate   = 0.01    # probabilty of changing omegas
+omega_rate   = 0.1    # probabilty of changing omegas
 omega_switch = True    # if true, randomly change omegas
 omega_sample = False   # sample omegas instead
 assert (omega_switch * omega_sample) ==0, "only get to do one type of omega update"
@@ -286,16 +286,16 @@ if __name__ == "__main__":
   np.random.seed(init_seed + 1000*chain_id)
   
   # run algorithm
-  # np.random.seed(init_seed + 1000*chain_id)
-  # sgnht = run_thermostats( problem, params, theta0, x0 )
-  # algoname = "SG-Thermostats"
-  # # view_posterior( problem, theta_range, sgnht["THETA"], algoname, burnin=1000 )
-  # # # if saveit:
-  # # #   pp.savefig("bf-%s-posterior_hist.pdf"%(algoname), format="pdf", dpi=600,bbox_inches="tight")
-  # # #   pp.savefig("../../papers/uai-2015/images/bf-%s-posterior_hist.pdf"%(algoname), format="pdf",dpi=600,bbox_inches="tight")
-  # view_theta_timeseries( problem, theta_range, sgnht["THETA"], algoname, howmany = 1000 )
-  # view_theta_hist( problem, theta_range, sgnht["THETA"], algoname, howmany = 1000 )
-  # view_X_timeseries( problem, sgnht["X"], algoname, howmany = 1000 )
[…]

### 2015-02-28 code/blowfly_problem/blowfly_wrapper.py

@@ -10,7 +10,8 @@ from abc_sgld.code.working_code import *
 from scipy import stats as spstats
 import pylab as pp
 
-default_epsilons = 0.5 + 0*np.array([ 0.08257003,  0.01013485,  0.01,  0.01, 0.01 , 0.01,  0.01,0.01,0.53027013,0.13338243] )
+#default_epsilons = 0.5 + 0*np.array([ 0.08257003,  0.01013485,  0.01,  0.01, 0.01 , 0.01,  0.01,0.01,0.53027013,0.13338243] )
+default_epsilons = np.array([0.5,0.5,0.5,0.5,0.25,0.25,0.25,0.25,0.75,0.75])
 
 # exponential distributed observations with Gamma(alpha,beta) prior over lambda
 problem_params = default_params()

### 2015-03-01 code/autoencoder_problem/nn.py
autoencoder

@@ -1,0 +1,165 @@
+import numpy as np
+import json
+import gzip
+import cPickle
+from time import time
+import matplotlib.pyplot as plt
+
+def load_mnist():
+    f = gzip.open('../../data/mnist.pkl.gz', 'rb')
+    data = cPickle.load(f)
+    f.close()
+    return data
+
+def sigmoid(x):
+    return 1.0 / (1.0 + np.exp(-x))
+
+def sigmoid_grad(x):
+    ex = np.exp(-x)
+    return ex / (1 + ex)**2
+
+def one_hot(j):
+    e = np.zeros((10, 1))
+    e[j] = 1.0
+    return e
+
+def load_sets_mnist():
+    (x_train, t_train), (x_valid, t_valid), (x_test, t_test) = load_mnist()
+    # Transform the data into something easier to work with (imo)
+    training_inputs = [np.reshape(x, (784, 1)) for x in x_train]
+    training_results = [one_hot(y) for y in t_train]
+    training_set = zip(training_inputs, training_inputs)
+
+    validation_inputs = [np.reshape(x, (784, 1)) for x in x_valid]
+    valiation_results = [one_hot(y) for y in t_valid]
+    validation_set = zip(validation_inputs, validation_inputs)
+
+    test_inputs = [np.reshape(x, (784, 1)) for x in x_test]
+    test_set = zip(test_inputs, test_inputs)
+
+    return training_set, validation_set, test_set
+    # v--- really slow
+    # return np.array(training_set), np.array(validation_set), np.array(test_set)
+
+def plot_digits(data, numcols, shape=(28,28)):
+    numdigits = data.shape[0]
+    numrows = int(numdigits/numcols)
+    for i in range(numdigits):
+        plt.subplot(numrows, numcols, i)
+        plt.axis('off')
+        plt.imshow(data[i].reshape(shape), interpolation='nearest', cmap='Greys')
+    plt.show()
+
+class NeuralNetwork(object):
+    def __init__(self, config):
+        # config is a list with the number of neurons per layer, so len(config) equals the number of layers.
+        self.config = config
+        self.num_layers = len(config)
+        # Initialize the biases
+        self.biases = np.array([np.random.randn(layer_size, 1) for layer_size in self.config[1:]])
+        # Initialize the weights
+        self.weights = []
+        for i in range(self.num_layers-1):
+            layer_size = self.config[i]
+            next_layer_size = self.config[i+1]
[…]

### 2015-03-01 code/autoencoder_problem/nn_wrapper.py

@@ -1,0 +1,127 @@
+import numpy as np
+# Theta=weights
+class NeuralNetworkProblem(object):
+    def __init__(self, nn, data):
+        self.nn = nn
+        self.training_set, self.validation_set, self.test_set = data
+
+    # Skip
+    def prior_rand(self):
+        pass
+
+    # Skip
+    def propose(self, theta):
+        pass
+
+    def loglike_x(self, X, omega, simulated=True):
+        Y = []
+        for i in omega:
+            Y.append(self.training_set[i][1])
+        # Y = self.training_set[omega, 1]
+        assert len(X) == len(Y)
+        likelihood = 0.0
+        for i in range(len(X)):
+            x = X[i]
+            y = Y[i]
+            if simulated:
+                x[x < 0.5] = 0
+                x[x >= 0.5] = 1
+                y_copy = y.copy()
+                y_copy[y_copy <= 0.5] = 0
+                y_copy[y_copy > 0.5] = 1
+                # (??) Take log?
+                likelihood += np.sum(x == y_copy) / float(len(y_copy))
+            else:
+                # We are given a sample so we don't need to do feedforward
+                likelihood += np.nan_to_num(np.sum(-y*np.log(x+10**-10)-(1-y)*np.log(1-x+10**-10)))
+
+        return likelihood / len(X)
+
+    # Log prior for some theta
+    def loglike_prior(self, theta):
+        pass
+
+    # Log posterior for some theta
+    def loglike_posterior(self, theta):
+        pass
+
+    # Skip
+    def loglike_proposal_theta(self, to_theta, from_theta):
+        pass
+
+    def simulate(self, theta, seed = None, S = 1):
+        samples = []
+        for batch_id in seed:
+            x = self.training_set[batch_id][0]
+            output = self.forwardpass(theta, x)
+            u = np.random.uniform(0, 1, output.shape)
+            larger = u > output
+            smaller = u < output
+            output[larger] = 1
+            output[smaller] = 0
+            samples.append(output)
+        return np.array(samples)
+
+    # Use backprop
+    def true_gradient(self, theta, omega):
+        self.nn.weights = theta
+        grad_b = 0*self.nn.biases
+        grad_w = 0*theta
+        for i in omega:
+            x, y = self.training_set[i]
+            gb, gw = self.nn.backpropagation(x, y)
+            grad_b += gb
+            grad_w += gw
+        # Return mean
+        return grad_w / len(omega)
+
+    # Does NN have this?
[…]

### 2015-03-01 code/autoencoder_problem/run.py

@@ -1,0 +1,72 @@
+# TOOD: Adapt this for NN case
+# Construct NN, load data
+# Pass it to nn_wrapper
+
+import numpy as np
+import pylab as pp
+import scipy as sp
+from scipy import stats as spstats
+from nn_wrapper import *
+from working_code import *
+from nn import *
+
+keep_x        = True
+init_seed     = 1
+T             = 5000 # nbr of samples
+verbose_rate  = 1000
+C             = 20.01    # injected noise variance parameter
+eta           = 0.001 # step size for Hamiltoniam dynamics
+#h = 0.0005
+
+# params for gradients
+d_theta = 0.001  # step size for gradient estimate
+S       = 5
+grad_params = {}
+
+# keep theta within bounds
+lower_bounds = np.array([0.001])
+upper_bounds = np.array([np.inf])
+
+if __name__ == "__main__":
+  # pp.close('all')
+  problem = NeuralNetworkProblem( NeuralNetwork([28*28, 100, 28*28]), load_sets_mnist() )
+  # Cheat by loading previous training NN
+  # problem.nn.load('latest_epoch.json')
+  chain_id = 1
+  params = {}
+  params["chain_id"]  = chain_id
+  params["init_seed"] = init_seed
+  params["T"]       = T
+  params["S"]       = S
+  params["d_theta"] = d_theta
+  params["eta"]     = eta
+  params["C"]       = C
+  params["batch_size"] = 100
+  params["verbose_rate"] = verbose_rate
+  params["grad_params"]  = {"logs":{"true":[],"true_abc":[],"2side_keps":[],"2side_sl":[]},\
+                            "record_2side_sl_grad":False, "record_2side_keps_grad":False,"record_true_abc_grad":False,"record_true_grad":False,
+                            "2side_keps": {"R": 100}}
+  params["lower_bounds"] = lower_bounds
+  params["upper_bounds"] = upper_bounds
+  params["keep_x"]       = keep_x
+
+  theta0 = problem.nn.weights
+  x0     = None
+
+  # initialize the simulator
+  np.random.seed(init_seed)
+  params["grad_func"] = problem.two_sided_keps_gradient
+
+  # init randomly for MCMC chain
+  np.random.seed(init_seed + 1000*chain_id)
+
+  # run algorithm
+  # outs = run_thermostats( problem, params, theta0, x0 )
+  #outs = run_sghmc( problem, params, theta0, x0 )
[…]

### 2015-03-01 code/autoencoder_problem/working_code.py

@@ -1,0 +1,389 @@
+import numpy as np
+import scipy as sp
+from scipy import stats as spstats
+import pylab as pp
+import pdb
+
+def get_omega(problem, batch_size):
+  get_omega.counter += 1
+  mini_batches = get_omega.mini_batches
+  if not mini_batches:
+    reconstruction_error, num_correct = problem.nn.test(problem.validation_set)
+    print "Reconstruction error: {}, average percentage of pixels correct: {}%".format(reconstruction_error, num_correct*100)
+    N = len(problem.training_set)
+    perms = np.random.permutation(N)
+    mini_batches += [perms[k:k+batch_size] for k in range(0, N, batch_size)]
+  if get_omega.counter == 10:
+    get_omega.counter = 0
+    reconstruction_error, num_correct = problem.nn.test(problem.validation_set)
+    print "Reconstruction error: {}, average percentage of pixels correct: {}%".format(reconstruction_error, num_correct*100)
+  return mini_batches.pop()
+get_omega.mini_batches = []
+get_omega.counter = 0
+
+def bounce_off_boundaries( theta, p, lower_bounds, upper_bounds ):
+  D = len(theta)
+
+  for d in range(D):
+    if lower_bounds is not None:
+      if theta[d] < lower_bounds[d]:
+        theta[d] = lower_bounds[d] + (lower_bounds[d]-theta[d]) #from Neal Handbook, p37
+        p[d] *= -1
+
+    if upper_bounds is not None:
+      if theta[d] > upper_bounds[d]:
+        theta[d] = upper_bounds[d] - (theta[d]-upper_bounds[d]) #from Neal Handbook, p37
+        p[d] *= -1
+  return theta, p
+
+
+def run_sgld( problem, params, theta, x = None ):
+
+  T             = params["T"]
+  S             = params["S"]
+  verbose_rate  = params["verbose_rate"]
+  batch_size    = params["batch_size"]
+  eta           = params["eta"] # Hamiltonain step size
+  d_theta       = params["d_theta"] # step used for estimating gradients
+  grad_U_func   = params["grad_func"]
+  grad_U_params = params["grad_params"]
+
+  keep_x        = params["keep_x"] # the HABC methods do not necessarily have x in the state space
+
+  # keep theta within bounds
+  lower_bounds = params["lower_bounds"]
[…]

### 2015-03-01 code/blowfly_problem/blowfly_convergence_plots.py
exp bias checking

@@ -1,0 +1,229 @@
+from abcpy.problems.blowfly.blowfly    import BlowflyProblem   as Problem
+from abcpy.problems.blowfly.blowfly    import default_params   as load_default_params
+from abcpy.plotting import *
+
+import pdb
+import numpy as np
+import pylab as pp
+
+def quick_convergence_single_algo( y, X, intervals ):
+  C = np.zeros( (len(y),len(intervals)))
+  
+  convs = []
+  count = 0.0
+  mean  = np.zeros(len(y))
+  
+  j = 0
+  for i in xrange(len(X)):
+    mean = count*mean + X[i]
+    count+=1
+    mean /= count
+    
+    convs.append( pow( y - mean, 2 )/pow(y_star,2) )
+    
+  convs = np.array(convs)
+  C = convs[intervals,:]
+  return C
+
+def compute_convergence( p, y_star, th, sm, sts, burnin, intervals, ignore_stats = False ):
+  t = th[burnin:,:]
+  s = sm[burnin:]
+  
+  have_stats = False
+  if len(sts) == len(th) and ignore_stats == False:
+    sts = sts[burnin:,:]
+    have_stats = True
+    print "Have stats!"
+  else:
+    print "Bummer, no stats!"
+    
+  cs = np.cumsum(s)
+  
+  N = len(t)
+  
+  idx = 0
+  times = []
+  convs = []
+  sims = []
+  statss = []
+  
+  count = 0.0
+  mean  = np.zeros(len(y_star))
+  while idx < N:
+    
+    theta = t[idx]
+    
+    if have_stats:
+      stats = sts[idx,:]
+    else:
+      outs = p.simulation_function(theta)
+      stats = p.statistics_function(outs)
+    
+    mean = count*mean + stats
+    count+=1
+    mean /= count
+    
+    convs.append( pow( y_star - mean, 2 )/pow(y_star,2) )
+    sims.append( cs[idx] )
+    times.append( idx)
+    statss.append(stats)
+    
+    idx += 1
+  
+  convs = np.array(convs)
+  sims  = np.array(sims)
+  times = np.array(times)
+  statss = np.array(statss)
+  
+  ids = []
[…]

### 2015-03-01 code/blowfly_problem/blowfly_runs.py

@@ -6,7 +6,46 @@ from blowfly_wrapper import *
 from habc_algos import *
 
 #def pp_vector( x ):
+def view_all_convs( problem, algo_convs, intervals, stats_2_use = [0,1,2,3,4,5,6,7,8]):
+  figsize = (16,6)
+  alpha=0.75
+  pp.rc('text', usetex=True)
+  pp.rc('font', family='times')
+  f = pp.figure( figsize=figsize )
+  sp=pp.subplot(1,1,1)
   
+  
+  for j in stats_2_use:
+    pp.subplot( 3,3,j+1)
+    
+    leg_names = []
+    for name, convs in algo_convs.iteritems():
+      pp.loglog( intervals, convs[:,j], lw=2 )
+      leg_names.append(name)
+    pp.title( problem.p.stats_names[j] )
+    pp.xlim( intervals[0],intervals[-1])
+  pp.legend( leg_names, loc=1,fancybox=True,prop={'size':16} )
+    
+  
+def quick_convergence_single_algo( y, X, intervals ):
+  C = np.zeros( (len(y),len(intervals)))
+  
+  convs = []
+  count = 0.0
+  mean  = np.zeros(len(y))
+  
+  j = 0
+  for i in xrange(len(X)):
+    mean = count*mean + X[i]
+    count+=1
+    mean /= count
+    
+    convs.append( pow( y - mean, 2 )/pow(y,2) )
+    
+  convs = np.array(convs)
+  C = convs[intervals,:]
+  return C
+    
 def view_results( problem, thetas, all_stats, burnin = 1 ):
   stats = all_stats.mean(1)
   self=problem
@@ -204,7 +243,7 @@ init_seed     = 4
 T             = 2000 # nbr of samples
 verbose_rate  = 10
 C             = 10.01    # injected noise variance parameter
-eta           = 0.01 # step size for Hamiltoniam dynamics
+eta           = 0.1 # step size for Hamiltoniam dynamics
 #h = 0.0005
 
 # params for gradients
@@ -282,30 +321,30 @@ if __name__ == "__main__":
   params["grad_func"] = problem.two_sided_sl_gradient
   #params["grad_func"] = problem.two_sided_keps_gradient
 
-  # init randomly for MCMC chain
-  np.random.seed(init_seed + 1000*chain_id)
-  
-  # run algorithm
-  np.random.seed(init_seed + 1000*chain_id)
[…]

### 2015-03-01 code/blowfly_problem/blowfly_wrapper.py

@@ -116,12 +116,12 @@ class generate_blowfly( object ):
       if seed is None:
         state = np.random.get_state()
       
-      X_plus[s] = self.simulate( theta_plus, seed=[seed] )
+      X_plus = self.simulate( theta_plus, omega, S )
       
       if seed is None:
         np.random.set_state(state)
         
-      X_minus[s] = self.simulate( theta_minus, seed=[seed] )
+      X_minus = self.simulate( theta_minus, omega, S )
       
       seeds.append(seed)
       

### 2015-03-01 code/blowfly_problem/habc_algos.py

@@ -306,7 +306,7 @@ def run_sgld( problem, params, theta, x = None ):
   OMEGAS = [omega]
   
   grads = np.zeros( (T,D) )
-  
+  V=0
   # sample...
   for t in xrange(T):
     
@@ -328,6 +328,11 @@ def run_sgld( problem, params, theta, x = None ):
     
     grads[t,:] = grad_U
     
+    if t > 60:
+    
+      V = np.var(grads[t-60:t+1],0)# *np.eye(D)
+
+      
     # initialize momentum  
     p = np.random.randn(D)
     
@@ -365,6 +370,12 @@ def run_sgld( problem, params, theta, x = None ):
     if np.mod(t+1,verbose_rate)==0:
       if keep_x:
         print "t = %04d    loglik = %3.3f    theta0 = %3.3f    x0 = %3.3f"%(t+1,loglike_x,theta[0],x[0][0])
+        grad_noise = ((0.5*eta*eta)**2)*V
+        injected_noise = eta**2
+        print "\t var ", V
+        print "\t grad noise ", grad_noise
+        print "\t inject noise", injected_noise
+        print "\t ratio ", grad_noise / injected_noise
       else:
         print "t = %04d    theta0 = %3.3f"%(t+1,theta[0])
       

### 2015-03-01 code/exponential_problem/exp_wrapper.py

@@ -7,14 +7,14 @@ import pylab as pp
 
 problem_params = default_params()
 
-problem_params["N"]               = 20
+problem_params["N"]               = 5
 problem_params["q_stddev"]        = 0.2
 problem_params["theta_star"]      = 0.15
 problem_params["epsilon"]         = 0.25*np.sqrt( 1.0 / (problem_params["N"]*problem_params["theta_star"]**2) )
 problem_params["alpha"]           = 0.1
 problem_params["beta"]            = 0.1
 problem_params["min_range"]       = 0.005
-problem_params["max_range"]       = 0.3
+problem_params["max_range"]       = 0.6
 #problem_params["alpha"]           = 3.75
 #problem_params["beta"]            = 1.01
 
@@ -67,20 +67,24 @@ class generate_exponential( object ):
     
   def simulate( self, theta, seed = None, S = 1 ):
     X = np.zeros( (S,1))  
-    
+
+    #if seed is not None:
+    #  # save current state
+    current_state = np.random.get_state()
+      
     for s in range(S):
       if seed is not None:
-        # save current state
-        current_state = np.random.get_state()
         np.random.seed(seed[s])
-    
+        #print "setting seed ",seed[s]
+      #else:
+      #  np.random.set_state( current_state )
     
       raw_outputs = self.p.simulation_function( theta )
       X[s] = self.p.statistics_function( raw_outputs)
       
-      if seed is not None:
-        # put back the current state
-        np.random.set_state( current_state )
+    if seed is not None:
+      # put back the current state
+      np.random.set_state( current_state )
     
     return X
   
@@ -159,23 +163,24 @@ class generate_exponential( object ):
     seeds   = []
     X_plus  = np.zeros( (S,1))
     X_minus = np.zeros( (S,1))
+    theta_minus = max(0.01,theta-d_theta)
+    theta_plus = theta_minus + 2*d_theta
     for s in range(S):
-      theta_minus = max(0.01,theta-d_theta)
-      theta_plus = theta_minus + 2*d_theta
       
       if omega is not None:
         seed = omega[s]
        
       if seed is None:
         state = np.random.get_state()
-      
-      X_plus[s] = self.simulate( theta_plus, seed=[seed] )[0]
-      
+      #pdb.set_trace()
[…]

### 2015-03-01 code/exponential_problem/figs_runs_posteriors.py

@@ -5,6 +5,25 @@ from scipy import stats as spstats
 from exp_wrapper import *
 from abc_sgld.code.working_code import *
 
+def total_variational_distance_errors( problem, theta, times ):
+  self=problem
+  errs = []
+  time_ids = []
+  nbr_sims = []
+  
+  for time_id in times:
+    if time_id <= len(theta):
+      #errs.append( bin_errors_1d(self.p.coarse_theta_range, self.p.posterior_cdf_bins, thetas[:time_id]) )
+      er = bin_errors_1d( self.p.coarse_theta_range, self.p.posterior_cdf_bins, theta[:time_id] )
+      
+      errs.append( er  )
+      time_ids.append(time_id)
+      
+  errs = np.array(errs)
+  time_ids = np.array(time_ids)
+  return errs, time_ids
+  
+    
 def view_theta_timeseries( problem, theta_range, samples, algoname, howmany = 1000 ):
   figsize = (16,6)
   alpha=0.75
@@ -65,18 +84,19 @@ def view_posterior( problem, theta_range, samples, algoname, burnin = 1000 ):
   set_label_fonsize( sp, 24 )
   
 
+burnin = 100
 
 keep_x        = True 
 init_seed     = 4
-T             = 50000 # nbr of samples
-verbose_rate  = 1000
-C             = 1.01    # injected noise variance parameter
-eta           = 0.01 # step size for Hamiltoniam dynamics
+T             = 10000 + burnin # nbr of samples
+verbose_rate  = 100
+C             = 50.01    # injected noise variance parameter
+eta           = 0.005 # step size for Hamiltoniam dynamics
 #h = 0.0005
 
 # params for gradients
-d_theta = 0.001  # step size for gradient estimate
-S       = 5 
+d_theta = 0.01  # step size for gradient estimate
+S       = 5
 grad_params = {}
   
 # keep theta within bounds
@@ -88,7 +108,7 @@ upper_bounds = np.array([np.inf])
 # common ransom seeds     #
 # ----------------------- #
 use_omega    = True    # use fixed random seeds for simulations
-omega_rate   = 0.01    # probabilty of changing omegas
+omega_rate   = 0.001    # probabilty of changing omegas
 omega_switch = True    # if true, randomly change omegas
[…]

### 2015-03-01 code/working_code.py

@@ -272,7 +272,7 @@ def run_sgld( problem, params, theta, x = None ):
   d_theta       = params["d_theta"] # step used for estimating gradients
   grad_U_func   = params["grad_func"]
   grad_U_params = params["grad_params"]
-  
+  mh_correction = params["mh_correction"]
   keep_x        = params["keep_x"] # the HABC methods do not necessarily have x in the state space
   
   # keep theta within bounds
@@ -317,22 +317,25 @@ def run_sgld( problem, params, theta, x = None ):
     # estimate stochastic gradient
     grad_U = grad_U_func( theta, d_theta, omega, S, grad_U_params)
     #
-    if grad_U_params["record_2side_sl_grad"]:
-      grad_U_dummy = problem.two_sided_sl_gradient( theta, d_theta, omega, S, grad_U_params )
-    if grad_U_params["record_2side_keps_grad"]:
-      grad_U_dummy = problem.two_sided_keps_gradient( theta, d_theta, omega, S, grad_U_params )
-    if grad_U_params["record_true_abc_grad"]:
-      grad_U_dummy = problem.true_abc_gradient( theta, d_theta, S, grad_U_params )
-    if grad_U_params["record_true_grad"]:
-      grad_U_dummy = problem.true_gradient( theta, grad_U_params )
+    # if grad_U_params["record_2side_sl_grad"]:
+    #   grad_U_dummy = problem.two_sided_sl_gradient( theta, d_theta, omega, S, grad_U_params )
+    # if grad_U_params["record_2side_keps_grad"]:
+    #   grad_U_dummy = problem.two_sided_keps_gradient( theta, d_theta, omega, S, grad_U_params )
+    # if grad_U_params["record_true_abc_grad"]:
+    #   grad_U_dummy = problem.true_abc_gradient( theta, d_theta, S, grad_U_params )
+    # if grad_U_params["record_true_grad"]:
+    #   grad_U_dummy = problem.true_gradient( theta, grad_U_params )
     
     grads[t,:] = grad_U
     
     # initialize momentum  
     p = np.random.randn(D)
+    current_p = p
     
     # take step momentum
     p = p - grad_U*eta/2.0
+    
+    current_theta = theta
     
     # full step position
     theta = theta + p*eta
@@ -340,6 +343,12 @@ def run_sgld( problem, params, theta, x = None ):
     # bounce position off parameter boundaries
     theta, p = bounce_off_boundaries( theta, p, lower_bounds, upper_bounds )
     
[…]

### 2015-03-01 code/autoencoder_problem_ted/nn.py
ted autoencoder fixes

@@ -1,0 +1,169 @@
+import numpy as np
+import json
+import gzip
+import cPickle
+from time import time
+import matplotlib.pyplot as plt
+import pdb
+
+def load_mnist():
+    f = gzip.open('../../data/mnist.pkl.gz', 'rb')
+    data = cPickle.load(f)
+    f.close()
+    return data
+
+def sigmoid(x):
+    return 1.0 / (1.0 + np.exp(-x))
+
+def sigmoid_grad(x):
+    ex = np.exp(-x)
+    return ex / (1 + ex)**2
+
+def one_hot(j):
+    e = np.zeros((10, 1))
+    e[j] = 1.0
+    return e
+
+def load_sets_mnist():
+  
+    (x_train, t_train), (x_valid, t_valid), (x_test, t_test) = load_mnist()
+    # Transform the data into something easier to work with (imo)
+    training_inputs = [np.reshape(x, (784, 1)) for x in x_train]
+    training_results = [one_hot(y) for y in t_train]
+    training_set = zip(training_inputs, training_inputs)
+
+    validation_inputs = [np.reshape(x, (784, 1)) for x in x_valid]
+    valiation_results = [one_hot(y) for y in t_valid]
+    validation_set = zip(validation_inputs, validation_inputs)
+
+    test_inputs = [np.reshape(x, (784, 1)) for x in x_test]
+    test_set = zip(test_inputs, test_inputs)
+
+    return training_set, validation_set, test_set
+    # v--- really slow
+    # return np.array(training_set), np.array(validation_set), np.array(test_set)
+
+def plot_digits(data, numcols, shape=(28,28)):
+    numdigits = data.shape[0]
+    numrows = int(numdigits/numcols)
+    for i in range(numdigits):
+        plt.subplot(numrows, numcols, i)
+        plt.axis('off')
+        plt.imshow(data[i].reshape(shape), interpolation='nearest', cmap='Greys')
+    plt.show()
+
+class NeuralNetwork(object):
+    def __init__(self, config):
+        # config is a list with the number of neurons per layer, so len(config) equals the number of layers.
+        self.config = config
+        self.num_layers = len(config)
+        # Initialize the biases
+        self.biases = np.array([0.001*np.random.randn(layer_size, 1) for layer_size in self.config[1:]])
+        # Initialize the weights
+        self.weights = []
+        for i in range(self.num_layers-1):
+            layer_size = self.config[i]
[…]

### 2015-03-01 code/autoencoder_problem_ted/nn_wrapper.py

@@ -1,0 +1,144 @@
+import numpy as np
+# Theta=weights
+class NeuralNetworkProblem(object):
+    def __init__(self, nn, data):
+        self.nn = nn
+        (self.X_train, self.T_train), (self.X_valid, self.T_valid), (self.X_test, self.T_test) = data
+        self.N = len(self.X_train)
+        #self.training_set, self.validation_set, self.test_set = data
+        self.prior_penalty = 0.1
+    # Skip
+    def prior_rand(self):
+        pass
+
+    # Skip
+    def propose(self, theta):
+        pass
+
+    def loglike_x(self, X, omega, simulated=True):
+        Y = self.X_train[omega,:].T
+        # Y = []
+        # for i in omega:
+        #     Y.append(self.training_set[i][1])
+        # # Y = self.training_set[omega, 1]
+        # assert len(X) == len(Y)
+        # likelihood = 0.0
+        # for i in range(len(X)):
+        #     x = X[i]
+        #     y = Y[i]
+        #     if simulated:
+        #         x[x < 0.5] = 0
+        #         x[x >= 0.5] = 1
+        #         y_copy = y.copy()
+        #         y_copy[y_copy <= 0.5] = 0
+        #         y_copy[y_copy > 0.5] = 1
+        #         # (??) Take log?
+        #         likelihood += np.sum(x == y_copy) / float(len(y_copy))
+        #     else:
+        #         # We are given a sample so we don't need to do feedforward
+        #         likelihood += np.nan_to_num(np.sum(-y*np.log(x+10**-10)-(1-y)*np.log(1-x+10**-10)))
+
+        n = len(omega)
+        likelihood_batch = Y*np.log( X + 1e-10 ) + (1-Y)*np.log(1 - X + 1e-10)
+        likelihood = self.N*likelihood_batch.sum()/n
+        return likelihood
+
+    # Log prior for some theta
+    def loglike_prior(self, theta):
+        pass
+
+    # Log posterior for some theta
+    def loglike_posterior(self, theta):
+        pass
+
+    # Skip
+    def loglike_proposal_theta(self, to_theta, from_theta):
+        pass
+
+    def simulate(self, theta, seed = None, S = 1):
+        self.set_weights( theta )
+        samples = self.forwardpass( self.X_train[seed,:])
+        # samples = []
+        # for batch_id in seed:
+        #     x = self.training_set[batch_id][0]
+        #     output = self.forwardpass(theta, x)
[…]

### 2015-03-01 code/autoencoder_problem_ted/run.py

@@ -1,0 +1,72 @@
+# TOOD: Adapt this for NN case
+# Construct NN, load data
+# Pass it to nn_wrapper
+
+import numpy as np
+import pylab as pp
+import scipy as sp
+from scipy import stats as spstats
+from nn_wrapper import *
+from working_code import *
+from nn import *
+
+keep_x        = True
+init_seed     = 1
+T             = 5000 # nbr of samples
+verbose_rate  = 1000
+C             = 20.01    # injected noise variance parameter
+eta           = 0.001 # step size for Hamiltoniam dynamics
+#h = 0.0005
+
+# params for gradients
+d_theta = 0.001  # step size for gradient estimate
+S       = 5
+grad_params = {}
+
+# keep theta within bounds
+lower_bounds = np.array([0.001])
+upper_bounds = np.array([np.inf])
+
+if __name__ == "__main__":
+  # pp.close('all')
+  problem = NeuralNetworkProblem( NeuralNetwork([28*28, 10, 28*28]), load_mnist() )
+  # Cheat by loading previous training NN
+  # problem.nn.load('latest_epoch.json')
+  chain_id = 1
+  params = {}
+  params["chain_id"]  = chain_id
+  params["init_seed"] = init_seed
+  params["T"]       = T
+  params["S"]       = S
+  params["d_theta"] = d_theta
+  params["eta"]     = eta
+  params["C"]       = C
+  params["batch_size"] = 100
+  params["verbose_rate"] = verbose_rate
+  params["grad_params"]  = {"logs":{"true":[],"true_abc":[],"2side_keps":[],"2side_sl":[]},\
+                            "record_2side_sl_grad":False, "record_2side_keps_grad":False,"record_true_abc_grad":False,"record_true_grad":False,
+                            "2side_keps": {"R": 100}}
+  params["lower_bounds"] = lower_bounds
+  params["upper_bounds"] = upper_bounds
+  params["keep_x"]       = keep_x
+
+  theta0 = problem.nn.weights
+  x0     = None
+
+  # initialize the simulator
+  np.random.seed(init_seed)
+  params["grad_func"] = problem.two_sided_keps_gradient
+
+  # init randomly for MCMC chain
+  np.random.seed(init_seed + 1000*chain_id)
+
+  # run algorithm
+  # outs = run_thermostats( problem, params, theta0, x0 )
+  #outs = run_sghmc( problem, params, theta0, x0 )
[…]

### 2015-03-01 code/autoencoder_problem_ted/working_code.py

@@ -1,0 +1,389 @@
+import numpy as np
+import scipy as sp
+from scipy import stats as spstats
+import pylab as pp
+import pdb
+
+def get_omega(problem, batch_size):
+  get_omega.counter += 1
+  mini_batches = get_omega.mini_batches
+  if not mini_batches:
+    reconstruction_error, num_correct = problem.nn.test(problem.X_valid.T,problem.X_valid.T)
+    print "Reconstruction error: {}, average percentage of pixels correct: {}%".format(reconstruction_error, num_correct*100)
+    N = problem.N
+    perms = np.random.permutation(N)
+    mini_batches += [perms[k:k+batch_size] for k in range(0, N, batch_size)]
+  if get_omega.counter == 10:
+    get_omega.counter = 0
+    reconstruction_error, num_correct = problem.nn.test(problem.X_valid.T,problem.X_valid.T)
+    print "Reconstruction error: {}, average percentage of pixels correct: {}%".format(reconstruction_error, num_correct*100)
+  return mini_batches.pop()
+get_omega.mini_batches = []
+get_omega.counter = 0
+
+def bounce_off_boundaries( theta, p, lower_bounds, upper_bounds ):
+  D = len(theta)
+
+  for d in range(D):
+    if lower_bounds is not None:
+      if theta[d] < lower_bounds[d]:
+        theta[d] = lower_bounds[d] + (lower_bounds[d]-theta[d]) #from Neal Handbook, p37
+        p[d] *= -1
+
+    if upper_bounds is not None:
+      if theta[d] > upper_bounds[d]:
+        theta[d] = upper_bounds[d] - (theta[d]-upper_bounds[d]) #from Neal Handbook, p37
+        p[d] *= -1
+  return theta, p
+
+
+def run_sgld( problem, params, theta, x = None ):
+
+  T             = params["T"]
+  S             = params["S"]
+  verbose_rate  = params["verbose_rate"]
+  batch_size    = params["batch_size"]
+  eta           = params["eta"] # Hamiltonain step size
+  d_theta       = params["d_theta"] # step used for estimating gradients
+  grad_U_func   = params["grad_func"]
+  grad_U_params = params["grad_params"]
+
+  keep_x        = params["keep_x"] # the HABC methods do not necessarily have x in the state space
+
+  # keep theta within bounds
[…]

### 2015-03-01 code/exponential_problem/exp_wrapper.py

@@ -7,14 +7,14 @@ import pylab as pp
 
 problem_params = default_params()
 
-problem_params["N"]               = 5
+problem_params["N"]               = 20
 problem_params["q_stddev"]        = 0.2
 problem_params["theta_star"]      = 0.15
 problem_params["epsilon"]         = 0.25*np.sqrt( 1.0 / (problem_params["N"]*problem_params["theta_star"]**2) )
 problem_params["alpha"]           = 0.1
 problem_params["beta"]            = 0.1
 problem_params["min_range"]       = 0.005
-problem_params["max_range"]       = 0.6
+problem_params["max_range"]       = 0.3
 #problem_params["alpha"]           = 3.75
 #problem_params["beta"]            = 1.01
 
@@ -53,7 +53,11 @@ class generate_exponential( object ):
   
   def loglike_x( self, x ):
     S,J = x.shape
-    L = logsumexp( self.kernel.logpdf( x ), 0 ) - np.log(S)
+    mu = x.mean()
+    vr = x.var()
+    f = spstats.norm( mu, np.sqrt(vr+self.p.epsilon**2) )
+    L = f.logpdf(self.y)
+    #L = logsumexp( self.kernel.logpdf( x ), 0 ) - np.log(S)
     return L
     
   def loglike_prior( self, theta ):

### 2015-03-01 code/exponential_problem/figs_runs_posteriors.py

@@ -88,7 +88,7 @@ burnin = 100
 
 keep_x        = True 
 init_seed     = 4
-T             = 10000 + burnin # nbr of samples
+T             = 50000 + burnin # nbr of samples
 verbose_rate  = 100
 C             = 50.01    # injected noise variance parameter
 eta           = 0.005 # step size for Hamiltoniam dynamics
@@ -107,7 +107,7 @@ upper_bounds = np.array([np.inf])
 # ----------------------- #
 # common ransom seeds     #
 # ----------------------- #
-use_omega    = True    # use fixed random seeds for simulations
+use_omega    = False    # use fixed random seeds for simulations
 omega_rate   = 0.001    # probabilty of changing omegas
 omega_switch = True    # if true, randomly change omegas
 omega_sample = False   # sample omegas instead
@@ -119,7 +119,7 @@ omega_params = {"use_omega":use_omega, \
 
 if __name__ == "__main__": 
   pp.close('all')
-  saveit = False
+  saveit = True
   chain_id = 1
   params = {}
   params["chain_id"]  = chain_id
@@ -148,7 +148,7 @@ if __name__ == "__main__":
   problem = generate_exponential( exp_problem )
   
   seeds = np.arange(8)
-  theta_range = np.linspace( 0.01, 0.6, 200 )
+  theta_range = np.linspace( 0.01, 0.3, 200 )
   
   problem = generate_exponential( exp_problem )
   params["grad_func"] = problem.two_sided_sl_gradient
@@ -158,24 +158,24 @@ if __name__ == "__main__":
   np.random.seed(init_seed + 1000*chain_id)
   
   # run algorithm
-  np.random.seed(init_seed + 1000*chain_id)
-  sgnht = run_thermostats( problem, params, theta0, x0 )
-  algoname = "SG-Thermostats"
-  view_posterior( problem, theta_range, sgnht["THETA"], algoname, burnin=1000 )
-  if saveit:
-    pp.savefig("exp-%s-posterior_hist.pdf"%(algoname), format="pdf", dpi=600,bbox_inches="tight")
-    pp.savefig("../../papers/uai-2015/images/exp-%s-posterior_hist.pdf"%(algoname), format="pdf", dpi=600,bbox_inches="tight")
[…]

### 2015-03-01 code/autoencoder_problem_ted/run.py
small batches and smaller R

@@ -41,11 +41,11 @@ if __name__ == "__main__":
   params["d_theta"] = d_theta
   params["eta"]     = eta
   params["C"]       = C
-  params["batch_size"] = 100
+  params["batch_size"] = 10
   params["verbose_rate"] = verbose_rate
   params["grad_params"]  = {"logs":{"true":[],"true_abc":[],"2side_keps":[],"2side_sl":[]},\
                             "record_2side_sl_grad":False, "record_2side_keps_grad":False,"record_true_abc_grad":False,"record_true_grad":False,
-                            "2side_keps": {"R": 100}}
+                            "2side_keps": {"R": 10}}
   params["lower_bounds"] = lower_bounds
   params["upper_bounds"] = upper_bounds
   params["keep_x"]       = keep_x

### 2015-03-01 code/autoencoder_problem_ted/nn_wrapper.py
gradient for prior had wrong sign

@@ -116,7 +116,7 @@ class NeuralNetworkProblem(object):
             # print grad
             gradient += (f_plus-f_minus) / delta
         gradient /= 2*d_theta*R
-        gradient_prior = self.prior_penalty*np.array([np.sign(theta[0]), np.sign(theta[1])]) # (??) sign of the weights?
+        gradient_prior = -self.prior_penalty*np.array([np.sign(theta[0]), np.sign(theta[1])]) # (??) sign of the weights?
         gradient += gradient_prior
         return -gradient
 

### 2015-03-01 code/autoencoder_problem_ted/run.py

@@ -13,7 +13,7 @@ from nn import *
 keep_x        = True
 init_seed     = 1
 T             = 5000 # nbr of samples
-verbose_rate  = 1000
+verbose_rate  = 10
 C             = 20.01    # injected noise variance parameter
 eta           = 0.001 # step size for Hamiltoniam dynamics
 #h = 0.0005
@@ -29,7 +29,7 @@ upper_bounds = np.array([np.inf])
 
 if __name__ == "__main__":
   # pp.close('all')
-  problem = NeuralNetworkProblem( NeuralNetwork([28*28, 10, 28*28]), load_mnist() )
+  problem = NeuralNetworkProblem( NeuralNetwork([28*28, 100, 28*28]), load_mnist() )
   # Cheat by loading previous training NN
   # problem.nn.load('latest_epoch.json')
   chain_id = 1
@@ -41,11 +41,11 @@ if __name__ == "__main__":
   params["d_theta"] = d_theta
   params["eta"]     = eta
   params["C"]       = C
-  params["batch_size"] = 10
+  params["batch_size"] = 200
   params["verbose_rate"] = verbose_rate
   params["grad_params"]  = {"logs":{"true":[],"true_abc":[],"2side_keps":[],"2side_sl":[]},\
                             "record_2side_sl_grad":False, "record_2side_keps_grad":False,"record_true_abc_grad":False,"record_true_grad":False,
-                            "2side_keps": {"R": 10}}
+                            "2side_keps": {"R": 20}}
   params["lower_bounds"] = lower_bounds
   params["upper_bounds"] = upper_bounds
   params["keep_x"]       = keep_x

### 2015-03-01 code/autoencoder_problem_ted/working_code.py

@@ -135,7 +135,7 @@ def run_sgld( problem, params, theta, x = None ):
 
     if np.mod(t+1,verbose_rate)==0:
       if keep_x:
-        print "t = %04d    loglik = %3.3f    theta0 = %3.3f    x0 = %3.3f"%(t+1,loglike_x,theta[0],x[0][0])
+        print "t = %04d    loglik = %3.3f"%(t+1,loglike_x ) #,theta[0],x[0][0])
       else:
         print "t = %04d    theta0 = %3.3f"%(t+1,theta[0])
 

### 2015-03-01 code/autoencoder_problem/nn_wrapper.py
include biases in theta

@@ -3,7 +3,10 @@ import numpy as np
 class NeuralNetworkProblem(object):
     def __init__(self, nn, data):
         self.nn = nn
+        (self.X_train, self.T_train), (self.X_valid, self.T_valid), (self.X_test, self.T_test) = data
+        self.N = len(self.X_train)
         self.training_set, self.validation_set, self.test_set = data
+        self.prior_penalty = 0.1
 
     # Skip
     def prior_rand(self):
@@ -14,28 +17,32 @@ class NeuralNetworkProblem(object):
         pass
 
     def loglike_x(self, X, omega, simulated=True):
-        Y = []
-        for i in omega:
-            Y.append(self.training_set[i][1])
-        # Y = self.training_set[omega, 1]
-        assert len(X) == len(Y)
-        likelihood = 0.0
-        for i in range(len(X)):
-            x = X[i]
-            y = Y[i]
-            if simulated:
-                x[x < 0.5] = 0
-                x[x >= 0.5] = 1
-                y_copy = y.copy()
-                y_copy[y_copy <= 0.5] = 0
-                y_copy[y_copy > 0.5] = 1
-                # (??) Take log?
-                likelihood += np.sum(x == y_copy) / float(len(y_copy))
-            else:
-                # We are given a sample so we don't need to do feedforward
-                likelihood += np.nan_to_num(np.sum(-y*np.log(x+10**-10)-(1-y)*np.log(1-x+10**-10)))
+        Y = self.X_train[omega,:].T
+        # Y = []
+        # for i in omega:
+        #     Y.append(self.training_set[i][1])
+        # # Y = self.training_set[omega, 1]
+        # assert len(X) == len(Y)
+        # likelihood = 0.0
+        # for i in range(len(X)):
+        #     x = X[i]
+        #     y = Y[i]
+        #     if simulated:
+        #         x[x < 0.5] = 0
+        #         x[x >= 0.5] = 1
+        #         y_copy = y.copy()
+        #         y_copy[y_copy <= 0.5] = 0
+        #         y_copy[y_copy > 0.5] = 1
+        #         # (??) Take log?
+        #         likelihood += np.sum(x == y_copy) / float(len(y_copy))
+        #     else:
+        #         # We are given a sample so we don't need to do feedforward
+        #         likelihood += np.nan_to_num(np.sum(-y*np.log(x+10**-10)-(1-y)*np.log(1-x+10**-10)))
 
[…]
